---
title: "Longitudinal Panel Analysis: AI Coding Tool Adoption and Development Productivity"
output: html_document
---

This notebook is the natural follow-up to the cross-sectional diagnostic in
`ai_adoption_cross_sectional.Rmd`.  The cross-section showed that the
estimator is not the bottleneck --- unobserved confounding is.  Here we use
within-repo temporal variation to absorb time-invariant confounders and
progressively build toward a credible causal estimate.

The causal question remains: "Does adopting AI coding tools increase
development productivity (measured by monthly commits)?"  We now have a
balanced monthly panel (999 repos × 27 months, 2024-01 through 2026-03)
with staggered treatment adoption (240 ever-treated repos, 759 never-treated)
and time-varying covariates.

The notebook follows a **progression of increasingly credible designs**:

- **Stage 0**: Naive before/after (no control group, no trend adjustment)
- **Stage 1**: Interrupted time series (trend-adjusted, no control group)
- **Stage 2**: TWFE DiD without time-varying covariates
- **Stage 3**: TWFE DiD with time-varying covariates
- **Stage 4**: Borusyak et al. imputation DiD (modern staggered estimator)

Each stage builds on the previous one's limitations.

# Setup

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "knitr", "kableExtra",
                   "fixest", "didimputation", "showtext", "scales",
                   "MatchIt", "cobalt", "data.table")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace,
                                       quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(fixest)
library(didimputation)
library(showtext)
library(scales)
library(MatchIt)
library(cobalt)

font_add("Linux Libertine", regular = "LinLibertine_Rah.ttf")
showtext_auto()

# Shared theme for all plots
theme_panel <- function(base_size = 13) {
  theme_minimal(base_size = base_size, base_family = "Linux Libertine") +
    theme(
      legend.position = "bottom",
      legend.margin = margin(t = -5, r = 0, b = -3, l = 0),
      legend.box.margin = margin(t = -5, r = 0, b = -3, l = 0),
      axis.title.y = element_text(size = rel(0.8)),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )
}
```

# Load and Prepare Data

```{r load_data, message=FALSE, warning=FALSE}
panel_raw <- read.csv("../data/ai_adoption_panel.csv",
                       stringsAsFactors = FALSE)
cat(sprintf("Raw panel: %d obs, %d repos, %d months\n",
            nrow(panel_raw), n_distinct(panel_raw$full_name),
            n_distinct(panel_raw$month)))
```

```{r prepare_data}
# treatment_month and months_since_treatment are "" (not NA) for never-treated
panel <- panel_raw %>%
  mutate(
    treatment_month = na_if(treatment_month, ""),
    treatment_date  = na_if(treatment_date, ""),
    first_ai_file   = na_if(first_ai_file, ""),
    months_since_treatment = as.integer(na_if(
      as.character(months_since_treatment), ""
    ))
  )

# Repos with treatment_month outside the panel window (e.g., 2026-04)
# are effectively never-treated within our observation period
panel_months <- sort(unique(panel$month))
panel <- panel %>%
  mutate(
    treatment_month = ifelse(
      !is.na(treatment_month) & !(treatment_month %in% panel_months),
      NA_character_, treatment_month
    ),
    treated = ifelse(is.na(treatment_month), 0L, treated),
    months_since_treatment = ifelse(is.na(treatment_month),
                                    NA_integer_, months_since_treatment)
  )

panel <- panel %>%
  mutate(
    # Numeric repo ID for fixest / didimputation
    repo_id = as.integer(factor(full_name)),
    # Numeric month index (1 = 2024-01, ..., 27 = 2026-03)
    month_idx = as.integer(factor(month, levels = sort(unique(month)))),
    # Log-transformed outcome
    log_commits = log1p(commits),
    # Treatment cohort for didimputation: first treated month_idx, 0 if never
    treatment_cohort = ifelse(
      is.na(treatment_month), 0L,
      as.integer(factor(treatment_month,
                        levels = sort(unique(month))))
    ),
    # Ever-treated indicator (repo-level, time-invariant)
    ever_treated = as.integer(!is.na(treatment_month)),
    # Log-transformed time-varying covariates
    log_contributors = log1p(active_contributors),
    log_stars         = log1p(new_stars),
    log_issues        = log1p(new_issues),
    log_forks         = log1p(new_forks),
    log_releases      = log1p(new_releases),
    # Binary covariates
    is_org = as.integer(owner_type == "Organization"),
    # Repo age at panel start (years)
    created_date = as.Date(substr(created_at, 1, 10)),
    repo_age_yr = as.numeric(
      difftime(as.Date("2024-01-01"), created_date, units = "days")
    ) / 365.25
  )

# For the event study: never-treated repos get NA event_time
panel <- panel %>%
  mutate(
    event_time = ifelse(is.na(months_since_treatment), NA_integer_,
                        as.integer(months_since_treatment))
  )

cat(sprintf("Ever-treated: %d repos  |  Never-treated: %d repos\n",
            sum(panel$ever_treated[panel$month_idx == 1]),
            sum(panel$ever_treated[panel$month_idx == 1] == 0)))
cat(sprintf("Treatment cohorts span month_idx %d to %d\n",
            min(panel$treatment_cohort[panel$treatment_cohort > 0]),
            max(panel$treatment_cohort[panel$treatment_cohort > 0])))
```

# Descriptive Analysis

## Treatment Timing

```{r treatment_timing, fig.width=8, fig.height=3}
cohort_counts <- panel %>%
  filter(ever_treated == 1) %>%
  distinct(full_name, treatment_month) %>%
  count(treatment_month, name = "n_repos")

ggplot(cohort_counts, aes(x = treatment_month, y = n_repos)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  labs(x = "Treatment Month", y = "Number of Repos",
       title = "Staggered AI Tool Adoption Timing") +
  theme_panel() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Treatment adoption is staggered across 2024-11 through 2026-03, with the
bulk of adoption occurring in 2025 and early 2026.  This staggered structure
is what motivates the modern DiD estimator in Stage 4.

## Average Trajectories

```{r trajectories, fig.width=8, fig.height=3}
trajectory <- panel %>%
  group_by(month, ever_treated) %>%
  summarise(
    mean_commits = mean(commits, na.rm = TRUE),
    mean_log     = mean(log_commits, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(group = ifelse(ever_treated == 1,
                        "Ever-treated (will adopt AI tools)",
                        "Never-treated (control)"))

ggplot(trajectory, aes(x = month, y = mean_log,
                       color = group, group = group)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  labs(x = "Month", y = "Mean log(1 + Commits)",
       color = NULL,
       title = "Average Monthly Commits by Treatment Group") +
  theme_panel() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The gap between ever-treated and never-treated repos is visible *throughout
the entire panel* --- including well before any repo adopted AI tools.  This
confirms the cross-sectional finding: repos that eventually adopt AI tools
are systematically more active.  The key question for DiD is whether the
*trajectories are parallel*, not whether the levels are equal.

## Pre-Treatment Covariate Balance

```{r balance, fig.width=8, fig.height=4}
# Use the first month as a baseline snapshot for time-invariant comparison
baseline <- panel %>%
  filter(month_idx == 1) %>%
  mutate(
    log_baseline_commits = log_commits,
    log_baseline_contrib = log_contributors
  )

balance_vars <- c("repo_age_yr", "is_org", "log_baseline_commits",
                  "log_baseline_contrib")
balance_labels <- c(
  "repo_age_yr"            = "Repo age (years)",
  "is_org"                 = "Organization owner",
  "log_baseline_commits"   = "log(Baseline commits)",
  "log_baseline_contrib"   = "log(Baseline contributors)"
)

compute_smd <- function(data, var, treat_col = "ever_treated") {
  t_vals <- data[[var]][data[[treat_col]] == 1]
  c_vals <- data[[var]][data[[treat_col]] == 0]
  pooled_sd <- sqrt((var(t_vals, na.rm = TRUE) +
                     var(c_vals, na.rm = TRUE)) / 2)
  if (pooled_sd < 1e-10) return(0)
  (mean(t_vals, na.rm = TRUE) - mean(c_vals, na.rm = TRUE)) / pooled_sd
}

smd_df <- data.frame(
  variable = balance_vars,
  label = balance_labels[balance_vars],
  smd = sapply(balance_vars, function(v) compute_smd(baseline, v))
) %>%
  mutate(label = reorder(label, abs(smd)))

ggplot(smd_df, aes(x = smd, y = label)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dotted",
             color = "red", alpha = 0.5) +
  labs(x = "Standardised Mean Difference (Ever-Treated vs Never-Treated)",
       y = NULL,
       title = "Baseline Covariate Balance",
       caption = "Dotted red lines mark the ±0.1 SMD threshold") +
  theme_panel() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```

```{r balance_table}
smd_df %>%
  arrange(desc(abs(smd))) %>%
  mutate(smd = sprintf("%.3f", smd)) %>%
  select(Covariate = label, SMD = smd) %>%
  kable(caption = "Baseline Standardised Mean Differences") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The balance diagnostics show that ever-treated and never-treated repos differ
at baseline --- the same selection story as the cross-section.  DiD does *not*
require baseline balance; it requires *parallel trends*.  But large baseline
differences make parallel trends less plausible a priori, which is why we
test pre-trends carefully in the event studies below.

# Stage 0: Naive Before/After Comparison

For each ever-treated repo, compare mean monthly commits before vs. after
adoption.  This is the simplest possible estimate --- and the most biased,
because it attributes all temporal trends to the treatment.

```{r stage0}
ba <- panel %>%
  filter(ever_treated == 1, !is.na(event_time)) %>%
  mutate(post = as.integer(event_time >= 0)) %>%
  group_by(full_name, post) %>%
  summarise(mean_commits = mean(commits, na.rm = TRUE),
            mean_log     = mean(log_commits, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = post,
              values_from = c(mean_commits, mean_log),
              names_sep = "_") %>%
  mutate(
    diff_raw = mean_commits_1 - mean_commits_0,
    diff_log = mean_log_1 - mean_log_0
  )

cat("=== Naive Before/After (Treated Repos Only) ===\n")
cat(sprintf("N repos: %d\n", nrow(ba)))
cat(sprintf("Mean pre-treatment commits:  %.1f\n", mean(ba$mean_commits_0)))
cat(sprintf("Mean post-treatment commits: %.1f\n", mean(ba$mean_commits_1)))
cat(sprintf("Mean within-repo change:     %.1f commits/month\n",
            mean(ba$diff_raw)))
cat(sprintf("Mean change in log scale:    %.3f\n", mean(ba$diff_log)))

# Paired t-test on log scale
t_ba <- t.test(ba$mean_log_1, ba$mean_log_0, paired = TRUE)
cat(sprintf("Paired t-test (log): t = %.2f, p = %.4f, 95%% CI [%.3f, %.3f]\n",
            t_ba$statistic, t_ba$p.value,
            t_ba$conf.int[1], t_ba$conf.int[2]))
```

```{r stage0_plot, fig.width=8, fig.height=3}
ggplot(ba, aes(x = diff_log)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(ba$diff_log), color = "red",
             linetype = "dashed", linewidth = 0.8) +
  labs(x = "Within-Repo Change in log(1 + Monthly Commits)",
       y = "Number of Repos",
       title = "Naive Before/After: Distribution of Within-Repo Changes",
       caption = "Red dashed line = mean change. No control group; confounded by secular trends.") +
  theme_panel()
```

The naive before/after comparison shows the raw within-repo change but has no
control group.  If all repos experienced a secular trend over 2024--2026
(e.g., growing commit activity due to platform growth), the estimate
confounds the treatment effect with the trend.

# Stage 1: Interrupted Time Series (No Control Group)

ITS models the pre-treatment trend explicitly and tests for a level shift
and/or slope change at treatment onset.  Still no control group.

```{r stage1}
# Restrict to ever-treated repos
its_data <- panel %>%
  filter(ever_treated == 1) %>%
  mutate(
    post = as.integer(treated == 1),
    # Time since treatment (0 for pre-treatment periods)
    time_since = ifelse(post == 1, event_time, 0)
  )

# ITS with repo FE: level shift + slope change
its_model <- feols(
  log_commits ~ month_idx + post + time_since | repo_id,
  data = its_data,
  cluster = ~repo_id
)

cat("=== ITS Model (Treated Repos Only, Repo FE) ===\n")
summary(its_model)
```

```{r stage1_plot, fig.width=8, fig.height=3}
# Average trajectory for treated repos, centred on event time
its_trajectory <- panel %>%
  filter(ever_treated == 1, !is.na(event_time),
         event_time >= -12, event_time <= 12) %>%
  group_by(event_time) %>%
  summarise(mean_log = mean(log_commits, na.rm = TRUE),
            se = sd(log_commits, na.rm = TRUE) / sqrt(n()),
            n = n(),
            .groups = "drop")

ggplot(its_trajectory, aes(x = event_time, y = mean_log)) +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = mean_log - 1.96 * se,
                  ymax = mean_log + 1.96 * se),
              alpha = 0.2, fill = "steelblue") +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(color = "steelblue", size = 1.5) +
  annotate("text", x = -6, y = max(its_trajectory$mean_log),
           label = "Pre-treatment", family = "Linux Libertine",
           size = 3.5, color = "grey40") +
  annotate("text", x = 6, y = max(its_trajectory$mean_log),
           label = "Post-treatment", family = "Linux Libertine",
           size = 3.5, color = "grey40") +
  labs(x = "Months Since AI Tool Adoption",
       y = "Mean log(1 + Commits)",
       title = "ITS: Average Trajectory of Treated Repos Around Adoption",
       caption = "No control group. Vertical dashed line = treatment onset.") +
  theme_panel()
```

The ITS level-shift coefficient (`post`) tests whether commits jump
discretely at adoption.  The slope-change coefficient (`time_since`) tests
whether the post-treatment trajectory differs from the pre-treatment trend.
However, without a control group, we cannot rule out that a contemporaneous
shock (e.g., a GitHub platform change, a seasonal pattern) drives the result.

# Stage 2: TWFE DiD Without Time-Varying Covariates

The classic two-way fixed effects specification absorbs all time-invariant
repo heterogeneity (α_i) and all common temporal shocks (γ_t).

```{r stage2_static}
# Static TWFE
twfe_base <- feols(
  log_commits ~ treated | repo_id + month_idx,
  data = panel,
  cluster = ~repo_id
)

cat("=== TWFE DiD (No Covariates) ===\n")
summary(twfe_base)
```

```{r stage2_event, fig.width=8, fig.height=4}
# Event study: use sunab() for clean staggered event study in fixest
# sunab(cohort, event_time) implements Sun & Abraham (2021) decomposition
# but we use i() for the classic TWFE event study first
twfe_es <- feols(
  log_commits ~ i(event_time, ref = -1) | repo_id + month_idx,
  data = panel %>% filter(is.na(event_time) |
                          (event_time >= -12 & event_time <= 12)),
  cluster = ~repo_id
)

cat("=== TWFE Event Study (No Covariates) ===\n")
summary(twfe_es)
```

```{r stage2_esplot, fig.width=8, fig.height=4}
iplot(twfe_es,
      main = "TWFE Event Study: No Covariates",
      xlab = "Months Since AI Tool Adoption",
      ylab = "Coefficient (log commits)")
```

```{r stage2_pretrends}
# Pre-trends F-test: joint test that all pre-treatment coefficients = 0
pre_coefs <- grep("event_time::-[0-9]", names(coef(twfe_es)), value = TRUE)
# Exclude the reference period (-1) which is dropped
pre_coefs <- setdiff(pre_coefs, "event_time::-1")

if (length(pre_coefs) > 0) {
  pre_test <- wald(twfe_es, pre_coefs)
  cat(sprintf("Pre-trends F-test: F = %.2f, p = %.4f\n",
              pre_test$stat, pre_test$p))
}
```

The event study plot shows coefficients relative to the month before adoption
(k = -1).  Pre-treatment coefficients near zero support the parallel trends
assumption.  Post-treatment coefficients show the dynamic treatment effect.

# Stage 3: TWFE DiD With Time-Varying Covariates

Adding time-varying controls absorbs repo-specific shocks that vary over
time --- the remaining threat after repo FE absorbs time-invariant
confounders and month FE absorbs common shocks.

We include: `log1p(active_contributors)`, `log1p(new_stars)`,
`log1p(new_issues)`, `log1p(new_forks)`, `log1p(new_releases)`.
We exclude `prs_merged` because PRs are mechanistically downstream of
commits (merging a PR is committing), so conditioning on them risks blocking
the causal path.

```{r stage3_static}
twfe_cov <- feols(
  log_commits ~ treated + log_contributors + log_stars + log_issues +
    log_forks + log_releases | repo_id + month_idx,
  data = panel,
  cluster = ~repo_id
)

cat("=== TWFE DiD (With Time-Varying Covariates) ===\n")
summary(twfe_cov)
```

```{r stage3_event, fig.width=8, fig.height=4}
twfe_cov_es <- feols(
  log_commits ~ i(event_time, ref = -1) + log_contributors + log_stars +
    log_issues + log_forks + log_releases | repo_id + month_idx,
  data = panel %>% filter(is.na(event_time) |
                          (event_time >= -12 & event_time <= 12)),
  cluster = ~repo_id
)

cat("=== TWFE Event Study (With Covariates) ===\n")
summary(twfe_cov_es)
```

```{r stage3_esplot, fig.width=8, fig.height=4}
iplot(twfe_cov_es,
      main = "TWFE Event Study: With Time-Varying Covariates",
      xlab = "Months Since AI Tool Adoption",
      ylab = "Coefficient (log commits)")
```

```{r stage3_pretrends}
pre_coefs_cov <- grep("event_time::-[0-9]",
                      names(coef(twfe_cov_es)), value = TRUE)
pre_coefs_cov <- setdiff(pre_coefs_cov, "event_time::-1")

if (length(pre_coefs_cov) > 0) {
  pre_test_cov <- wald(twfe_cov_es, pre_coefs_cov)
  cat(sprintf("Pre-trends F-test (with covariates): F = %.2f, p = %.4f\n",
              pre_test_cov$stat, pre_test_cov$p))
}
```

```{r stage23_comparison}
# Side-by-side comparison
etable(twfe_base, twfe_cov,
       headers = c("TWFE (no covariates)", "TWFE (with covariates)"),
       se.below = TRUE,
       keep = "treated")
```

Adding time-varying covariates addresses observable repo-specific shocks.
If the ATT changes substantially, time-varying confounders matter.  However,
both TWFE specifications share a fundamental limitation under staggered
treatment: already-treated units serve as implicit controls for later-treated
units, and if treatment effects are heterogeneous, this can produce biased
estimates with negative weights.  Stage 4 addresses this.

# Stage 4: Borusyak et al. Imputation DiD

The imputation estimator of Borusyak, Jaravel, and Spiess (2024) avoids the
negative-weighting problem of TWFE.  It fits a model for Y(0) using only
untreated observations (never-treated + not-yet-treated), imputes the
counterfactual for each treated (i, t) cell, and averages the residuals.

```{r stage4_prep}
# didimputation requires: idname, tname, gname (cohort: 0 = never-treated),
# yname, and optionally first_stage for covariates.
# Ensure data types are correct
did_data <- panel %>%
  mutate(
    repo_id = as.integer(repo_id),
    month_idx = as.integer(month_idx),
    treatment_cohort = as.integer(treatment_cohort)
  ) %>%
  as.data.frame()

cat(sprintf("Panel for didimputation: %d obs, %d repos, %d months\n",
            nrow(did_data), n_distinct(did_data$repo_id),
            n_distinct(did_data$month_idx)))
cat(sprintf("Treatment cohorts: %d (0 = never-treated)\n",
            n_distinct(did_data$treatment_cohort)))
```

## Static ATT

```{r stage4_static}
# Static ATT (single average treatment effect)
did_static <- did_imputation(
  data = did_data,
  yname = "log_commits",
  gname = "treatment_cohort",
  tname = "month_idx",
  idname = "repo_id",
  first_stage = ~ log_contributors + log_stars + log_issues +
                  log_forks + log_releases,
  cluster_var = "repo_id"
)

cat("=== Borusyak et al. Imputation DiD: Static ATT ===\n")
print(did_static)
```

## Event Study

```{r stage4_event}
# Event study with pre-trends
did_es <- did_imputation(
  data = did_data,
  yname = "log_commits",
  gname = "treatment_cohort",
  tname = "month_idx",
  idname = "repo_id",
  first_stage = ~ log_contributors + log_stars + log_issues +
                  log_forks + log_releases,
  horizon = TRUE,
  pretrends = TRUE,
  cluster_var = "repo_id"
)

cat("=== Borusyak et al. Event Study ===\n")
print(did_es)
```

```{r stage4_esplot, fig.width=8, fig.height=4}
# Build event study plot from did_es output
es_df <- as.data.frame(did_es) %>%
  mutate(
    event_time = as.integer(as.character(term)),
    is_pre = event_time < 0
  ) %>%
  filter(event_time >= -12, event_time <= 12)

ggplot(es_df, aes(x = event_time, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "grey70") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.15, fill = "steelblue") +
  geom_point(aes(color = is_pre), size = 2) +
  geom_line(color = "steelblue", linewidth = 0.6) +
  scale_color_manual(values = c("TRUE" = "grey50", "FALSE" = "steelblue"),
                     labels = c("TRUE" = "Pre-treatment", "FALSE" = "Post-treatment"),
                     name = NULL) +
  labs(x = "Months Since AI Tool Adoption",
       y = "Estimated Effect on log(1 + Commits)",
       title = "Borusyak et al. Imputation DiD: Event Study",
       caption = "95% CIs. Pre-treatment coefficients test parallel trends.") +
  theme_panel()
```

## Pre-Trends Assessment

```{r stage4_pretrends}
pre_es <- es_df %>% filter(event_time < 0)

if (nrow(pre_es) > 0) {
  cat("Pre-treatment coefficients (Borusyak):\n")
  cat(sprintf("  Max |coef|: %.4f (at k = %d)\n",
              max(abs(pre_es$estimate)),
              pre_es$event_time[which.max(abs(pre_es$estimate))]))
  cat(sprintf("  Mean |coef|: %.4f\n", mean(abs(pre_es$estimate))))

  # Check how many pre-treatment CIs exclude zero
  n_sig <- sum(pre_es$conf.low > 0 | pre_es$conf.high < 0)
  cat(sprintf("  Pre-treatment periods with CI excluding zero: %d / %d\n",
              n_sig, nrow(pre_es)))
}
```

# Control Group Assessment

Do we need to match the control group?  We assess this through three
diagnostics: (1) the event-study pre-trends from Stages 2--4 (already shown),
(2) pre-treatment trajectory comparison, and (3) robustness to trimming
structurally inactive repos.

## Pre-Treatment Trajectory Comparison

```{r pretreat_trajectory, fig.width=8, fig.height=3}
# Earliest treatment month is 2024-11 (month_idx ~ 11)
# Use months 1-10 (2024-01 to 2024-10) as the "pure pre-treatment" window
earliest_treat_idx <- min(panel$treatment_cohort[panel$treatment_cohort > 0])

pre_traj <- panel %>%
  filter(month_idx < earliest_treat_idx) %>%
  group_by(month, month_idx, ever_treated) %>%
  summarise(mean_log = mean(log_commits, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(group = ifelse(ever_treated == 1,
                        "Ever-treated", "Never-treated"))

ggplot(pre_traj, aes(x = month_idx, y = mean_log,
                     color = group, group = group)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  labs(x = "Month Index",
       y = "Mean log(1 + Commits)",
       color = NULL,
       title = "Pre-Treatment Trajectories (Before Any Repo Adopts)",
       caption = paste0("Months 1-", earliest_treat_idx - 1,
                        " (before any treatment). Parallel = good for DiD.")) +
  theme_panel()
```

## Trimming Inactive Repos

Many never-treated repos may be structurally dormant (zero commits throughout
the panel).  These contribute trivially parallel trends (flat at zero) but
dilute the comparison.  We test robustness by dropping repos with fewer than
10 total commits across the entire panel.

```{r trim_inactive}
# Identify active repos
repo_total <- panel %>%
  group_by(full_name, repo_id, ever_treated) %>%
  summarise(total_commits = sum(commits, na.rm = TRUE), .groups = "drop")

cat("Repos by activity level:\n")
cat(sprintf("  Total commits = 0:  %d repos (%d control, %d treated)\n",
            sum(repo_total$total_commits == 0),
            sum(repo_total$total_commits == 0 & repo_total$ever_treated == 0),
            sum(repo_total$total_commits == 0 & repo_total$ever_treated == 1)))
cat(sprintf("  Total commits < 10: %d repos (%d control, %d treated)\n",
            sum(repo_total$total_commits < 10),
            sum(repo_total$total_commits < 10 & repo_total$ever_treated == 0),
            sum(repo_total$total_commits < 10 & repo_total$ever_treated == 1)))
cat(sprintf("  Total commits >= 10: %d repos (%d control, %d treated)\n",
            sum(repo_total$total_commits >= 10),
            sum(repo_total$total_commits >= 10 & repo_total$ever_treated == 0),
            sum(repo_total$total_commits >= 10 & repo_total$ever_treated == 1)))

# Re-run Borusyak on trimmed sample
active_repos <- repo_total %>% filter(total_commits >= 10)
did_data_trim <- did_data %>% filter(repo_id %in% active_repos$repo_id)

cat(sprintf("\nTrimmed panel: %d obs, %d repos\n",
            nrow(did_data_trim), n_distinct(did_data_trim$repo_id)))
```

```{r trim_borusyak}
did_trim <- did_imputation(
  data = did_data_trim,
  yname = "log_commits",
  gname = "treatment_cohort",
  tname = "month_idx",
  idname = "repo_id",
  first_stage = ~ log_contributors + log_stars + log_issues +
                  log_forks + log_releases,
  cluster_var = "repo_id"
)

cat("=== Borusyak DiD on Trimmed Sample (>= 10 total commits) ===\n")
print(did_trim)
```

## Propensity Score Matching

The parallel trends assumption is more plausible when treated and control
units are comparable at baseline.  We use propensity score matching (PSM) on
pre-treatment observables to select a control group that resembles the
treated repos *before* adoption --- then run Borusyak DiD on this matched
panel.

The matching covariates are all computed from the pure pre-treatment window
(before any repo adopts, months 1 through `r earliest_treat_idx - 1`).
This avoids contaminating the propensity score with post-treatment outcomes.

```{r psm_data}
# Collapse to repo-level pre-treatment summary for matching
match_data <- panel %>%
  filter(month_idx < earliest_treat_idx) %>%
  group_by(full_name, repo_id, ever_treated, is_org, queried_language,
           repo_age_yr) %>%
  summarise(
    pre_mean_log_commits = mean(log_commits, na.rm = TRUE),
    pre_sd_log_commits   = sd(log_commits, na.rm = TRUE),
    pre_mean_contrib     = mean(log_contributors, na.rm = TRUE),
    pre_mean_stars       = mean(log_stars, na.rm = TRUE),
    pre_mean_issues      = mean(log_issues, na.rm = TRUE),
    pre_mean_forks       = mean(log_forks, na.rm = TRUE),
    pre_mean_releases    = mean(log_releases, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # Replace NA sd (repos with constant commits) with 0
    pre_sd_log_commits = ifelse(is.na(pre_sd_log_commits), 0,
                                pre_sd_log_commits)
  )

cat(sprintf("Matching pool: %d treated, %d control\n",
            sum(match_data$ever_treated == 1),
            sum(match_data$ever_treated == 0)))
```

```{r psm_estimate}
# Estimate propensity score via logistic regression
ps_formula <- ever_treated ~ pre_mean_log_commits + pre_sd_log_commits +
  pre_mean_contrib + pre_mean_stars + pre_mean_issues + pre_mean_forks +
  pre_mean_releases + is_org + repo_age_yr

ps_model <- glm(ps_formula, data = match_data, family = binomial)
cat("=== Propensity Score Model ===\n")
summary(ps_model)
```

```{r psm_overlap, fig.width=8, fig.height=3}
match_data$pscore <- predict(ps_model, type = "response")

ggplot(match_data, aes(x = pscore,
                       fill = factor(ever_treated,
                                     labels = c("Control", "Treated")))) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  labs(x = "Propensity Score", y = "Number of Repos", fill = NULL,
       title = "Propensity Score Distribution: Overlap Check",
       caption = "Good overlap is needed for matching. Regions without overlap are trimmed.") +
  theme_panel()
```

```{r psm_match, fig.width=8, fig.height=5}
# 1:3 nearest-neighbour matching on the propensity score
m_match <- matchit(
  ps_formula,
  data = match_data,
  method = "nearest",
  distance = "glm",
  ratio = 3,
  replace = FALSE,
  caliper = 0.2
)

summary(m_match)

match_labels <- c(
  "pre_mean_log_commits" = "Pre-treatment mean log(commits)",
  "pre_sd_log_commits"   = "Pre-treatment SD log(commits)",
  "pre_mean_contrib"     = "Pre-treatment mean log(contributors)",
  "pre_mean_stars"       = "Pre-treatment mean log(stars)",
  "pre_mean_issues"      = "Pre-treatment mean log(issues)",
  "pre_mean_forks"       = "Pre-treatment mean log(forks)",
  "pre_mean_releases"    = "Pre-treatment mean log(releases)",
  "is_org"               = "Organization owner",
  "repo_age_yr"          = "Repo age (years)"
)

love.plot(m_match, binary = "std", thresholds = c(m = 0.1),
          var.names = match_labels,
          colors = c("red", "steelblue"),
          shapes = c(17, 16),
          title = "Covariate Balance: Before vs After PSM") +
  theme_panel()
```

```{r psm_matched_trajectory, fig.width=8, fig.height=3}
# Matched sample: check pre-treatment trajectories
matched_repo_data <- match.data(m_match)
matched_ids <- as.integer(matched_repo_data$repo_id)
panel_matched <- panel %>% filter(repo_id %in% matched_ids)

cat(sprintf("Matched sample: %d treated + %d control = %d repos (%d panel obs)\n",
            sum(matched_repo_data$ever_treated == 1),
            sum(matched_repo_data$ever_treated == 0),
            nrow(matched_repo_data),
            nrow(panel_matched)))

matched_traj <- panel_matched %>%
  filter(month_idx < earliest_treat_idx) %>%
  group_by(month, month_idx, ever_treated) %>%
  summarise(mean_log = mean(log_commits, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(group = ifelse(ever_treated == 1,
                        "Ever-treated", "Matched control"))

ggplot(matched_traj, aes(x = month_idx, y = mean_log,
                         color = group, group = group)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  labs(x = "Month Index",
       y = "Mean log(1 + Commits)",
       color = NULL,
       title = "Pre-Treatment Trajectories After PSM",
       caption = paste0("Months 1-", earliest_treat_idx - 1,
                        ". Matched 1:3 NN on propensity score.")) +
  theme_panel()
```

## Borusyak DiD on PSM-Matched Panel

```{r matched_borusyak}
did_data_matched <- did_data %>% filter(repo_id %in% matched_ids)

did_matched <- did_imputation(
  data = did_data_matched,
  yname = "log_commits",
  gname = "treatment_cohort",
  tname = "month_idx",
  idname = "repo_id",
  first_stage = ~ log_contributors + log_stars + log_issues +
                  log_forks + log_releases,
  cluster_var = "repo_id"
)

cat("=== Borusyak DiD on PSM-Matched Sample: Static ATT ===\n")
print(did_matched)
```

```{r matched_borusyak_es}
did_matched_es <- did_imputation(
  data = did_data_matched,
  yname = "log_commits",
  gname = "treatment_cohort",
  tname = "month_idx",
  idname = "repo_id",
  first_stage = ~ log_contributors + log_stars + log_issues +
                  log_forks + log_releases,
  horizon = TRUE,
  pretrends = TRUE,
  cluster_var = "repo_id"
)

cat("=== Borusyak Event Study on PSM-Matched Sample ===\n")
print(did_matched_es)
```

```{r matched_esplot, fig.width=8, fig.height=4}
es_matched_df <- as.data.frame(did_matched_es) %>%
  mutate(
    event_time = as.integer(as.character(term)),
    is_pre = event_time < 0
  ) %>%
  filter(event_time >= -12, event_time <= 12)

ggplot(es_matched_df, aes(x = event_time, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "grey70") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.15, fill = "#59A14F") +
  geom_point(aes(color = is_pre), size = 2) +
  geom_line(color = "#59A14F", linewidth = 0.6) +
  scale_color_manual(values = c("TRUE" = "grey50", "FALSE" = "#59A14F"),
                     labels = c("TRUE" = "Pre-treatment",
                                "FALSE" = "Post-treatment"),
                     name = NULL) +
  labs(x = "Months Since AI Tool Adoption",
       y = "Estimated Effect on log(1 + Commits)",
       title = "Borusyak DiD Event Study: PSM-Matched Sample",
       caption = "95% CIs. Matched 1:3 NN on propensity score + time-varying covariates.") +
  theme_panel()
```

The PSM-matched panel restricts the control group to the repos most
comparable to the treated units on pre-treatment observables.  If the
Borusyak ATT is similar across the full, trimmed, and PSM-matched samples,
the control group composition does not drive the result.  If the estimates
diverge, the parallel trends assumption may be sensitive to which controls
are included --- and the matched estimate is more credible because it
relies on a more comparable control group.

# Summary Comparison

```{r summary_table}
# Collect all estimates into a comparison table
results <- data.frame(
  Design = character(),
  Estimate = numeric(),
  SE = numeric(),
  CI_low = numeric(),
  CI_high = numeric(),
  N_obs = integer(),
  stringsAsFactors = FALSE
)

# Stage 0: Naive before/after
results <- rbind(results, data.frame(
  Design = "Stage 0: Naive before/after",
  Estimate = mean(ba$diff_log),
  SE = sd(ba$diff_log) / sqrt(nrow(ba)),
  CI_low = t_ba$conf.int[1],
  CI_high = t_ba$conf.int[2],
  N_obs = nrow(ba)
))

# Stage 1: ITS level shift
its_ct <- summary(its_model)$coeftable
its_idx <- which(rownames(its_ct) == "post")
results <- rbind(results, data.frame(
  Design = "Stage 1: ITS (level shift)",
  Estimate = its_ct[its_idx, "Estimate"],
  SE = its_ct[its_idx, "Std. Error"],
  CI_low = its_ct[its_idx, "Estimate"] - 1.96 * its_ct[its_idx, "Std. Error"],
  CI_high = its_ct[its_idx, "Estimate"] + 1.96 * its_ct[its_idx, "Std. Error"],
  N_obs = nobs(its_model)
))

# Stage 2: TWFE (no covariates)
tw_ct <- summary(twfe_base)$coeftable
tw_idx <- which(rownames(tw_ct) == "treated")
results <- rbind(results, data.frame(
  Design = "Stage 2: TWFE (no covariates)",
  Estimate = tw_ct[tw_idx, "Estimate"],
  SE = tw_ct[tw_idx, "Std. Error"],
  CI_low = tw_ct[tw_idx, "Estimate"] - 1.96 * tw_ct[tw_idx, "Std. Error"],
  CI_high = tw_ct[tw_idx, "Estimate"] + 1.96 * tw_ct[tw_idx, "Std. Error"],
  N_obs = nobs(twfe_base)
))

# Stage 3: TWFE (with covariates)
twc_ct <- summary(twfe_cov)$coeftable
twc_idx <- which(rownames(twc_ct) == "treated")
results <- rbind(results, data.frame(
  Design = "Stage 3: TWFE (with covariates)",
  Estimate = twc_ct[twc_idx, "Estimate"],
  SE = twc_ct[twc_idx, "Std. Error"],
  CI_low = twc_ct[twc_idx, "Estimate"] - 1.96 * twc_ct[twc_idx, "Std. Error"],
  CI_high = twc_ct[twc_idx, "Estimate"] + 1.96 * twc_ct[twc_idx, "Std. Error"],
  N_obs = nobs(twfe_cov)
))

# Stage 4: Borusyak (full sample)
bj <- as.data.frame(did_static)
results <- rbind(results, data.frame(
  Design = "Stage 4: Borusyak DiD (full)",
  Estimate = bj$estimate[1],
  SE = bj$std.error[1],
  CI_low = bj$conf.low[1],
  CI_high = bj$conf.high[1],
  N_obs = nrow(did_data)
))

# Stage 4: Borusyak (trimmed)
bt <- as.data.frame(did_trim)
results <- rbind(results, data.frame(
  Design = "Stage 4: Borusyak DiD (trimmed)",
  Estimate = bt$estimate[1],
  SE = bt$std.error[1],
  CI_low = bt$conf.low[1],
  CI_high = bt$conf.high[1],
  N_obs = nrow(did_data_trim)
))

# Stage 4: Borusyak (matched)
bm <- as.data.frame(did_matched)
results <- rbind(results, data.frame(
  Design = "Stage 4: Borusyak DiD (matched)",
  Estimate = bm$estimate[1],
  SE = bm$std.error[1],
  CI_low = bm$conf.low[1],
  CI_high = bm$conf.high[1],
  N_obs = nrow(did_data_matched)
))

results %>%
  mutate(
    Estimate = sprintf("%.3f", Estimate),
    SE       = sprintf("%.3f", SE),
    `95% CI` = sprintf("[%.3f, %.3f]", CI_low, CI_high),
    `N obs`  = format(N_obs, big.mark = ",")
  ) %>%
  select(Design, Estimate, SE, `95% CI`, `N obs`) %>%
  kable(caption = "Summary: All Estimates of the ATT of AI Tool Adoption on log(1 + Monthly Commits)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r summary_plot, fig.width=8, fig.height=4}
results$Design <- factor(results$Design, levels = rev(results$Design))

ggplot(results, aes(x = Estimate, y = Design)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high),
                  size = 0.6, color = "steelblue") +
  labs(x = "Estimated ATT on log(1 + Monthly Commits)",
       y = NULL,
       title = "Comparison of All Estimates",
       caption = "95% CIs. Progression from naive to credible designs.") +
  theme_panel() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```

# Interpretation

The progression from Stage 0 to Stage 4 illustrates the tutorial's core
message: *the design matters more than the estimator*.

- **Stages 0--1** (no control group) produce estimates that confound the
  treatment effect with secular trends.  They are the longitudinal analogue
  of the bivariate cross-sectional comparison.

- **Stages 2--3** (TWFE DiD) absorb time-invariant confounders (repo FE) and
  common shocks (month FE).  Adding time-varying covariates in Stage 3
  addresses repo-specific shocks.  However, TWFE is biased under staggered
  treatment with heterogeneous effects.

- **Stage 4** (Borusyak et al.) resolves the staggered-treatment bias and
  provides clean event-study estimates.  The robustness of the ATT across
  the full, trimmed, and matched samples addresses concerns about control
  group composition.

The event-study plots across Stages 2--4 serve a dual purpose: they test the
parallel trends assumption (pre-treatment coefficients near zero) *and*
visualise the dynamic treatment effect (post-treatment coefficients).

Compare the final Borusyak ATT with the cross-sectional coefficients from
the companion notebook.  The cross-sectional M_pretreat coefficient reflects
both the causal effect and selection bias (productive repos adopt AI tools).
The longitudinal ATT, by absorbing within-repo variation, strips away the
selection component --- the difference between the two estimates is an
upper bound on the magnitude of selection bias in the cross-section.
