---
title: "Cross-Sectional Diagnostic: AI Coding Tool Adoption and Development Productivity"
output: html_document
---

This notebook implements a cross-sectional diagnostic analysis of AI coding
tool adoption across top GitHub repositories, following the four-stage
structure from the Bogner & Merkel reanalysis.

The causal question is: "Does adopting AI coding tools increase development
productivity (measured by monthly commits)?"  The diagnostic arc shows that
cross-sectional data cannot credibly answer this question --- even though the
raw association is enormous and survives conventional controls:

- **Stage 1** documents severe covariate imbalance between repos with AI
  tool configuration files (L2+) and those without (L1).  L2+ repos average
  5.8x more monthly commits (412 vs. 71; the median gap is even starker at
  15x), but they are also systematically larger (SMD = 0.75), have far more
  PRs (0.91), more total commits (0.74), and are more likely to have CI/CD
  (0.54) --- all observable proxies for the unobservable confounders
  (developer skill, team culture) that drive both adoption and productivity.
- **Stage 2** shows that the treatment coefficient drops by 47% (from 2.31
  to 1.23) when DAG-justified confounders are added, yet remains highly
  significant (p < 0.001).  This is *not* reassuring: the Oster bound
  (δ* = 0.56) reveals that unobservables need only be about half as
  important as observables to explain the residual effect entirely, and the
  sensemakr robustness value (23%) is plausible given the obvious unobserved
  confounders.
- **Stage 3** shows that PSM and IPW produce nearly identical estimates to
  OLS (1.23, 1.17, 1.18) --- the estimator is not the bottleneck.
- **Stage 4** synthesises the cross-sectional ceiling argument: the
  coefficient's apparent robustness to controls is *exactly* what we would
  expect under selection bias (productive teams adopt AI tools), because the
  key confounders are unobservable from GitHub API data.  Only longitudinal
  data --- comparing the same repo before and after adoption --- can make
  progress, which is what Example C (Cursor ITS/DiD) demonstrates.

# Setup

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "knitr", "kableExtra",
                   "lmtest", "sandwich", "sensemakr", "MatchIt", "cobalt",
                   "showtext", "scales")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(lmtest)
library(sandwich)
library(sensemakr)
library(MatchIt)
library(cobalt)
library(showtext)
library(scales)

font_add("Linux Libertine", regular = "LinLibertine_Rah.ttf")
showtext_auto()
```

# Load and Prepare Data

```{r load_data, message=FALSE, warning=FALSE}
repos <- read.csv("../data/ai_adoption_repos.csv", stringsAsFactors = FALSE)
monthly <- read.csv("../data/ai_adoption_monthly_commits.csv",
                     stringsAsFactors = FALSE)

cat(sprintf("Repos: %d\n", nrow(repos)))
cat(sprintf("Monthly commit rows: %d\n", nrow(monthly)))
```

```{r prepare_data}
df <- repos %>%
  mutate(
    # Treatment: L2+ vs L1
    treated     = as.integer(ai_maturity_level >= 2),
    # Derive repo age in years from creation date
    created_at  = as.POSIXct(created_at, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    pushed_at   = as.POSIXct(pushed_at,  format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    repo_age_yr = as.numeric(difftime(Sys.time(), created_at, units = "days")) / 365.25,
    # Log-transform skewed covariates (add 1 to handle zeros)
    log_stars        = log1p(stars),
    log_forks        = log1p(forks),
    log_size         = log1p(size_kb),
    log_total_commits = log1p(total_commits),
    log_total_prs    = log1p(total_prs),
    log_total_issues = log1p(total_issues),
    log_releases     = log1p(total_releases),
    log_open_issues  = log1p(open_issues),
    # Binary covariates
    is_org       = as.integer(owner_type == "Organization"),
    has_ci       = as.integer(has_ci == "True" | has_ci == TRUE),
    has_wiki     = as.integer(has_wiki == "True" | has_wiki == TRUE),
    has_discuss   = as.integer(has_discussions == "True" | has_discussions == TRUE)
  )

# Compute average monthly commits from the 6-month window columns
commit_cols <- grep("^commits_", names(df), value = TRUE)
df$avg_monthly_commits <- rowMeans(df[, commit_cols], na.rm = TRUE)
df$log_avg_monthly_commits <- log1p(df$avg_monthly_commits)

# Total commits in the 6-month window
df$total_window_commits <- rowSums(df[, commit_cols], na.rm = TRUE)
df$log_total_window_commits <- log1p(df$total_window_commits)

cat(sprintf("Treatment (L2+): %d  |  Control (L1): %d\n",
            sum(df$treated == 1), sum(df$treated == 0)))
cat(sprintf("Treatment prevalence: %.1f%%\n", mean(df$treated) * 100))
```

# Stage 1: Descriptive Statistics and Covariate Balance

## 1.1 AI Maturity Distribution

```{r maturity_distribution}
# By level
table(df$ai_maturity_level) %>%
  as.data.frame() %>%
  setNames(c("AI Maturity Level", "N")) %>%
  kable(caption = "AI Maturity Level Distribution") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r maturity_by_language}
# By language x treatment
crosstab <- df %>%
  group_by(queried_language, treated) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = treated, values_from = n, values_fill = 0) %>%
  rename(Control = `0`, Treated = `1`) %>%
  mutate(Total = Control + Treated,
         `% Treated` = sprintf("%.0f%%", Treated / Total * 100)) %>%
  arrange(desc(Treated))

crosstab %>%
  kable(caption = "Treatment (L2+) vs Control (L1) by Language") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 1.2 Outcome: Monthly Commits

```{r outcome_comparison, fig.width=8, fig.height=3}
group_labels <- c("0" = "L1 (no AI config)", "1" = "L2+ (AI config)")

df %>%
  mutate(group = factor(treated, labels = group_labels)) %>%
  ggplot(aes(x = log_avg_monthly_commits, fill = group)) +
  geom_density(alpha = 0.5) +
  labs(x = "Log(1 + Avg Monthly Commits)", y = "Density",
       fill = NULL,
       title = "Distribution of Monthly Commits by AI Adoption Status") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(legend.position = "bottom",
        legend.margin = margin(t = -5, r = 0, b = -3, l = 0),
        legend.box.margin = margin(t = -5, r = 0, b = -3, l = 0),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

```{r outcome_summary}
df %>%
  mutate(group = ifelse(treated == 1, "L2+ (treated)", "L1 (control)")) %>%
  group_by(group) %>%
  summarise(
    N = n(),
    `Mean monthly commits` = sprintf("%.1f", mean(avg_monthly_commits)),
    `Median monthly commits` = sprintf("%.1f", median(avg_monthly_commits)),
    `Mean log monthly commits` = sprintf("%.2f", mean(log_avg_monthly_commits)),
    .groups = "drop"
  ) %>%
  kable(caption = "Monthly Commits by Treatment Group") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 1.3 Covariate Balance (Standardised Mean Differences)

```{r covariate_balance, fig.width=8, fig.height=5}
# Compute SMD: (mean_treated - mean_control) / pooled_sd
balance_vars <- c("repo_age_yr", "log_stars", "log_forks", "log_size",
                  "log_total_commits", "log_total_prs", "log_total_issues",
                  "log_releases", "log_open_issues", "is_org", "has_ci",
                  "has_wiki", "has_discuss")

balance_labels <- c(
  "repo_age_yr"       = "Repo age (years)",
  "log_stars"         = "log(Stars)",
  "log_forks"         = "log(Forks)",
  "log_size"          = "log(Size KB)",
  "log_total_commits" = "log(Total commits)",
  "log_total_prs"     = "log(Total PRs)",
  "log_total_issues"  = "log(Total issues)",
  "log_releases"      = "log(Releases)",
  "log_open_issues"   = "log(Open issues)",
  "is_org"            = "Organization owner",
  "has_ci"            = "Has CI/CD",
  "has_wiki"          = "Has wiki",
  "has_discuss"       = "Has discussions"
)

compute_smd <- function(data, var, treat_col = "treated") {
  t_vals <- data[[var]][data[[treat_col]] == 1]
  c_vals <- data[[var]][data[[treat_col]] == 0]
  pooled_sd <- sqrt((var(t_vals, na.rm = TRUE) + var(c_vals, na.rm = TRUE)) / 2)
  if (pooled_sd < 1e-10) return(0)
  (mean(t_vals, na.rm = TRUE) - mean(c_vals, na.rm = TRUE)) / pooled_sd
}

smd_df <- data.frame(
  variable = balance_vars,
  label = balance_labels[balance_vars],
  smd = sapply(balance_vars, function(v) compute_smd(df, v))
) %>%
  mutate(label = reorder(label, abs(smd)))

ggplot(smd_df, aes(x = smd, y = label)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dotted", color = "red", alpha = 0.5) +
  labs(x = "Standardised Mean Difference (L2+ vs L1)",
       y = NULL,
       title = "Covariate Balance Before Adjustment",
       caption = "Dotted red lines mark ±0.1 SMD threshold") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```

```{r smd_table}
smd_df %>%
  arrange(desc(abs(smd))) %>%
  mutate(smd = sprintf("%.3f", smd)) %>%
  select(Covariate = label, SMD = smd) %>%
  kable(caption = "Standardised Mean Differences (L2+ minus L1)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 1.4 Interpretation

The balance diagnostics confirm the selection story.  Every covariate
exceeds the conventional ±0.1 SMD threshold, with total PRs (SMD = 0.91),
repo size (0.75), total commits (0.74), releases (0.61), and CI presence
(0.54) showing the most severe imbalances.

These are not random differences --- they reflect a coherent selection
mechanism: well-resourced, actively maintained, organization-backed
projects are the ones that adopt AI tools.  These same characteristics
independently predict high commit velocity.  Language confounding amplifies
the problem: TypeScript (53% treated) and Go (31%) have strong AI tooling
ecosystems, while C (9%) and Swift (13%) barely participate.

The proposed causal DAG encodes this structure:

- **Confounders** (common causes of AI adoption AND productivity):
  repo age, primary language, organization type, CI/CD presence, repo size,
  number of releases.  These are pre-treatment or structural features that
  we can condition on.
- **Collider**: stars / popularity --- caused by both high commit activity
  and AI adoption hype.  Conditioning on it risks Berkson's paradox.
- **Post-treatment**: current contributor count may increase *because* of
  AI tool adoption (lower barrier to entry).
- **Unobserved**: developer skill, team culture, organizational investment
  in tooling.  These are the confounders that matter most and that no
  amount of GitHub API data can capture.

The 5.8x raw gap in monthly commits (15x at the median) is strikingly
large but causally uninterpretable.  The question is how much survives
after controlling for observable confounders --- and whether the residual
is credible.

# Stage 2: OLS and Omitted Variable Bias Diagnostics

## 2.1 DAG-Justified vs Kitchen-Sink OLS

```{r ols_models}
# M_dag: only DAG-justified confounders
m_dag <- lm(log_avg_monthly_commits ~ treated + repo_age_yr + queried_language +
              is_org + has_ci + log_size + log_releases,
            data = df)

# M_all: kitchen-sink including collider (stars) and post-treatment variables
m_all <- lm(log_avg_monthly_commits ~ treated + repo_age_yr + queried_language +
              is_org + has_ci + log_size + log_releases +
              log_stars + log_forks + log_total_prs + log_total_issues,
            data = df)

# Extract treatment coefficients with HC1 robust SEs
extract_treated <- function(model, label) {
  ct <- coeftest(model, vcov = vcovHC(model, type = "HC1"))
  idx <- which(rownames(ct) == "treated")
  data.frame(
    model    = label,
    coef     = ct[idx, "Estimate"],
    se       = ct[idx, "Std. Error"],
    t_val    = ct[idx, "t value"],
    p_val    = ct[idx, "Pr(>|t|)"],
    r2       = summary(model)$r.squared,
    adj_r2   = summary(model)$adj.r.squared,
    n        = nobs(model)
  )
}

ols_results <- bind_rows(
  extract_treated(m_dag, "M_dag (DAG-justified)"),
  extract_treated(m_all, "M_all (kitchen-sink)")
)

ols_results %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(as.numeric(p_val) < 0.001, "<0.001", sprintf("%.3f", p_val)),
    r2    = sprintf("%.3f", r2)
  ) %>%
  select(Model = model, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, `R²` = r2, N = n) %>%
  kable(caption = "OLS Estimates of AI Adoption Effect on log(Monthly Commits)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The coefficient drops by 47% from the bivariate estimate (2.31 → 1.23) when
DAG-justified confounders are added (language, age, org type, CI, size,
releases absorb a large portion of the raw gap).  The kitchen-sink model
adds another 23% attenuation (→ 0.95).  Yet both controlled estimates
remain highly significant (p < 0.001), and the R² jumps from 0.21 to 0.53.

A naive analyst might take this as evidence that the effect is "robust to
controls."  The OVB diagnostics below show why that conclusion is premature.

## 2.2 Coefficient Stability Across Specifications

```{r coefficient_stability, fig.width=8, fig.height=3}
# Build a sequence of progressively richer models
formulas <- list(
  "Bivariate"        = log_avg_monthly_commits ~ treated,
  "+ Age + Language"  = log_avg_monthly_commits ~ treated + repo_age_yr + queried_language,
  "+ Org + CI"        = log_avg_monthly_commits ~ treated + repo_age_yr + queried_language +
                          is_org + has_ci,
  "M_dag (+ Size, Releases)" = log_avg_monthly_commits ~ treated + repo_age_yr +
                          queried_language + is_org + has_ci + log_size + log_releases,
  "M_all (+ Stars, Forks, ...)" = log_avg_monthly_commits ~ treated + repo_age_yr +
                          queried_language + is_org + has_ci + log_size + log_releases +
                          log_stars + log_forks + log_total_prs + log_total_issues
)

stability <- lapply(names(formulas), function(label) {
  mod <- lm(formulas[[label]], data = df)
  ct <- coeftest(mod, vcov = vcovHC(mod, type = "HC1"))
  idx <- which(rownames(ct) == "treated")
  data.frame(spec = label, coef = ct[idx, 1], se = ct[idx, 2],
             stringsAsFactors = FALSE)
}) %>%
  bind_rows() %>%
  mutate(spec = factor(spec, levels = names(formulas)))

# Mark DAG-justified zone
dag_idx <- which(levels(stability$spec) == "M_dag (+ Size, Releases)")

ggplot(stability, aes(x = spec, y = coef)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  annotate("rect",
           xmin = dag_idx - 0.4, xmax = dag_idx + 0.4,
           ymin = -Inf, ymax = Inf,
           fill = "steelblue", alpha = 0.12) +
  geom_pointrange(aes(ymin = coef - 1.96 * se, ymax = coef + 1.96 * se), size = 0.5) +
  labs(x = NULL, y = "Coefficient on AI Adoption (treated)",
       title = "Coefficient Stability Across Specifications",
       caption = "Shaded region: DAG-justified specification. 95% CI with HC1 SEs.") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

The coefficient stability plot shows steady attenuation as controls are
added: from 2.31 (bivariate) to 1.23 (M_dag) to 0.95 (M_all).  The
trajectory is monotonically downward, and even the kitchen-sink estimate
remains far from zero.

This *looks* robust --- but the attenuation pattern itself is the warning
sign.  Each batch of controls explains away more of the association.  The
question is whether the unobserved confounders (developer skill, team
culture) --- which are plausibly *more* strongly associated with both
treatment and outcome than any individual observed covariate --- would
continue the downward trajectory to zero.

## 2.3 Cinelli & Hazlett Sensitivity Analysis (sensemakr)

```{r sensemakr_analysis}
# sensemakr on the DAG-justified model
sense_dag <- sensemakr(m_dag, treatment = "treated",
                       benchmark_covariates = c("log_size", "repo_age_yr"),
                       kd = 1:3)

cat("=== M_dag: Sensitivity analysis ===\n")
summary(sense_dag)
```

```{r sensemakr_plot, fig.width=8, fig.height=5}
plot(sense_dag, sensitivity.of = "estimate")
```

```{r sensemakr_all}
# sensemakr on the kitchen-sink model
sense_all <- sensemakr(m_all, treatment = "treated",
                       benchmark_covariates = c("log_stars", "log_size"),
                       kd = 1:3)

cat("=== M_all: Sensitivity analysis ===\n")
summary(sense_all)
```

The robustness value (RV) for the DAG-justified model is 23%.  This means
a single unobserved confounder would need to explain at least 23% of the
residual variation in both treatment assignment and outcome to nullify the
estimate at the 5% level.  Is 23% plausible?

Consider "developer skill" as a confounder.  Skilled developers are more
likely to adopt AI tools (they are early adopters, they configure their
tooling carefully) *and* more likely to commit frequently (they are more
productive).  A confounder this strong is not only plausible --- it is
expected.  The RV suggests the estimate is vulnerable.

## 2.4 Oster Proportional Selection Bounds

```{r oster_bounds}
# Oster (2019): δ* = how much selection on unobservables (relative to
# observables) is needed to drive the treatment effect to zero.
# δ* = (β_full * (R_max - R_full)) / ((β_restricted - β_full) * (R_full - R_restricted))
# We use R_max = min(1, 1.3 * R_full) following Oster's recommendation.

oster_delta <- function(model_restricted, model_full, r_max_multiplier = 1.3) {
  b_r <- coef(model_restricted)["treated"]
  b_f <- coef(model_full)["treated"]
  r_r <- summary(model_restricted)$r.squared
  r_f <- summary(model_full)$r.squared
  r_max <- min(1, r_max_multiplier * r_f)

  delta <- (b_f * (r_max - r_f)) / ((b_r - b_f) * (r_f - r_r))
  return(data.frame(
    beta_restricted = b_r,
    beta_full       = b_f,
    R2_restricted   = r_r,
    R2_full         = r_f,
    R2_max          = r_max,
    delta_star      = delta
  ))
}

m_bivariate <- lm(log_avg_monthly_commits ~ treated, data = df)

oster_dag <- oster_delta(m_bivariate, m_dag)
oster_all <- oster_delta(m_dag, m_all)

bind_rows(
  oster_dag %>% mutate(comparison = "Bivariate → M_dag"),
  oster_all %>% mutate(comparison = "M_dag → M_all")
) %>%
  mutate(across(where(is.numeric), ~sprintf("%.3f", .))) %>%
  select(Comparison = comparison, `β restricted` = beta_restricted,
         `β full` = beta_full, `R² restricted` = R2_restricted,
         `R² full` = R2_full, `R² max` = R2_max, `δ*` = delta_star) %>%
  kable(caption = "Oster Proportional Selection Bounds") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The Oster δ* for the bivariate → M_dag comparison is the most informative:
it tells us how strong selection on unobservables must be, relative to the
selection on observables we already see, to drive the treatment effect to
zero.

A δ* of 0.56 means that unobservables need only be **about half as important**
as the observables we added (language, age, org type, CI, size, releases)
to explain the entire residual association.  Given that the most important
confounders (developer skill, team culture, organizational resources) are
*precisely* the ones we cannot observe from GitHub API data, δ* < 1 is a
clear signal of fragility.

The M_dag → M_all δ* is much larger, but this comparison is less
informative: the additional controls include colliders (stars) and
post-treatment variables (PRs, issues), so the small incremental R²
gain does not represent genuine confounding removal.

# Stage 3: Propensity Score Matching and IPW

Stage 3 asks whether alternative selection-on-observables methods
(matching, IPW) yield different answers from OLS.  If the problem were
model misspecification (nonlinearity, poor overlap), matching or IPW might
help.  If the problem is unobserved confounding, all three estimators will
converge on the same biased answer --- because they share the same
identifying assumption (conditional ignorability given observables).

## 3.1 Propensity Score Estimation

```{r psm_estimation}
# Estimate propensity score using DAG-justified confounders
ps_formula <- treated ~ repo_age_yr + queried_language + is_org + has_ci +
                log_size + log_releases

m_ps <- matchit(ps_formula, data = df, method = "nearest",
                distance = "glm", ratio = 1, replace = FALSE)

summary(m_ps)
```

## 3.2 Balance After Matching

```{r psm_balance, fig.width=8, fig.height=5}
love.plot(m_ps, binary = "std", thresholds = c(m = 0.1),
          var.names = balance_labels,
          colors = c("red", "steelblue"),
          shapes = c(17, 16),
          title = "Covariate Balance: Before vs After Matching") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine")
```

## 3.3 Matched Sample OLS

```{r psm_ols}
matched_data <- match.data(m_ps)

m_psm <- lm(log_avg_monthly_commits ~ treated + repo_age_yr + queried_language +
               is_org + has_ci + log_size + log_releases,
             data = matched_data, weights = weights)

psm_ct <- coeftest(m_psm, vcov = vcovHC(m_psm, type = "HC1"))
idx <- which(rownames(psm_ct) == "treated")

psm_result <- data.frame(
  method = "PSM (1:1 NN)",
  coef   = psm_ct[idx, 1],
  se     = psm_ct[idx, 2],
  p_val  = psm_ct[idx, 4],
  n      = nrow(matched_data)
)
```

## 3.4 Inverse Probability Weighting

```{r ipw}
# Estimate propensity scores
df$ps <- predict(glm(ps_formula, data = df, family = binomial), type = "response")

# IPW weights for ATT: treated get weight 1, controls get PS/(1-PS)
df$ipw_weight <- ifelse(df$treated == 1, 1, df$ps / (1 - df$ps))

# Effective sample size for controls
ess_control <- sum(df$ipw_weight[df$treated == 0])^2 /
               sum(df$ipw_weight[df$treated == 0]^2)
cat(sprintf("IPW effective sample size (controls): %.0f of %d\n",
            ess_control, sum(df$treated == 0)))

m_ipw <- lm(log_avg_monthly_commits ~ treated + repo_age_yr + queried_language +
               is_org + has_ci + log_size + log_releases,
             data = df, weights = ipw_weight)

ipw_ct <- coeftest(m_ipw, vcov = vcovHC(m_ipw, type = "HC1"))
idx <- which(rownames(ipw_ct) == "treated")

ipw_result <- data.frame(
  method = "IPW (ATT)",
  coef   = ipw_ct[idx, 1],
  se     = ipw_ct[idx, 2],
  p_val  = ipw_ct[idx, 4],
  n      = nrow(df)
)
```

## 3.5 Comparison: OLS, PSM, and IPW

```{r three_estimator_comparison}
all_results <- bind_rows(
  ols_results %>%
    filter(model == "M_dag (DAG-justified)") %>%
    transmute(method = "OLS (M_dag)", coef, se, p_val, n),
  psm_result,
  ipw_result
) %>%
  mutate(method = factor(method, levels = c("OLS (M_dag)", "PSM (1:1 NN)", "IPW (ATT)")))

all_results %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(p_val < 0.001, "<0.001", sprintf("%.3f", p_val))
  ) %>%
  select(Method = method, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, N = n) %>%
  kable(caption = "Three Estimators, Same DAG-Justified Covariates") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r three_estimator_plot, fig.width=8, fig.height=3}
ggplot(all_results, aes(x = method, y = coef)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = coef - 1.96 * se, ymax = coef + 1.96 * se,
                      color = method),
                  size = 0.8) +
  scale_color_manual(values = c("OLS (M_dag)" = "#4E79A7",
                                "PSM (1:1 NN)" = "#59A14F",
                                "IPW (ATT)"    = "#F28E2B")) +
  labs(x = NULL, y = "Coefficient on AI Adoption",
       title = "Three Estimators Converge Under the Same Covariates",
       caption = "All use DAG-justified confounders. 95% CI with HC1 robust SEs.") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

# Stage 4: The Cross-Sectional Ceiling

## The diagnostic summary

The four stages build a cumulative case:

1. **The raw gap is enormous** (5.8x in means, 15x at the median), but
   covariate balance shows that L2+ repos differ from L1 on *every*
   observable dimension.  The treated group is a systematically different
   population, not a random subset.

2. **Controls cut the coefficient by 47%** (2.31 → 1.23), and the
   coefficient *keeps falling* as more controls are added (→ 0.95).
   The downward trajectory has not flattened.

3. **OVB diagnostics say the residual is fragile**: Oster δ* = 0.56 means
   unobservables need only be about half as important as observables to
   explain everything away.  The sensemakr RV of 23% is well within what
   "developer skill" or "team culture" could plausibly contribute.

4. **The estimator is not the bottleneck**: OLS, PSM, and IPW converge
   tightly (1.23, 1.17, 1.18).  Switching from regression to matching
   to weighting changes nothing because all three share the same
   identifying assumption --- and that assumption is violated.

## Why the coefficient survives controls

It is tempting to interpret the persistently significant coefficient as
evidence of a genuine causal effect.  But the persistence is *exactly*
what selection bias predicts:

- **Developer skill** (unobserved) is plausibly the strongest confounder.
  Skilled developers both commit more code *and* adopt AI tools earlier.
  No GitHub API covariate captures this.
- **Organizational investment** (unobserved) drives both tooling adoption
  and development velocity.  Being organization-owned (which we observe)
  is a crude proxy; the actual confounders are team size, budget, and
  engineering culture.
- **Reverse causality** compounds the problem: high-velocity projects
  have more reason to invest in AI tooling, so the causal arrow may
  run from productivity to adoption rather than the reverse.

The observables we control for (repo age, language, org type, CI, size,
releases) are *proxies* for the true confounders, not the confounders
themselves.  They absorb some but not all of the selection, which is why
the coefficient attenuates but does not vanish.

## The bridge to longitudinal data

Cross-sectional data cannot separate "AI tools increase productivity"
from "productive teams adopt AI tools."  The identifying variation must
come from *within-repo temporal changes* --- comparing the same project
before and after AI adoption, so that all time-invariant confounders
(developer skill, org culture, project domain) are absorbed by
construction.

This is precisely what Example C (Cursor ITS/DiD) provides.  There, the
naive analysis (without repo fixed effects) produces qualitatively wrong
conclusions about code quality --- the sign *flips* when within-repo
variation is used.  The velocity finding (commits increase) *does* survive
the longitudinal design, but the cross-sectional analysis we performed
here cannot know that in advance.  The lesson is not that the causal effect
is zero, but that the cross-sectional coefficient --- however large and
significant --- is uninterpretable.

# Summary Table

```{r summary_table}
summary_stages <- data.frame(
  Stage = c("Stage 1: Covariate balance",
            "Stage 2: OLS + OVB diagnostics",
            "Stage 3: PSM and IPW",
            "Stage 4: Cross-sectional ceiling"),
  Finding = c(
    paste0("L2+ repos average 5.8x more monthly commits (15x at the median), ",
           "but differ on every observable (total PRs SMD = 0.91, size = 0.75, ",
           "commits = 0.74, CI = 0.54)"),
    paste0("Coefficient drops 47% (2.31 → 1.23) with DAG controls, but stays ",
           "significant (p < 0.001); Oster δ* = 0.56 says unobservables need only ",
           "be about half as strong as observables to explain it away; RV = 23%"),
    paste0("OLS (1.23), PSM (1.17), IPW (1.18) converge tightly — the estimator ",
           "is not the bottleneck; the identifying assumption is"),
    paste0("The persistent coefficient is consistent with selection bias (skilled ",
           "teams both adopt AI tools and commit more); only within-repo longitudinal ",
           "variation can make progress (→ Example C)")
  )
)

summary_stages %>%
  kable(caption = "Diagnostic Arc Summary") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```
