---
title: "Longitudinal Panel Analysis: AI Coding Tool Adoption and Development Productivity"
output: html_document
---

This notebook is the natural follow-up to the cross-sectional diagnostic in
`ai_adoption_cross_sectional.Rmd`.  The cross-section showed that the
estimator is not the bottleneck --- unobserved confounding is.  Here we use
within-repo temporal variation to absorb time-invariant confounders and
progressively build toward a credible causal estimate.

The causal question remains: "Does adopting AI coding tools increase
development productivity (measured by monthly commits)?"  We now have a
balanced monthly panel (999 repos × 27 months, 2024-01 through 2026-03)
with staggered treatment adoption (240 ever-treated repos, 759 never-treated)
and time-varying covariates.

The notebook follows a **progression of increasingly credible designs**:

- **Stage 0**: Naive before/after (no control group, no trend adjustment)
- **Stage 1**: Interrupted time series (trend-adjusted, no control group)
- **Stage 2**: Modern staggered DiD with control group --- Borusyak et al.
  (2024) imputation estimator *and* Callaway & Sant'Anna (2021)
  outcome-regression estimator

Each stage builds on the previous one's limitations.  We skip the classic
TWFE DiD estimator because it has known problems under staggered treatment
with heterogeneous effects (Goodman-Bacon 2021; de Chaisemartin &
D'Haultfoeuille 2020) and go directly to modern estimators.  Using two
complementary approaches lets us assess sensitivity to how covariates are
handled (global imputation vs. group-time outcome regression).

# Setup

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "knitr", "kableExtra",
                   "fixest", "didimputation", "clock", "recipes", "did", 
                   "showtext", "scales", "MatchIt", "cobalt", "data.table")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace,
                                       quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(fixest)
library(didimputation)
library(did)
library(showtext)
library(scales)
library(MatchIt)
library(cobalt)

font_add("Linux Libertine", regular = "LinLibertine_Rah.ttf")
showtext_auto()

plots_dir <- file.path(dirname(getwd()), "plots")
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)

# Shared theme for all plots
theme_panel <- function(base_size = 13) {
  theme_minimal(base_size = base_size, base_family = "Linux Libertine") +
    theme(
      legend.position = "bottom",
      legend.margin = margin(t = -5, r = 0, b = -3, l = 0),
      legend.box.margin = margin(t = -5, r = 0, b = -3, l = 0),
      axis.title.y = element_text(size = rel(0.8)),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )
}
```

# Load and Prepare Data

```{r load_data, message=FALSE, warning=FALSE}
panel_raw <- read.csv("../data/ai_adoption_panel.csv",
                       stringsAsFactors = FALSE)
cat(sprintf("Raw panel: %d obs, %d repos, %d months\n",
            nrow(panel_raw), n_distinct(panel_raw$full_name),
            n_distinct(panel_raw$month)))
```

```{r prepare_data}
# treatment_month and months_since_treatment are "" (not NA) for never-treated
panel <- panel_raw %>%
  mutate(
    treatment_month = na_if(treatment_month, ""),
    treatment_date  = na_if(treatment_date, ""),
    first_ai_file   = na_if(first_ai_file, ""),
    months_since_treatment = as.integer(na_if(
      as.character(months_since_treatment), ""
    ))
  )

# Repos with treatment_month outside the panel window (e.g., 2026-04)
# are effectively never-treated within our observation period
panel_months <- sort(unique(panel$month))
panel <- panel %>%
  mutate(
    treatment_month = ifelse(
      !is.na(treatment_month) & !(treatment_month %in% panel_months),
      NA_character_, treatment_month
    ),
    treated = ifelse(is.na(treatment_month), 0L, treated),
    months_since_treatment = ifelse(is.na(treatment_month),
                                    NA_integer_, months_since_treatment)
  )

panel <- panel %>%
  mutate(
    # Numeric repo ID for fixest / didimputation
    repo_id = as.integer(factor(full_name)),
    # Numeric month index (1 = 2024-01, ..., 27 = 2026-03)
    month_idx = as.integer(factor(month, levels = sort(unique(month)))),
    # Log-transformed outcome
    log_commits = log1p(commits),
    # Treatment cohort for didimputation: first treated month_idx, 0 if never
    treatment_cohort = ifelse(
      is.na(treatment_month), 0L,
      as.integer(factor(treatment_month,
                        levels = sort(unique(month))))
    ),
    # Ever-treated indicator (repo-level, time-invariant)
    ever_treated = as.integer(!is.na(treatment_month)),
    # Log-transformed time-varying covariates
    log_contributors = log1p(active_contributors),
    log_stars         = log1p(new_stars),
    log_issues        = log1p(new_issues),
    log_forks         = log1p(new_forks),
    log_releases      = log1p(new_releases),
    # Binary covariates
    is_org = as.integer(owner_type == "Organization"),
    # Repo age at panel start (years)
    created_date = as.Date(substr(created_at, 1, 10)),
    repo_age_yr = as.numeric(
      difftime(as.Date("2024-01-01"), created_date, units = "days")
    ) / 365.25
  )

# For the event study: never-treated repos get NA event_time
panel <- panel %>%
  mutate(
    event_time = ifelse(is.na(months_since_treatment), NA_integer_,
                        as.integer(months_since_treatment))
  )

cat(sprintf("Ever-treated: %d repos  |  Never-treated: %d repos\n",
            sum(panel$ever_treated[panel$month_idx == 1]),
            sum(panel$ever_treated[panel$month_idx == 1] == 0)))
cat(sprintf("Treatment cohorts span month_idx %d to %d\n",
            min(panel$treatment_cohort[panel$treatment_cohort > 0]),
            max(panel$treatment_cohort[panel$treatment_cohort > 0])))
```

# Descriptive Analysis

## Treatment Timing

```{r treatment_timing, fig.width=8, fig.height=3}
cohort_counts <- panel %>%
  filter(ever_treated == 1) %>%
  distinct(full_name, treatment_month) %>%
  count(treatment_month, name = "n_repos")

p_timing <- ggplot(cohort_counts, aes(x = treatment_month, y = n_repos)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = n_repos), vjust = -0.4, size = 3.5,
            family = "Linux Libertine") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Treatment Month", y = "Number of Repos") +
  theme_panel() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_timing)

ggsave(file.path(plots_dir, "treatment_timing.pdf"),
       p_timing, width = 8, height = 3, device = cairo_pdf)
```

Treatment adoption is staggered across 2024-11 through 2026-03, with the
bulk of adoption occurring in 2025 and early 2026.  This staggered structure
is what motivates the modern DiD estimator in Stage 2.

## Average Trajectories

```{r trajectories, fig.width=8, fig.height=3}
trajectory <- panel %>%
  group_by(month, ever_treated) %>%
  summarise(
    mean_commits = mean(commits, na.rm = TRUE),
    mean_log     = mean(log_commits, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(group = ifelse(ever_treated == 1,
                        "Ever-treated (will adopt AI tools)",
                        "Never-treated (control)"))

ggplot(trajectory, aes(x = month, y = mean_log,
                       color = group, group = group)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  labs(x = "Month", y = "Mean log(1 + Commits)",
       color = NULL,
       title = "Average Monthly Commits by Treatment Group") +
  theme_panel() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The gap between ever-treated and never-treated repos is visible *throughout
the entire panel* --- including well before any repo adopted AI tools.  This
confirms the cross-sectional finding: repos that eventually adopt AI tools
are systematically more active.  The key question for DiD is whether the
*trajectories are parallel*, not whether the levels are equal.

## Pre-Treatment Covariate Balance

```{r balance, fig.width=8, fig.height=4}
# Use the first month as a baseline snapshot for time-invariant comparison
baseline <- panel %>%
  filter(month_idx == 1) %>%
  mutate(
    log_baseline_commits = log_commits,
    log_baseline_contrib = log_contributors
  )

balance_vars <- c("repo_age_yr", "is_org", "log_baseline_commits",
                  "log_baseline_contrib")
balance_labels <- c(
  "repo_age_yr"            = "Repo age (years)",
  "is_org"                 = "Organization owner",
  "log_baseline_commits"   = "log(Baseline commits)",
  "log_baseline_contrib"   = "log(Baseline contributors)"
)

compute_smd <- function(data, var, treat_col = "ever_treated") {
  t_vals <- data[[var]][data[[treat_col]] == 1]
  c_vals <- data[[var]][data[[treat_col]] == 0]
  pooled_sd <- sqrt((var(t_vals, na.rm = TRUE) +
                     var(c_vals, na.rm = TRUE)) / 2)
  if (pooled_sd < 1e-10) return(0)
  (mean(t_vals, na.rm = TRUE) - mean(c_vals, na.rm = TRUE)) / pooled_sd
}

smd_df <- data.frame(
  variable = balance_vars,
  label = balance_labels[balance_vars],
  smd = sapply(balance_vars, function(v) compute_smd(baseline, v))
) %>%
  mutate(label = reorder(label, abs(smd)))

ggplot(smd_df, aes(x = smd, y = label)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dotted",
             color = "red", alpha = 0.5) +
  labs(x = "Standardised Mean Difference (Ever-Treated vs Never-Treated)",
       y = NULL,
       title = "Baseline Covariate Balance",
       caption = "Dotted red lines mark the ±0.1 SMD threshold") +
  theme_panel() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```

```{r balance_table}
smd_df %>%
  arrange(desc(abs(smd))) %>%
  mutate(smd = sprintf("%.3f", smd)) %>%
  select(Covariate = label, SMD = smd) %>%
  kable(caption = "Baseline Standardised Mean Differences") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The balance diagnostics show that ever-treated and never-treated repos differ
at baseline --- the same selection story as the cross-section.  DiD does *not*
require baseline balance; it requires *parallel trends*.  But large baseline
differences make parallel trends less plausible a priori, which is why we
test pre-trends carefully in the event studies below.

# Stage 0: Naive Before/After Comparison

For each ever-treated repo, compare mean monthly commits before vs. after
adoption.  This is the simplest possible estimate --- and the most biased,
because it attributes all temporal trends to the treatment.

```{r stage0}
ba <- panel %>%
  filter(ever_treated == 1, !is.na(event_time)) %>%
  mutate(post = as.integer(event_time >= 0)) %>%
  group_by(full_name, post) %>%
  summarise(mean_commits = mean(commits, na.rm = TRUE),
            mean_log     = mean(log_commits, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = post,
              values_from = c(mean_commits, mean_log),
              names_sep = "_") %>%
  mutate(
    diff_raw = mean_commits_1 - mean_commits_0,
    diff_log = mean_log_1 - mean_log_0
  )

cat("=== Naive Before/After (Treated Repos Only) ===\n")
cat(sprintf("N repos: %d\n", nrow(ba)))
cat(sprintf("Mean pre-treatment commits:  %.1f\n", mean(ba$mean_commits_0)))
cat(sprintf("Mean post-treatment commits: %.1f\n", mean(ba$mean_commits_1)))
cat(sprintf("Mean within-repo change:     %.1f commits/month\n",
            mean(ba$diff_raw)))
cat(sprintf("Mean change in log scale:    %.3f\n", mean(ba$diff_log)))

# Paired t-test on log scale
t_ba <- t.test(ba$mean_log_1, ba$mean_log_0, paired = TRUE)
cat(sprintf("Paired t-test (log): t = %.2f, p = %.4f, 95%% CI [%.3f, %.3f]\n",
            t_ba$statistic, t_ba$p.value,
            t_ba$conf.int[1], t_ba$conf.int[2]))
```

```{r stage0_plot, fig.width=8, fig.height=3}
ggplot(ba, aes(x = diff_log)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(ba$diff_log), color = "red",
             linetype = "dashed", linewidth = 0.8) +
  labs(x = "Within-Repo Change in log(1 + Monthly Commits)",
       y = "Number of Repos",
       title = "Naive Before/After: Distribution of Within-Repo Changes",
       caption = "Red dashed line = mean change. No control group; confounded by secular trends.") +
  theme_panel()
```

The naive before/after comparison shows the raw within-repo change but has no
control group.  If all repos experienced a secular trend over 2024--2026
(e.g., growing commit activity due to platform growth), the estimate
confounds the treatment effect with the trend.

# Stage 1: Interrupted Time Series (No Control Group)

ITS models the pre-treatment trend explicitly and tests for a level shift
and/or slope change at treatment onset.  Still no control group.

```{r stage1}
# Restrict to ever-treated repos
its_data <- panel %>%
  filter(ever_treated == 1) %>%
  mutate(
    post = as.integer(treated == 1),
    # Time since treatment (0 for pre-treatment periods)
    time_since = ifelse(post == 1, event_time, 0)
  )

# ITS with repo FE: level shift + slope change
its_model <- feols(
  log_commits ~ month_idx + post + time_since | repo_id,
  data = its_data,
  cluster = ~repo_id
)

cat("=== ITS Model (Treated Repos Only, Repo FE) ===\n")
summary(its_model)
```

```{r stage1_plot, fig.width=8, fig.height=3}
# Distribution of log(commits) by event time for treated repos
its_boxdata <- panel %>%
  filter(ever_treated == 1, !is.na(event_time),
         event_time >= -12, event_time <= 12) %>%
  mutate(period = ifelse(event_time < 0, "Pre-treatment", "Post-treatment"),
         period = factor(period, levels = c("Pre-treatment", "Post-treatment")))

p_its <- ggplot(its_boxdata, aes(x = factor(event_time), y = log_commits,
                                 fill = period)) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3, linewidth = 0.3) +
  geom_vline(xintercept = 12.5, linetype = "dashed", color = "grey50") +
  scale_fill_manual(values = c("Pre-treatment" = "grey80",
                               "Post-treatment" = "steelblue")) +
  labs(x = "Months Since AI Tool Adoption (Event Time)",
       y = "log(1 + Commits)",
       fill = NULL,
       caption = paste0("Treated repos only (n = 240). Each repo's timeline is ",
                        "aligned to its own adoption date.\n",
                        "No control group; contemporaneous shocks remain confounded.")) +
  theme_panel() +
  theme(axis.text.x = element_text(size = rel(0.75)))

print(p_its)

ggsave(file.path(plots_dir, "its_boxplot.pdf"),
       p_its, width = 8, height = 3, device = cairo_pdf)
```

The ITS level-shift coefficient (`post`) tests whether commits jump
discretely at adoption.  The slope-change coefficient (`time_since`) tests
whether the post-treatment trajectory differs from the pre-treatment trend.
However, without a control group, we cannot rule out that a contemporaneous
shock (e.g., a GitHub platform change, a seasonal pattern) drives the result.

# Stage 2: Borusyak et al. Imputation DiD

Unlike the classic TWFE DiD estimator---which suffers from negative-weighting
bias under staggered treatment with heterogeneous effects (Goodman-Bacon
2021; de Chaisemartin & D'Haultfoeuille 2020; see Roth et al. 2023 for a
survey)---the imputation estimator of Borusyak, Jaravel, and Spiess (2024)
avoids this problem entirely.  It fits a model for Y(0) using only untreated
observations (never-treated + not-yet-treated), imputes the counterfactual
for each treated (i, t) cell, and averages the residuals.

```{r stage2_prep}
# didimputation requires: idname, tname, gname (cohort: 0 = never-treated),
# yname, and optionally first_stage for covariates.
# Ensure data types are correct
did_data <- panel %>%
  mutate(
    repo_id = as.integer(repo_id),
    month_idx = as.integer(month_idx),
    treatment_cohort = as.integer(treatment_cohort)
  ) %>%
  as.data.frame()

cat(sprintf("Panel for didimputation: %d obs, %d repos, %d months\n",
            nrow(did_data), n_distinct(did_data$repo_id),
            n_distinct(did_data$month_idx)))
cat(sprintf("Treatment cohorts: %d (0 = never-treated)\n",
            n_distinct(did_data$treatment_cohort)))
```

## Static ATT

```{r stage2_static}
# Static ATT (single average treatment effect)
did_static <- did_imputation(
  data = did_data,
  yname = "log_commits",
  gname = "treatment_cohort",
  tname = "month_idx",
  idname = "repo_id",
  first_stage = ~ log_contributors + log_stars + log_issues +
                  log_forks + log_releases,
  cluster_var = "repo_id"
)

cat("=== Borusyak et al. Imputation DiD: Static ATT ===\n")
print(did_static)
```

## Event Study

```{r stage2_event}
# Event study with pre-trends
did_es <- did_imputation(
  data = did_data,
  yname = "log_commits",
  gname = "treatment_cohort",
  tname = "month_idx",
  idname = "repo_id",
  first_stage = ~ log_contributors + log_stars + log_issues +
                  log_forks + log_releases,
  horizon = TRUE,
  pretrends = TRUE,
  cluster_var = "repo_id"
)

cat("=== Borusyak et al. Event Study ===\n")
print(did_es)
```

```{r stage2_esplot_prep}
# Prepare Borusyak event study data for combined plot
es_borusyak <- as.data.frame(did_es) %>%
  mutate(
    event_time = as.integer(as.character(term)),
    estimator = "Borusyak (with covariates)"
  ) %>%
  filter(event_time >= -12, event_time <= 12)
```

## Callaway & Sant'Anna (2021) DiD

We now estimate the same ATT using the Callaway & Sant'Anna (2021)
estimator **without covariates** (`xformla = ~1`).  This deliberately
omits the time-varying covariates that Borusyak includes in its
first-stage imputation model.  The comparison is informative: if the
Borusyak estimate with time-varying covariates is substantially lower
than the CS estimate without covariates, this is consistent with
covariate attenuation --- the covariates absorb part of the causal
pathway because they are post-treatment mediators, not pure confounders.

We use the never-treated repos (n = 759) as the control group.
Without covariates in the model, every treatment cohort --- regardless
of size --- can be estimated as a simple difference in means, so no
cohorts need to be dropped.

```{r csdid_static}
cs_att <- att_gt(
  yname  = "log_commits",
  tname  = "month_idx",
  idname = "repo_id",
  gname  = "treatment_cohort",
  xformla = ~1,
  data   = did_data,
  est_method = "reg",
  control_group = "nevertreated",
  clustervars = "repo_id",
  print_details = FALSE
)

cs_agg <- aggte(cs_att, type = "simple", na.rm = TRUE)
cat("=== Callaway & Sant'Anna DiD: Aggregate ATT ===\n")
summary(cs_agg)
```

```{r csdid_event}
cs_es <- aggte(cs_att, type = "dynamic", na.rm = TRUE)
cat("=== Callaway & Sant'Anna DiD: Event Study ===\n")
summary(cs_es)
```

## Combined Event Study

```{r combined_esplot, fig.width=8, fig.height=3}
# Prepare CS event study data
es_cs <- data.frame(
  event_time = cs_es$egt,
  estimate   = cs_es$att.egt,
  std.error  = cs_es$se.egt
) %>%
  mutate(
    conf.low  = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    estimator = "Callaway & Sant'Anna (no covariates)"
  ) %>%
  filter(event_time >= -12, event_time <= 12)

# Combine both estimators
es_combined <- bind_rows(
  es_borusyak %>% select(event_time, estimate, conf.low, conf.high, estimator),
  es_cs       %>% select(event_time, estimate, conf.low, conf.high, estimator)
)

dodge <- position_dodge(width = 0.4)

p_event <- ggplot(es_combined, aes(x = event_time, y = estimate,
                                   color = estimator, fill = estimator)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "grey70") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.12, color = NA, position = dodge) +
  geom_point(size = 1.8, position = dodge) +
  geom_line(linewidth = 0.5, position = dodge) +
  scale_color_manual(
    values = c("Borusyak (with covariates)" = "steelblue",
               "Callaway & Sant'Anna (no covariates)" = "#E15759")) +
  scale_fill_manual(
    values = c("Borusyak (with covariates)" = "steelblue",
               "Callaway & Sant'Anna (no covariates)" = "#E15759")) +
  labs(x = "Months Since AI Tool Adoption",
       y = "Estimated Effect on log(1 + Commits)",
       color = NULL, fill = NULL) +
  theme_panel()

print(p_event)

ggsave(file.path(plots_dir, "event_study_comparison.pdf"),
       p_event, width = 8, height = 3, device = cairo_pdf)
```

## Pre-Trends Assessment

We assess the parallel trends assumption for each estimator.  For
Borusyak, we use a joint Wald test on the pre-treatment event-study
coefficients (k = -12 to -1).  For Callaway & Sant'Anna, we use the
package's built-in pre-test (`Wpval`), which accounts for the
covariance structure across event-study coefficients and uses
simultaneous confidence bands to control the family-wise error rate.

```{r pretrends_test}
# --- Borusyak pre-trends ---
pre_borusyak <- es_borusyak %>% filter(event_time < 0)

cat("=== Borusyak Pre-Trends (k = -12 to -1) ===\n")
if (nrow(pre_borusyak) > 0) {
  cat(sprintf("  Max |coef|:  %.4f (at k = %d)\n",
              max(abs(pre_borusyak$estimate)),
              pre_borusyak$event_time[which.max(abs(pre_borusyak$estimate))]))
  cat(sprintf("  Mean |coef|: %.4f\n", mean(abs(pre_borusyak$estimate))))

  n_sig_bj <- sum(pre_borusyak$conf.low > 0 | pre_borusyak$conf.high < 0)
  cat(sprintf("  Periods with pointwise CI excluding zero: %d / %d\n",
              n_sig_bj, nrow(pre_borusyak)))

  pre_coefs_bj <- pre_borusyak$estimate
  pre_se_bj    <- pre_borusyak$std.error
  wald_bj <- sum((pre_coefs_bj / pre_se_bj)^2)
  df_bj   <- length(pre_coefs_bj)
  pval_bj <- pchisq(wald_bj, df = df_bj, lower.tail = FALSE)
  cat(sprintf("  Joint Wald chi2(%d) = %.2f, p = %.4f\n", df_bj, wald_bj, pval_bj))
}

# --- CS pre-trends: use the did package's built-in pre-test ---
cat("\n=== Callaway & Sant'Anna Pre-Trends ===\n")

# Built-in Wald pre-test (may be NULL if covariance matrix is singular)
if (!is.null(cs_es$Wpval)) {
  cat(sprintf("  Built-in Wald pre-test p-value: %.4f\n", cs_es$Wpval))
} else {
  cat("  Built-in Wald pre-test: not available (singular covariance matrix)\n")
}

# Simultaneous confidence bands (critical value stored in cs_es$crit.val.egt)
pre_cs <- es_cs %>% filter(event_time < 0)
crit_val <- cs_es$crit.val.egt
if (nrow(pre_cs) > 0) {
  cat(sprintf("  Critical value for simultaneous bands: %.3f (vs. 1.96 pointwise)\n",
              crit_val))
  n_sig_simult <- sum(abs(pre_cs$estimate) > crit_val * pre_cs$std.error)
  cat(sprintf("  Periods with simult. band excluding zero (k = -12 to -1): %d / %d\n",
              n_sig_simult, nrow(pre_cs)))

  cat(sprintf("  Max |coef|:  %.4f (at k = %d)\n",
              max(abs(pre_cs$estimate)),
              pre_cs$event_time[which.max(abs(pre_cs$estimate))]))
  cat(sprintf("  Mean |coef|: %.4f\n", mean(abs(pre_cs$estimate))))
}
```

## Why Do Borusyak and CS Estimates Diverge?

The two estimators differ along two dimensions that can each shift the
point estimate.

**Covariate attenuation in Borusyak.** The Borusyak first-stage model is
$Y_{it} = \alpha_i + \delta_t + X_{it}\beta + \varepsilon_{it}$,
estimated on untreated observations only.  When *imputing* the
counterfactual $Y(0)$ for a treated $(i, t)$ cell, it plugs in the
**post-treatment values** of $X_{it}$.  If AI adoption causes an
increase in contributors or stars, and those in turn drive more commits,
the imputed $Y(0)$ is inflated, shrinking the residual
(actual $-$ imputed) and **attenuating** the ATT estimate.  The CS
specification omits these covariates entirely, so it is unaffected by
this mechanism.

**Aggregation weights.** Borusyak weights each treated $(i, t)$ cell
equally, while CS (`type = "simple"`) first averages within
cohort-time cells, then across cohorts weighted by cohort size.  With
heterogeneous effects across cohorts, different weighting schemes
produce different aggregates.

The gap between the Borusyak ATT and the CS ATT is consistent with
both mechanisms: covariate attenuation pulls Borusyak down, and
different aggregation weights shift the two estimates further apart.
Together, the pair of estimates brackets a plausible range for the ATT.

# Control Group Assessment

Do we need to match the control group?  We assess this through three
diagnostics: (1) the event-study pre-trends from Stage 2 (already shown),
(2) pre-treatment trajectory comparison, and (3) robustness to trimming
structurally inactive repos.

## Pre-Treatment Trajectory Comparison

```{r pretreat_trajectory, fig.width=8, fig.height=3}
# Earliest treatment month is 2024-11 (month_idx ~ 11)
# Use months 1-10 (2024-01 to 2024-10) as the "pure pre-treatment" window
earliest_treat_idx <- min(panel$treatment_cohort[panel$treatment_cohort > 0])

pre_traj <- panel %>%
  filter(month_idx < earliest_treat_idx) %>%
  group_by(month, month_idx, ever_treated) %>%
  summarise(mean_log = mean(log_commits, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(group = ifelse(ever_treated == 1,
                        "Ever-treated", "Never-treated"))

ggplot(pre_traj, aes(x = month_idx, y = mean_log,
                     color = group, group = group)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  labs(x = "Month Index",
       y = "Mean log(1 + Commits)",
       color = NULL,
       title = "Pre-Treatment Trajectories (Before Any Repo Adopts)",
       caption = paste0("Months 1-", earliest_treat_idx - 1,
                        " (before any treatment). Parallel = good for DiD.")) +
  theme_panel()
```

## Trimming Inactive Repos

Many never-treated repos may be structurally dormant (zero commits throughout
the panel).  These contribute trivially parallel trends (flat at zero) but
dilute the comparison.  We test robustness by dropping repos with fewer than
10 total commits across the entire panel.

```{r trim_inactive}
# Identify active repos
repo_total <- panel %>%
  group_by(full_name, repo_id, ever_treated) %>%
  summarise(total_commits = sum(commits, na.rm = TRUE), .groups = "drop")

cat("Repos by activity level:\n")
cat(sprintf("  Total commits = 0:  %d repos (%d control, %d treated)\n",
            sum(repo_total$total_commits == 0),
            sum(repo_total$total_commits == 0 & repo_total$ever_treated == 0),
            sum(repo_total$total_commits == 0 & repo_total$ever_treated == 1)))
cat(sprintf("  Total commits < 10: %d repos (%d control, %d treated)\n",
            sum(repo_total$total_commits < 10),
            sum(repo_total$total_commits < 10 & repo_total$ever_treated == 0),
            sum(repo_total$total_commits < 10 & repo_total$ever_treated == 1)))
cat(sprintf("  Total commits >= 10: %d repos (%d control, %d treated)\n",
            sum(repo_total$total_commits >= 10),
            sum(repo_total$total_commits >= 10 & repo_total$ever_treated == 0),
            sum(repo_total$total_commits >= 10 & repo_total$ever_treated == 1)))

# Re-run Borusyak on trimmed sample
active_repos <- repo_total %>% filter(total_commits >= 10)
did_data_trim <- did_data %>% filter(repo_id %in% active_repos$repo_id)

cat(sprintf("\nTrimmed panel: %d obs, %d repos\n",
            nrow(did_data_trim), n_distinct(did_data_trim$repo_id)))
```

```{r trim_borusyak}
did_trim <- did_imputation(
  data = did_data_trim,
  yname = "log_commits",
  gname = "treatment_cohort",
  tname = "month_idx",
  idname = "repo_id",
  first_stage = ~ log_contributors + log_stars + log_issues +
                  log_forks + log_releases,
  cluster_var = "repo_id"
)

cat("=== Borusyak DiD on Trimmed Sample (>= 10 total commits) ===\n")
print(did_trim)
```

## Propensity Score Matching

The parallel trends assumption is more plausible when treated and control
units are comparable at baseline.  We use propensity score matching (PSM) on
pre-treatment observables to select a control group that resembles the
treated repos *before* adoption --- then run Borusyak DiD on this matched
panel.

The matching covariates are all computed from the pure pre-treatment window
(before any repo adopts, months 1 through `r earliest_treat_idx - 1`).
This avoids contaminating the propensity score with post-treatment outcomes.

```{r psm_data}
# Collapse to repo-level pre-treatment summary for matching
match_data <- panel %>%
  filter(month_idx < earliest_treat_idx) %>%
  group_by(full_name, repo_id, ever_treated, is_org, queried_language,
           repo_age_yr) %>%
  summarise(
    pre_mean_log_commits = mean(log_commits, na.rm = TRUE),
    pre_sd_log_commits   = sd(log_commits, na.rm = TRUE),
    pre_mean_contrib     = mean(log_contributors, na.rm = TRUE),
    pre_mean_stars       = mean(log_stars, na.rm = TRUE),
    pre_mean_issues      = mean(log_issues, na.rm = TRUE),
    pre_mean_forks       = mean(log_forks, na.rm = TRUE),
    pre_mean_releases    = mean(log_releases, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # Replace NA sd (repos with constant commits) with 0
    pre_sd_log_commits = ifelse(is.na(pre_sd_log_commits), 0,
                                pre_sd_log_commits)
  )

cat(sprintf("Matching pool: %d treated, %d control\n",
            sum(match_data$ever_treated == 1),
            sum(match_data$ever_treated == 0)))
```

```{r psm_estimate}
# Estimate propensity score via logistic regression
ps_formula <- ever_treated ~ pre_mean_log_commits + pre_sd_log_commits +
  pre_mean_contrib + pre_mean_stars + pre_mean_issues + pre_mean_forks +
  pre_mean_releases + is_org + repo_age_yr

ps_model <- glm(ps_formula, data = match_data, family = binomial)
cat("=== Propensity Score Model ===\n")
summary(ps_model)
```

```{r psm_overlap, fig.width=8, fig.height=3}
match_data$pscore <- predict(ps_model, type = "response")

ggplot(match_data, aes(x = pscore,
                       fill = factor(ever_treated,
                                     labels = c("Control", "Treated")))) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  labs(x = "Propensity Score", y = "Number of Repos", fill = NULL,
       title = "Propensity Score Distribution: Overlap Check",
       caption = "Good overlap is needed for matching. Regions without overlap are trimmed.") +
  theme_panel()
```

```{r psm_match, fig.width=8, fig.height=5}
# 1:1 nearest-neighbour matching on the propensity score
m_match <- matchit(
  ps_formula,
  data = match_data,
  method = "nearest",
  distance = "glm",
  ratio = 1,
  replace = FALSE,
  caliper = 0.2
)

summary(m_match)

match_labels <- c(
  "pre_mean_log_commits" = "Pre-treatment mean log(commits)",
  "pre_sd_log_commits"   = "Pre-treatment SD log(commits)",
  "pre_mean_contrib"     = "Pre-treatment mean log(contributors)",
  "pre_mean_stars"       = "Pre-treatment mean log(stars)",
  "pre_mean_issues"      = "Pre-treatment mean log(issues)",
  "pre_mean_forks"       = "Pre-treatment mean log(forks)",
  "pre_mean_releases"    = "Pre-treatment mean log(releases)",
  "is_org"               = "Organization owner",
  "repo_age_yr"          = "Repo age (years)"
)

love.plot(m_match, binary = "std", thresholds = c(m = 0.1),
          var.names = match_labels,
          colors = c("red", "steelblue"),
          shapes = c(17, 16),
          title = "Covariate Balance: Before vs After PSM") +
  theme_panel()
```

```{r psm_matched_trajectory, fig.width=8, fig.height=3}
# Matched sample: check pre-treatment trajectories
matched_repo_data <- match.data(m_match)
matched_ids <- as.integer(matched_repo_data$repo_id)
panel_matched <- panel %>% filter(repo_id %in% matched_ids)

cat(sprintf("Matched sample: %d treated + %d control = %d repos (%d panel obs)\n",
            sum(matched_repo_data$ever_treated == 1),
            sum(matched_repo_data$ever_treated == 0),
            nrow(matched_repo_data),
            nrow(panel_matched)))

matched_traj <- panel_matched %>%
  filter(month_idx < earliest_treat_idx) %>%
  group_by(month, month_idx, ever_treated) %>%
  summarise(mean_log = mean(log_commits, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(group = ifelse(ever_treated == 1,
                        "Ever-treated", "Matched control"))

ggplot(matched_traj, aes(x = month_idx, y = mean_log,
                         color = group, group = group)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  labs(x = "Month Index",
       y = "Mean log(1 + Commits)",
       color = NULL,
       title = "Pre-Treatment Trajectories After PSM",
       caption = paste0("Months 1-", earliest_treat_idx - 1,
                        ". Matched 1:1 NN on propensity score.")) +
  theme_panel()
```

## Borusyak DiD on PSM-Matched Panel

**Note:** This robustness check is included in the notebook for completeness
but is *not* presented in the paper.  The paper reports Borusyak DiD on the
full and trimmed samples; the matched-panel result is available here for
readers who wish to assess sensitivity to control group composition further.

```{r matched_borusyak}
did_data_matched <- did_data %>% filter(repo_id %in% matched_ids)

did_matched <- did_imputation(
  data = did_data_matched,
  yname = "log_commits",
  gname = "treatment_cohort",
  tname = "month_idx",
  idname = "repo_id",
  first_stage = ~ log_contributors + log_stars + log_issues +
                  log_forks + log_releases,
  cluster_var = "repo_id"
)

cat("=== Borusyak DiD on PSM-Matched Sample: Static ATT ===\n")
print(did_matched)
```

```{r matched_borusyak_es}
did_matched_es <- did_imputation(
  data = did_data_matched,
  yname = "log_commits",
  gname = "treatment_cohort",
  tname = "month_idx",
  idname = "repo_id",
  first_stage = ~ log_contributors + log_stars + log_issues +
                  log_forks + log_releases,
  horizon = TRUE,
  pretrends = TRUE,
  cluster_var = "repo_id"
)

cat("=== Borusyak Event Study on PSM-Matched Sample ===\n")
print(did_matched_es)
```

```{r matched_esplot, fig.width=8, fig.height=4}
es_matched_df <- as.data.frame(did_matched_es) %>%
  mutate(
    event_time = as.integer(as.character(term)),
    is_pre = event_time < 0
  ) %>%
  filter(event_time >= -12, event_time <= 12)

ggplot(es_matched_df, aes(x = event_time, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "grey70") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.15, fill = "#59A14F") +
  geom_point(aes(color = is_pre), size = 2) +
  geom_line(color = "#59A14F", linewidth = 0.6) +
  scale_color_manual(values = c("TRUE" = "grey50", "FALSE" = "#59A14F"),
                     labels = c("TRUE" = "Pre-treatment",
                                "FALSE" = "Post-treatment"),
                     name = NULL) +
  labs(x = "Months Since AI Tool Adoption",
       y = "Estimated Effect on log(1 + Commits)",
       title = "Borusyak DiD Event Study: PSM-Matched Sample",
       caption = "95% CIs. Matched 1:1 NN on propensity score + time-varying covariates.") +
  theme_panel()
```

The PSM-matched panel restricts the control group to the repos most
comparable to the treated units on pre-treatment observables.  If the
Borusyak ATT is similar across the full, trimmed, and PSM-matched samples,
the control group composition does not drive the result.  If the estimates
diverge, the parallel trends assumption may be sensitive to which controls
are included --- and the matched estimate is more credible because it
relies on a more comparable control group.

# Summary Comparison

```{r summary_table}
# Collect all estimates into a comparison table
results <- data.frame(
  Design = character(),
  Estimate = numeric(),
  SE = numeric(),
  CI_low = numeric(),
  CI_high = numeric(),
  N_obs = integer(),
  stringsAsFactors = FALSE
)

# Stage 0: Naive before/after
results <- rbind(results, data.frame(
  Design = "Stage 0: Naive before/after",
  Estimate = mean(ba$diff_log),
  SE = sd(ba$diff_log) / sqrt(nrow(ba)),
  CI_low = t_ba$conf.int[1],
  CI_high = t_ba$conf.int[2],
  N_obs = nrow(ba)
))

# Stage 1: ITS level shift
its_ct <- summary(its_model)$coeftable
its_idx <- which(rownames(its_ct) == "post")
results <- rbind(results, data.frame(
  Design = "Stage 1: ITS (level shift)",
  Estimate = its_ct[its_idx, "Estimate"],
  SE = its_ct[its_idx, "Std. Error"],
  CI_low = its_ct[its_idx, "Estimate"] - 1.96 * its_ct[its_idx, "Std. Error"],
  CI_high = its_ct[its_idx, "Estimate"] + 1.96 * its_ct[its_idx, "Std. Error"],
  N_obs = nobs(its_model)
))

# Stage 2: Borusyak (full sample)
bj <- as.data.frame(did_static)
results <- rbind(results, data.frame(
  Design = "Stage 2: Borusyak DiD (full)",
  Estimate = bj$estimate[1],
  SE = bj$std.error[1],
  CI_low = bj$conf.low[1],
  CI_high = bj$conf.high[1],
  N_obs = nrow(did_data)
))

# Stage 2: Borusyak (trimmed)
bt <- as.data.frame(did_trim)
results <- rbind(results, data.frame(
  Design = "Stage 2: Borusyak DiD (trimmed)",
  Estimate = bt$estimate[1],
  SE = bt$std.error[1],
  CI_low = bt$conf.low[1],
  CI_high = bt$conf.high[1],
  N_obs = nrow(did_data_trim)
))

# Stage 2: Borusyak (matched) -- not in paper
bm <- as.data.frame(did_matched)
results <- rbind(results, data.frame(
  Design = "Stage 2: Borusyak DiD (matched)*",
  Estimate = bm$estimate[1],
  SE = bm$std.error[1],
  CI_low = bm$conf.low[1],
  CI_high = bm$conf.high[1],
  N_obs = nrow(did_data_matched)
))

# Stage 2: Callaway & Sant'Anna
cs_est <- cs_agg$overall.att
cs_se  <- cs_agg$overall.se
results <- rbind(results, data.frame(
  Design = "Stage 2: Callaway & Sant'Anna DiD",
  Estimate = cs_est,
  SE = cs_se,
  CI_low = cs_est - 1.96 * cs_se,
  CI_high = cs_est + 1.96 * cs_se,
  N_obs = nrow(did_data)
))

results %>%
  mutate(
    Estimate = sprintf("%.3f", Estimate),
    SE       = sprintf("%.3f", SE),
    `95% CI` = sprintf("[%.3f, %.3f]", CI_low, CI_high),
    `N obs`  = format(N_obs, big.mark = ",")
  ) %>%
  select(Design, Estimate, SE, `95% CI`, `N obs`) %>%
  kable(caption = "Summary: All Estimates of the ATT of AI Tool Adoption on log(1 + Monthly Commits). * = not presented in paper.") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r summary_plot, fig.width=8, fig.height=4}
results$Design <- factor(results$Design, levels = rev(results$Design))

ggplot(results, aes(x = Estimate, y = Design)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high),
                  size = 0.6, color = "steelblue") +
  labs(x = "Estimated ATT on log(1 + Monthly Commits)",
       y = NULL,
       title = "Comparison of All Estimates",
       caption = "95% CIs. Progression from naive to credible designs.") +
  theme_panel() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```

# Interpretation

The progression from Stage 0 to Stage 2 illustrates the tutorial's core
message: *the design matters more than the estimator*.

- **Stage 0** (naive before/after) is the simplest longitudinal approach but
  confounds the treatment effect with any secular trend.

- **Stage 1** (ITS) improves on Stage 0 by explicitly modeling the
  pre-treatment trend, but still lacks a control group.  Without a
  counterfactual from untreated units, the estimate cannot distinguish the
  treatment effect from contemporaneous shocks (e.g., platform-wide changes,
  seasonal patterns) that affect all repositories around the same time.
  This limitation motivates the key insight of design-based identification:
  *modeling alone cannot substitute for the right research design*.

- **Stage 2** introduces a control group and exploits staggered adoption
  as quasi-random variation, using two complementary modern DiD estimators
  that avoid the negative-weighting problems of classic TWFE under
  staggered treatment with heterogeneous effects.  The **Borusyak et al.
  imputation estimator** fits an outcome model for $Y(0)$ and imputes
  counterfactuals, while the **Callaway & Sant'Anna outcome-regression
  estimator** conditions on covariate values at the pre-treatment base
  period for each cohort.

  The divergence between the two estimates is informative:  Borusyak's
  first-stage uses time-varying covariates that may be post-treatment
  mediators (e.g., contributors and stars that *increase because of* AI
  adoption), attenuating the ATT by absorbing part of the causal pathway.
  The Callaway & Sant'Anna specification omits these covariates entirely,
  so it is unaffected by this attenuation.  Additionally, the two
  estimators use different aggregation weights for heterogeneous
  cohort-level effects.  Together, the pair of estimates brackets a
  plausible range for the ATT.

  The robustness of the Borusyak ATT across the full, trimmed, and
  PSM-matched samples addresses concerns about control group composition.

The event-study plots serve a dual purpose: they test the parallel trends
assumption (pre-treatment coefficients near zero) *and* visualise the
dynamic treatment effect (post-treatment coefficients).

Compare the final Borusyak ATT with the cross-sectional coefficients from
the companion notebook.  The cross-sectional M_pretreat coefficient reflects
both the causal effect and selection bias (productive repos adopt AI tools).
The longitudinal ATT, by absorbing within-repo variation, strips away the
selection component --- the difference between the two estimates is an
upper bound on the magnitude of selection bias in the cross-section.
