---
title: "Cross-Sectional Analysis: AI Coding Tool Adoption and Development Productivity"
output: html_document
---

This notebook implements the cross-sectional analysis of AI coding tool
adoption for Section 4.1 of the paper.  The analysis follows a three-stage
arc that starts from what a typical SE researcher would do, diagnoses why
it fails, and shows that switching estimators cannot help:

- **Stage 1** starts from the typical approach: a naive comparison and a
  kitchen-sink regression.  The raw gap is enormous and the kitchen-sink
  coefficient is large and highly significant.  A naive analyst might stop
  here.
- **Stage 2** diagnoses why we should distrust the kitchen-sink result.
  The key insight is **temporal collapse**: MSR snapshot data merges
  pre-treatment and post-treatment states of time-varying covariates
  (stars, PRs, issues, CI, size, releases) into a single node, destroying
  the acyclicity that DAGs require and making covariate selection
  incoherent.  We derive a **temporal DAG** from the paper's Figure 3 and
  use the panel dataset to recover genuinely pre-treatment values of
  time-varying covariates, yielding a DAG-justified adjustment set that
  goes far beyond the three structural covariates alone.
- **Stage 3** shows that OLS, PSM, and IPW converge on the same estimate
  under pre-treatment covariates --- the estimator is not the bottleneck;
  the identifying assumption is.

The takeaway: cross-sectional MSR data cannot credibly answer causal
questions about AI tool adoption.  Only within-repo longitudinal variation
can make progress (Section 4.2).

# Setup

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "knitr", "kableExtra",
                   "lmtest", "sandwich", "MatchIt", "cobalt",
                   "showtext", "scales", "dagitty")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(lmtest)
library(sandwich)
library(MatchIt)
library(cobalt)
library(showtext)
library(scales)
library(dagitty)

font_add("Linux Libertine", regular = "LinLibertine_Rah.ttf")
showtext_auto()
```

# Load and Prepare Data

```{r load_data, message=FALSE, warning=FALSE}
repos <- read.csv("../data/ai_adoption_repos.csv", stringsAsFactors = FALSE)
monthly <- read.csv("../data/ai_adoption_monthly_commits.csv",
                     stringsAsFactors = FALSE)

cat(sprintf("Repos: %d\n", nrow(repos)))
cat(sprintf("Monthly commit rows: %d\n", nrow(monthly)))
```

```{r prepare_data}
df <- repos %>%
  mutate(
    # Treatment: L2+ vs L1
    treated     = as.integer(ai_maturity_level >= 2),
    # Derive repo age in years from creation date
    created_at  = as.POSIXct(created_at, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    pushed_at   = as.POSIXct(pushed_at,  format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    repo_age_yr = as.numeric(difftime(Sys.time(), created_at, units = "days")) / 365.25,
    # Log-transform skewed covariates (add 1 to handle zeros)
    log_stars        = log1p(stars),
    log_forks        = log1p(forks),
    log_size         = log1p(size_kb),
    log_total_commits = log1p(total_commits),
    log_total_prs    = log1p(total_prs),
    log_total_issues = log1p(total_issues),
    log_releases     = log1p(total_releases),
    log_open_issues  = log1p(open_issues),
    # Binary covariates
    is_org       = as.integer(owner_type == "Organization"),
    has_ci       = as.integer(has_ci == "True" | has_ci == TRUE),
    has_wiki     = as.integer(has_wiki == "True" | has_wiki == TRUE),
    has_discuss   = as.integer(has_discussions == "True" | has_discussions == TRUE)
  )

# Compute average monthly commits from the 6-month window columns
commit_cols <- grep("^commits_", names(df), value = TRUE)
df$avg_monthly_commits <- rowMeans(df[, commit_cols], na.rm = TRUE)
df$log_avg_monthly_commits <- log1p(df$avg_monthly_commits)

# Total commits in the 6-month window
df$total_window_commits <- rowSums(df[, commit_cols], na.rm = TRUE)
df$log_total_window_commits <- log1p(df$total_window_commits)

cat(sprintf("Treatment (L2+): %d  |  Control (L1): %d\n",
            sum(df$treated == 1), sum(df$treated == 0)))
cat(sprintf("Treatment prevalence: %.1f%%\n", mean(df$treated) * 100))
```

## Compute Pre-Treatment Covariates from Panel Data

The panel dataset provides monthly observations (commits, stars, issues,
forks, PRs merged, releases, active contributors) for each repository
across the full observation window (Jan 2024 -- Mar 2026).  For treated
repos the exact treatment month is known; for control repos all months
are pre-treatment.  We aggregate only pre-treatment months to obtain
genuinely pre-treatment values of time-varying covariates that the
cross-sectional snapshot collapses.

```{r panel_pre_treatment, message=FALSE, warning=FALSE}
panel <- read.csv("../data/ai_adoption_panel.csv", stringsAsFactors = FALSE)

pre_treat_covs <- panel %>%
  group_by(full_name) %>%
  mutate(
    repo_trt_month = first(na.omit(
      ifelse(treatment_month == "", NA_character_, treatment_month)
    ))
  ) %>%
  ungroup() %>%
  mutate(
    is_pre = is.na(repo_trt_month) | month < repo_trt_month
  ) %>%
  filter(is_pre) %>%
  group_by(full_name) %>%
  summarise(
    pre_stars        = sum(new_stars, na.rm = TRUE),
    pre_issues       = sum(new_issues, na.rm = TRUE),
    pre_forks        = sum(new_forks, na.rm = TRUE),
    pre_prs_merged   = sum(prs_merged, na.rm = TRUE),
    pre_releases     = sum(new_releases, na.rm = TRUE),
    pre_avg_contribs = mean(active_contributors, na.rm = TRUE),
    pre_avg_commits  = mean(commits, na.rm = TRUE),
    pre_months       = n(),
    .groups = "drop"
  )

df <- df %>%
  left_join(pre_treat_covs, by = "full_name") %>%
  mutate(
    log_pre_stars    = log1p(pre_stars),
    log_pre_issues   = log1p(pre_issues),
    log_pre_forks    = log1p(pre_forks),
    log_pre_prs      = log1p(pre_prs_merged),
    log_pre_releases = log1p(pre_releases),
    log_pre_contribs = log1p(pre_avg_contribs),
    log_pre_commits  = log1p(pre_avg_commits)
  )

n_matched  <- sum(!is.na(df$pre_stars))
n_missing  <- sum(is.na(df$pre_stars))
n_miss_trt <- sum(is.na(df$pre_stars) & df$treated == 1)
n_miss_ctl <- sum(is.na(df$pre_stars) & df$treated == 0)

cat(sprintf("Repos with panel-derived pre-treatment covariates: %d of %d\n",
            n_matched, nrow(df)))
cat(sprintf("Repos missing panel data: %d (treated: %d, control: %d)\n",
            n_missing, n_miss_trt, n_miss_ctl))
if (n_matched < nrow(df)) {
  cat("  -> These repos are either absent from the panel dataset or\n")
  cat("     were treated in the first panel month (no pre-treatment data).\n")
  cat("  -> DAG-justified analyses will use the matched subset.\n")
}
cat(sprintf("Treated repos — median pre-treatment months: %.0f\n",
            median(df$pre_months[df$treated == 1], na.rm = TRUE)))
cat(sprintf("Control repos — median pre-treatment months: %.0f\n",
            median(df$pre_months[df$treated == 0], na.rm = TRUE)))

# Clean subset for DAG-justified analyses
df_dag <- df %>% filter(!is.na(log_pre_stars))
cat(sprintf("DAG-justified analysis sample: %d repos (%d treated, %d control)\n",
            nrow(df_dag), sum(df_dag$treated == 1), sum(df_dag$treated == 0)))
```

# Stage 1: What Would a Typical SE Researcher Do?

A typical MSR study would start by comparing treated and control groups,
then run a regression with all available covariates.  We follow this
approach first, then diagnose why it fails.

## 1.1 Descriptive Comparison (Table 1)

Table 1 presents the outcome and all covariates side by side for the
control (L1: no AI config) and treated (L2+: AI config present) groups,
with standardised mean differences (SMDs).  Covariates are grouped by
their temporal status --- a distinction we will motivate in Stage 2.

```{r table1_descriptive}
# Helper: compute SMD
compute_smd <- function(data, var, treat_col = "treated") {
  t_vals <- data[[var]][data[[treat_col]] == 1]
  c_vals <- data[[var]][data[[treat_col]] == 0]
  pooled_sd <- sqrt((var(t_vals, na.rm = TRUE) + var(c_vals, na.rm = TRUE)) / 2)
  if (pooled_sd < 1e-10) return(0)
  (mean(t_vals, na.rm = TRUE) - mean(c_vals, na.rm = TRUE)) / pooled_sd
}

# Create language dummy variables
languages <- sort(unique(df$queried_language))
for (lang in languages) {
  col_name <- paste0("lang_", gsub("[^A-Za-z]", "", lang))
  df[[col_name]] <- as.integer(df$queried_language == lang)
}
lang_vars <- paste0("lang_", gsub("[^A-Za-z]", "", languages))
lang_labels <- setNames(languages, lang_vars)
lang_temporal <- setNames(rep("Pre-treatment", length(languages)), lang_vars)

# Define variables, labels, and temporal status
table1_vars <- c(
  "avg_monthly_commits",
  "repo_age_yr", "is_org",
  lang_vars,
  "log_stars", "log_forks", "log_size",
  "log_total_commits", "log_total_prs", "log_total_issues",
  "log_releases", "log_open_issues",
  "has_ci", "has_wiki", "has_discuss"
)

table1_labels <- c(
  "avg_monthly_commits" = "Avg monthly commits",
  "repo_age_yr"         = "Repo age (years)",
  "is_org"              = "Organization owner",
  lang_labels,
  "log_stars"           = "log(Stars)",
  "log_forks"           = "log(Forks)",
  "log_size"            = "log(Size KB)",
  "log_total_commits"   = "log(Total commits)",
  "log_total_prs"       = "log(Total PRs)",
  "log_total_issues"    = "log(Total issues)",
  "log_releases"        = "log(Releases)",
  "log_open_issues"     = "log(Open issues)",
  "has_ci"              = "Has CI/CD",
  "has_wiki"            = "Has wiki",
  "has_discuss"         = "Has discussions"
)

table1_temporal <- c(
  "avg_monthly_commits" = "Outcome",
  "repo_age_yr"         = "Pre-treatment",
  "is_org"              = "Pre-treatment",
  lang_temporal,
  "log_stars"           = "Time-varying",
  "log_forks"           = "Time-varying",
  "log_size"            = "Time-varying",
  "log_total_commits"   = "Time-varying",
  "log_total_prs"       = "Time-varying",
  "log_total_issues"    = "Time-varying",
  "log_releases"        = "Time-varying",
  "log_open_issues"     = "Time-varying",
  "has_ci"              = "Time-varying",
  "has_wiki"            = "Time-varying",
  "has_discuss"         = "Time-varying"
)

table1_df <- data.frame(
  variable = table1_vars,
  label    = table1_labels[table1_vars],
  temporal = table1_temporal[table1_vars],
  stringsAsFactors = FALSE
) %>%
  rowwise() %>%
  mutate(
    control_mean = mean(df[[variable]][df$treated == 0], na.rm = TRUE),
    treated_mean = mean(df[[variable]][df$treated == 1], na.rm = TRUE),
    smd          = compute_smd(df, variable)
  ) %>%
  ungroup()

table1_df %>%
  mutate(
    control_mean = sprintf("%.2f", control_mean),
    treated_mean = sprintf("%.2f", treated_mean),
    smd          = sprintf("%.3f", smd)
  ) %>%
  select(
    Variable = label,
    `Temporal Status` = temporal,
    `Control (L1) Mean` = control_mean,
    `Treated (L2+) Mean` = treated_mean,
    SMD = smd
  ) %>%
  kable(caption = "Table 1: Descriptive Comparison by AI Adoption Status (means and SMDs)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The treated group (L2+) differs from the control group (L1) on *every*
dimension.  Every covariate exceeds the conventional +/-0.1 SMD threshold.
The raw gap in monthly commits is enormous.  These are not random
differences --- they reflect a coherent selection mechanism:
well-resourced, actively maintained, organization-backed projects are the
ones that adopt AI tools.

The programming language distribution also differs markedly: TypeScript
and Go repositories are heavily represented among adopters, while C and
Swift barely participate.

## 1.2 Kitchen-Sink Regression

A typical SE researcher might now run a regression controlling for
everything available.

```{r kitchen_sink}
# Kitchen-sink: all available covariates
m_kitchen <- lm(log_avg_monthly_commits ~ treated + repo_age_yr +
                  queried_language + is_org + has_ci + log_size +
                  log_releases + log_stars + log_forks + log_total_prs +
                  log_total_issues,
                data = df)

# Extract treatment coefficient with HC1 robust SEs
extract_treated <- function(model, label) {
  ct <- coeftest(model, vcov = vcovHC(model, type = "HC1"))
  idx <- which(rownames(ct) == "treated")
  data.frame(
    model    = label,
    coef     = ct[idx, "Estimate"],
    se       = ct[idx, "Std. Error"],
    t_val    = ct[idx, "t value"],
    p_val    = ct[idx, "Pr(>|t|)"],
    r2       = summary(model)$r.squared,
    adj_r2   = summary(model)$adj.r.squared,
    n        = nobs(model)
  )
}

kitchen_result <- extract_treated(m_kitchen, "Kitchen-sink (all covariates)")

kitchen_result %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(as.numeric(p_val) < 0.001, "<0.001", sprintf("%.3f", p_val)),
    r2    = sprintf("%.3f", r2)
  ) %>%
  select(Model = model, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, `R²` = r2, N = n) %>%
  kable(caption = "Kitchen-Sink OLS: AI Adoption Effect on log(Monthly Commits)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The coefficient is positive and highly significant (p < 0.001), surviving
a battery of controls.  A naive analyst might conclude that AI adoption
*causes* higher productivity and stop here.  But should we trust this?

# Stage 2: Why Should We Distrust This?

## 2.1 The Temporal Collapse Problem

To reason about which covariates to include in a regression, we need a
causal DAG.  But the cross-sectional design introduces a fundamental
problem that is pervasive in MSR studies.

Consider a variable like "stars."  In the true data-generating process,
stars accumulate over time.  The actual causal DAG has time-indexed nodes:

- **stars\_{t-1}** (before AI adoption) is a **confounder**: popular repos
  attract AI-savvy contributors (stars\_{t-1} -> adoption\_t) *and* have
  more active contributor bases (stars\_{t-1} -> commits\_t).
- **stars\_{t+1}** (after AI adoption) is a **collider**: both AI adoption
  and high commit activity increase visibility and attract stars
  (adoption\_t -> stars\_{t+1} <- commits\_t).

In this true temporal DAG, the adjustment set is unambiguous: condition on
stars\_{t-1}, never on stars\_{t+1}.

But a typical MSR study queries the GitHub API once at the time of the
study.  This **collapses** both time-indexed nodes into a single "stars"
variable that contains both pre-treatment and post-treatment variation.
The resulting graph has both incoming and outgoing arrows between "stars"
and both treatment and outcome --- it is **cyclic**, not a DAG.  No
adjustment formula applies.

This temporal collapse is not specific to stars.  Every time-varying
covariate in our dataset (PRs, issues, CI presence, repo size, releases)
suffers from the same problem.  The true temporal DAG is well-defined;
the cross-sectional snapshot destroys it.

(See the paper's Figures X and Y for the true temporal DAG and the
collapsed cross-sectional graph.)

This connects to a recognized problem in the causal inference literature:
Hernan & Robins (2020, Ch. 20) describe this as "treatment-confounder
feedback," and Richardson & Robins have critiqued the use of
cross-sectional DAGs precisely because the nodes must be well-defined
temporal events, not time-collapsed summaries.

## 2.2 Covariate Classification

The temporal collapse argument divides our covariates into two groups:

```{r covariate_classification}
classif <- data.frame(
  Covariate = c("Repo age", "Primary language", "Organization owner",
                "Stars", "Forks", "Repo size (KB)", "Total commits",
                "Total PRs", "Total issues", "Releases", "Open issues",
                "CI/CD presence", "Wiki", "Discussions"),
  `Temporal Status` = c(rep("Pre-treatment", 3), rep("Temporally ambiguous", 11)),
  Justification = c(
    "Determined at creation; cannot be affected by AI adoption",
    "Determined at creation; cannot be affected by AI adoption",
    "Determined at creation; cannot be affected by AI adoption",
    "Accumulates before and after adoption",
    "Accumulates before and after adoption",
    "Includes post-adoption code growth",
    "Includes post-adoption commits",
    "Includes post-adoption PRs",
    "Includes post-adoption issues",
    "Includes post-adoption releases",
    "Snapshot mixes pre- and post-adoption state",
    "May be adopted as part of same modernization wave",
    "Snapshot mixes pre- and post-adoption state",
    "Snapshot mixes pre- and post-adoption state"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

classif %>%
  kable(caption = "Covariate Classification by Temporal Status") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

**Consequence:** Only the three structurally pre-treatment covariates
(repo age, language, organization type) are defensible controls *in the
cross-sectional snapshot*.  But with access to the panel dataset, we can
recover genuinely pre-treatment values of time-varying covariates,
restoring the clean temporal structure that the snapshot destroyed.

## 2.3 Temporal DAG and Panel-Derived Pre-Treatment Covariates

The paper's DAG (Figure 3) identifies confounders and a collider
(Stars).  As noted in the paper, Stars' role is temporally ambiguous: it
is a confounder *before* adoption (popular repos attract contributors
and adopt tools) and a collider *after* (both adoption and velocity
increase visibility).  The same ambiguity applies to all time-varying
characteristics of a project.

The **temporal DAG** resolves this by grouping time-varying covariates
into four high-level concepts --- **Popularity** (stars), **Demand**
(issues), **Activity** (forks, PRs, commits, releases), and **Team
Size** (active contributors) --- and splitting each into a pre-treatment
node (confounder) and a post-treatment node (collider).  The back-door
criterion applied to this temporal DAG yields an unambiguous adjustment
set.

```{r temporal_dag, fig.width=8, fig.height=6}
temporal_dag <- dagitty('dag {
  bb="0,0,1,1"

  "AI Adoption" [exposure, pos="0.30,0.45"]
  "Dev Velocity" [outcome, pos="0.70,0.45"]

  "Domain" [pos="0.08,0.05"]
  "Organization" [pos="0.28,0.05"]
  "Team Skill" [pos="0.50,0.05"]
  "Proj Maturity" [pos="0.72,0.05"]

  "Popularity_pre" [pos="0.15,0.22"]
  "Demand_pre" [pos="0.38,0.22"]
  "Activity_pre" [pos="0.62,0.22"]
  "Team Size_pre" [pos="0.85,0.22"]

  "Code Gen" [pos="0.45,0.55"]
  "Workflow" [pos="0.55,0.55"]

  "Popularity_post" [pos="0.15,0.75"]
  "Demand_post" [pos="0.38,0.75"]
  "Activity_post" [pos="0.62,0.75"]
  "Team Size_post" [pos="0.85,0.75"]

  "Domain" -> "AI Adoption"
  "Domain" -> "Dev Velocity"
  "Organization" -> "AI Adoption"
  "Organization" -> "Dev Velocity"
  "Team Skill" -> "AI Adoption"
  "Team Skill" -> "Dev Velocity"
  "Proj Maturity" -> "AI Adoption"
  "Proj Maturity" -> "Dev Velocity"

  "Popularity_pre" -> "AI Adoption"
  "Popularity_pre" -> "Dev Velocity"
  "Demand_pre" -> "AI Adoption"
  "Demand_pre" -> "Dev Velocity"
  "Activity_pre" -> "AI Adoption"
  "Activity_pre" -> "Dev Velocity"
  "Team Size_pre" -> "AI Adoption"
  "Team Size_pre" -> "Dev Velocity"

  "AI Adoption" -> "Code Gen" -> "Dev Velocity"
  "AI Adoption" -> "Workflow" -> "Dev Velocity"

  "AI Adoption" -> "Popularity_post"
  "Dev Velocity" -> "Popularity_post"
  "AI Adoption" -> "Demand_post"
  "Dev Velocity" -> "Demand_post"
  "AI Adoption" -> "Activity_post"
  "Dev Velocity" -> "Activity_post"
  "AI Adoption" -> "Team Size_post"
  "Dev Velocity" -> "Team Size_post"
}')

plot(temporal_dag)
```

Applying the **back-door criterion** to the temporal DAG:

```{r adjustment_sets}
adj_sets <- adjustmentSets(temporal_dag, type = "minimal")
cat("Minimal sufficient adjustment sets (back-door criterion):\n")
print(adj_sets)
```

Every pre-treatment node (both structurally pre-treatment and
panel-derived) is a confounder that blocks a back-door path.  Code Gen
and Workflow are mediators (on the causal path).  Every \*\_post node is a
collider --- conditioning on any of them would *open* a spurious path.

The key insight: **Team Skill remains entirely unobserved** --- no
variable in the panel or cross-sectional data captures developer skill
or expertise.  The panel-derived pre-treatment covariates (Popularity,
Demand, Activity, Team Size) absorb more of the selection than the
structurally pre-treatment variables alone, but Team Skill is the
confounder that no GitHub API data can address.

### Mapping DAG Nodes to Observable Covariates

```{r dag_mapping}
dag_map <- data.frame(
  `DAG Node` = c("Domain", "Organization", "Proj Maturity",
                 "Team Skill",
                 "Popularity_pre", "Demand_pre",
                 "Activity_pre", "Activity_pre", "Activity_pre", "Activity_pre",
                 "Team Size_pre"),
  `Observable Covariate` = c(
    "queried_language",
    "is_org",
    "repo_age_yr",
    "Unobserved",
    "log_pre_stars",
    "log_pre_issues",
    "log_pre_forks", "log_pre_prs", "log_pre_commits", "log_pre_releases",
    "log_pre_contribs"
  ),
  `DAG Role` = c(rep("Confounder", 11)),
  `Source` = c(rep("Cross-sectional (structural)", 3),
               "Unobserved",
               rep("Panel-derived (pre-treatment)", 7)),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dag_map %>%
  kable(caption = "Mapping Temporal DAG Nodes to Observable Covariates") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### DAG-Justified Regression

With the panel-derived pre-treatment covariates, we can now run a
regression that conditions on the full DAG-justified adjustment set
(minus Team Skill, which remains unobserved):

```{r dag_justified_regression}
m_dag <- lm(log_avg_monthly_commits ~ treated + repo_age_yr +
              queried_language + is_org +
              log_pre_stars + log_pre_issues + log_pre_forks +
              log_pre_prs + log_pre_releases + log_pre_contribs +
              log_pre_commits,
            data = df_dag)

dag_result <- extract_treated(m_dag, "DAG-justified (panel pre-treatment)")

dag_result %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(as.numeric(p_val) < 0.001, "<0.001", sprintf("%.3f", p_val)),
    r2    = sprintf("%.3f", r2)
  ) %>%
  select(Model = model, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, `R²` = r2, N = n) %>%
  kable(caption = "DAG-Justified OLS: Pre-Treatment Covariates from Panel Data") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The DAG-justified specification controls for substantially more of the
selection mechanism than the three structural covariates alone.
Pre-treatment Activity is particularly important: it captures baseline
development velocity, the strongest predictor of both AI adoption
(high-velocity projects have incentive to adopt) and future velocity
(persistence).  Including it effectively asks: *conditional on repos
having similar pre-treatment activity, popularity, demand, and team
size, do AI-adopting repos subsequently exhibit higher commit rates?*
Team Skill remains entirely unobserved --- no covariate in the data
captures developer expertise.

## 2.4 Bivariate, Kitchen-Sink, and DAG-Justified OLS (Table 2)

We compare three specifications: the bivariate regression (no controls),
the kitchen-sink model from Stage 1 (all available cross-sectional
covariates, many temporally ambiguous), and the DAG-justified model
(only pre-treatment covariates backed by the temporal DAG).

```{r ols_comparison}
# Bivariate
m_bivariate <- lm(log_avg_monthly_commits ~ treated, data = df)

ols_results <- bind_rows(
  extract_treated(m_bivariate, "Bivariate (no controls)"),
  extract_treated(m_kitchen,   "Kitchen-sink (all covariates)"),
  extract_treated(m_dag,       "DAG-justified (panel pre-treatment)")
)

ols_results %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(as.numeric(p_val) < 0.001, "<0.001", sprintf("%.3f", p_val)),
    r2    = sprintf("%.3f", r2)
  ) %>%
  select(Model = model, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, `R²` = r2, N = n) %>%
  kable(caption = "Table 2: OLS Estimates of AI Adoption Effect on log(Monthly Commits)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The bivariate coefficient is enormous, reflecting raw selection.  The
kitchen-sink coefficient attenuates after adding all available covariates,
but its interpretation is **uninterpretable**: the collapsed cross-sectional
graph is not a DAG, so no adjustment formula tells us whether we are
removing confounding or blocking causal paths through temporally ambiguous
variables.  The DAG-justified specification conditions only on covariates
that the temporal DAG identifies as pre-treatment confounders --- each
computed from panel data before the treatment month.  This is the only
specification with a principled justification from the back-door criterion.

## 2.5 Sensitivity Analysis: Oster Proportional Selection Bounds

How robust is the DAG-justified estimate to unobserved confounding?
We apply the Oster (2019) proportional selection method, which asks:
*given how much the coefficient changed when we added observable
controls, how strong would selection on unobservables need to be ---
relative to that same selection on observables --- to drive the
coefficient all the way to zero?*

The key quantity is $\delta^*$.  The conventional threshold is
$\delta^* \geq 1$: if unobservables would need to be *at least as
strong as* all the observables combined to eliminate the effect, the
result is considered robust.  If $\delta^* < 1$, the result is fragile.

```{r oster_bounds}
oster_delta <- function(model_restricted, model_full, r_max_multiplier = 1.3) {
  b_r <- coef(model_restricted)["treated"]
  b_f <- coef(model_full)["treated"]
  r_r <- summary(model_restricted)$r.squared
  r_f <- summary(model_full)$r.squared
  r_max <- min(1, r_max_multiplier * r_f)

  delta <- (b_f * (r_max - r_f)) / ((b_r - b_f) * (r_f - r_r))
  return(data.frame(
    beta_restricted = b_r,
    beta_full       = b_f,
    R2_restricted   = r_r,
    R2_full         = r_f,
    R2_max          = r_max,
    delta_star      = delta
  ))
}

oster_dag <- oster_delta(m_bivariate, m_dag)

oster_dag %>%
  mutate(across(where(is.numeric), ~sprintf("%.3f", .))) %>%
  mutate(comparison = "Bivariate -> DAG-justified (panel)") %>%
  select(Comparison = comparison, `beta restricted` = beta_restricted,
         `beta full` = beta_full, `R2 restricted` = R2_restricted,
         `R2 full` = R2_full, `R2 max` = R2_max, `delta*` = delta_star) %>%
  kable(caption = "Oster Proportional Selection Bounds") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The $\delta^*$ is well below 1, indicating that the estimate is **not
robust** to unobserved confounding.  Three reinforcing factors drive
this result:

1. **Massive coefficient drop**: the coefficient falls substantially
   from the bivariate to the DAG-justified model, meaning the
   observables already "explain away" most of the naive effect.
2. **Very high $R^2$**: the DAG-justified model explains most of the
   outcome variance, leaving little residual room for any additional
   variable --- observed or unobserved --- to exploit.
3. **Large $R^2$ jump**: the observables increase explanatory power
   dramatically from the bivariate to the full model, inflating the
   denominator and further shrinking $\delta^*$.

In plain language: given how much the coefficient already dropped and
how much variance the observables already explain, even a relatively
weak unobserved confounder could eliminate the remaining effect.  The
most important confounders --- developer skill and team culture --- are
precisely the ones we cannot observe from GitHub API data, and they are
plausibly strong enough to account for the residual association.

# Stage 3: Can a Better Estimator Help?

Stage 3 asks whether alternative selection-on-observables methods
(matching, IPW) yield different answers from OLS.  If the problem were
model misspecification (nonlinearity, poor overlap), matching or IPW might
help.  If the problem is unobserved confounding, all three estimators will
converge on the same biased answer --- because they share the same
identifying assumption (conditional ignorability given observables).

We use only the DAG-justified pre-treatment covariates, since they are
the only defensible adjustment set.

## 3.1 Propensity Score Matching

```{r psm_estimation}
ps_formula_dag <- treated ~ repo_age_yr + queried_language + is_org +
                    log_pre_stars + log_pre_issues + log_pre_forks +
                    log_pre_prs + log_pre_releases + log_pre_contribs +
                    log_pre_commits

m_ps <- matchit(ps_formula_dag, data = df_dag, method = "nearest",
                distance = "glm", ratio = 1, replace = FALSE)

summary(m_ps)
```

```{r psm_ols}
matched_data <- match.data(m_ps)

m_psm <- lm(log_avg_monthly_commits ~ treated + repo_age_yr +
               queried_language + is_org +
               log_pre_stars + log_pre_issues + log_pre_forks +
               log_pre_prs + log_pre_releases + log_pre_contribs +
               log_pre_commits,
             data = matched_data, weights = weights)

psm_ct <- coeftest(m_psm, vcov = vcovHC(m_psm, type = "HC1"))
idx <- which(rownames(psm_ct) == "treated")

psm_result <- data.frame(
  method = "PSM (1:1 NN)",
  coef   = psm_ct[idx, 1],
  se     = psm_ct[idx, 2],
  p_val  = psm_ct[idx, 4],
  n      = nrow(matched_data)
)
```

## 3.2 Inverse Probability Weighting

```{r ipw}
df_dag$ps <- predict(glm(ps_formula_dag, data = df_dag, family = binomial),
                     type = "response")

df_dag$ipw_weight <- ifelse(df_dag$treated == 1, 1,
                            df_dag$ps / (1 - df_dag$ps))

ess_control <- sum(df_dag$ipw_weight[df_dag$treated == 0])^2 /
               sum(df_dag$ipw_weight[df_dag$treated == 0]^2)
cat(sprintf("IPW effective sample size (controls): %.0f of %d\n",
            ess_control, sum(df_dag$treated == 0)))

m_ipw <- lm(log_avg_monthly_commits ~ treated + repo_age_yr +
               queried_language + is_org +
               log_pre_stars + log_pre_issues + log_pre_forks +
               log_pre_prs + log_pre_releases + log_pre_contribs +
               log_pre_commits,
             data = df_dag, weights = ipw_weight)

ipw_ct <- coeftest(m_ipw, vcov = vcovHC(m_ipw, type = "HC1"))
idx <- which(rownames(ipw_ct) == "treated")

ipw_result <- data.frame(
  method = "IPW (ATT)",
  coef   = ipw_ct[idx, 1],
  se     = ipw_ct[idx, 2],
  p_val  = ipw_ct[idx, 4],
  n      = nrow(df_dag)
)
```

## 3.3 Three-Estimator Comparison (Table 3)

```{r three_estimator_comparison}
all_results <- bind_rows(
  ols_results %>%
    filter(model == "DAG-justified (panel pre-treatment)") %>%
    transmute(method = "OLS (DAG-justified)", coef, se, p_val, n),
  psm_result,
  ipw_result
) %>%
  mutate(method = factor(method, levels = c("OLS (DAG-justified)",
                                            "PSM (1:1 NN)", "IPW (ATT)")))

all_results %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(p_val < 0.001, "<0.001", sprintf("%.3f", p_val))
  ) %>%
  select(Method = method, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, N = n) %>%
  kable(caption = "Table 3: Three Estimators, Same DAG-Justified Pre-Treatment Covariates") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The three estimators target slightly different estimands --- OLS
approximates the ATE while PSM and IPW estimate the ATT --- and operate
on different effective samples (PSM discards unmatched controls; IPW
reweights them).  These differences account for the modest variation in
point estimates, but all three coefficients are positive, significant,
and their confidence intervals overlap.  The substantive conclusion is
the same: switching from regression to matching to weighting changes
nothing because all three share the same identifying assumption ---
conditional ignorability given pre-treatment covariates --- and that
assumption is violated by the unobserved confounders that no GitHub API
data can capture.  **Using PSM or IPW does not automatically make a study
"causal inference."**
