---
title: "Cross-Sectional Diagnostic: AI Coding Tool Adoption and Development Productivity"
output: html_document
---

This notebook implements a cross-sectional diagnostic analysis of AI coding
tool adoption across top GitHub repositories, following the four-stage
structure from the Bogner & Merkel reanalysis.

The causal question is: "Does adopting AI coding tools increase development
productivity (measured by monthly commits)?"  The diagnostic arc shows that
cross-sectional data cannot credibly answer this question --- even though the
raw association is enormous and survives conventional controls:

- **Stage 1** documents severe covariate imbalance between repos with AI
  tool configuration files (L2+) and those without (L1).  L2+ repos average
  5.8x more monthly commits, but they are also systematically larger
  (SMD = 0.75), have far more PRs (0.91), more total commits (0.74), and
  are more likely to have CI/CD (0.54) --- all observable proxies for the
  unobservable confounders (developer skill, team culture) that drive both
  adoption and productivity.  Critically, the DAG review reveals that
  several covariates (CI, repo size, releases) have ambiguous causal
  status: they may be confounders *or* mediators, and the cross-sectional
  design cannot distinguish between the two.
- **Stage 2** compares three nested OLS specifications: M_pretreat (only
  the three unambiguously pre-treatment confounders), M_extended (adds
  covariates with ambiguous mediator risk), and M_all (kitchen-sink).
  The coefficient attenuates monotonically, but the drop from M_pretreat to
  M_extended is uninterpretable (confounding removal or over-conditioning on
  mediators?).  Oster bounds and sensemakr robustness values confirm the
  estimate is fragile under plausible unobserved confounding.
- **Stage 3** shows that PSM and IPW produce nearly identical estimates to
  OLS --- the estimator is not the bottleneck.
- **Stage 4** synthesises the cross-sectional ceiling argument: the
  coefficient's apparent robustness to controls is *exactly* what we would
  expect under selection bias (productive teams adopt AI tools), compounded
  by the inability to distinguish confounders from mediators in the
  cross-section.  Only longitudinal data --- comparing the same repo before
  and after adoption --- can make progress, which is what Example C
  (Cursor ITS/DiD) demonstrates.

# Setup

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "knitr", "kableExtra",
                   "lmtest", "sandwich", "sensemakr", "MatchIt", "cobalt",
                   "showtext", "scales")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(lmtest)
library(sandwich)
library(sensemakr)
library(MatchIt)
library(cobalt)
library(showtext)
library(scales)

font_add("Linux Libertine", regular = "LinLibertine_Rah.ttf")
showtext_auto()
```

# Load and Prepare Data

```{r load_data, message=FALSE, warning=FALSE}
repos <- read.csv("../data/ai_adoption_repos.csv", stringsAsFactors = FALSE)
monthly <- read.csv("../data/ai_adoption_monthly_commits.csv",
                     stringsAsFactors = FALSE)

cat(sprintf("Repos: %d\n", nrow(repos)))
cat(sprintf("Monthly commit rows: %d\n", nrow(monthly)))
```

```{r prepare_data}
df <- repos %>%
  mutate(
    # Treatment: L2+ vs L1
    treated     = as.integer(ai_maturity_level >= 2),
    # Derive repo age in years from creation date
    created_at  = as.POSIXct(created_at, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    pushed_at   = as.POSIXct(pushed_at,  format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    repo_age_yr = as.numeric(difftime(Sys.time(), created_at, units = "days")) / 365.25,
    # Log-transform skewed covariates (add 1 to handle zeros)
    log_stars        = log1p(stars),
    log_forks        = log1p(forks),
    log_size         = log1p(size_kb),
    log_total_commits = log1p(total_commits),
    log_total_prs    = log1p(total_prs),
    log_total_issues = log1p(total_issues),
    log_releases     = log1p(total_releases),
    log_open_issues  = log1p(open_issues),
    # Binary covariates
    is_org       = as.integer(owner_type == "Organization"),
    has_ci       = as.integer(has_ci == "True" | has_ci == TRUE),
    has_wiki     = as.integer(has_wiki == "True" | has_wiki == TRUE),
    has_discuss   = as.integer(has_discussions == "True" | has_discussions == TRUE)
  )

# Compute average monthly commits from the 6-month window columns
commit_cols <- grep("^commits_", names(df), value = TRUE)
df$avg_monthly_commits <- rowMeans(df[, commit_cols], na.rm = TRUE)
df$log_avg_monthly_commits <- log1p(df$avg_monthly_commits)

# Total commits in the 6-month window
df$total_window_commits <- rowSums(df[, commit_cols], na.rm = TRUE)
df$log_total_window_commits <- log1p(df$total_window_commits)

cat(sprintf("Treatment (L2+): %d  |  Control (L1): %d\n",
            sum(df$treated == 1), sum(df$treated == 0)))
cat(sprintf("Treatment prevalence: %.1f%%\n", mean(df$treated) * 100))
```

# Stage 1: Descriptive Statistics and Covariate Balance

## 1.1 AI Maturity Distribution

```{r maturity_distribution}
# By level
table(df$ai_maturity_level) %>%
  as.data.frame() %>%
  setNames(c("AI Maturity Level", "N")) %>%
  kable(caption = "AI Maturity Level Distribution") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r maturity_by_language}
# By language x treatment
crosstab <- df %>%
  group_by(queried_language, treated) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = treated, values_from = n, values_fill = 0) %>%
  rename(Control = `0`, Treated = `1`) %>%
  mutate(Total = Control + Treated,
         `% Treated` = sprintf("%.0f%%", Treated / Total * 100)) %>%
  arrange(desc(Treated))

crosstab %>%
  kable(caption = "Treatment (L2+) vs Control (L1) by Language") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 1.2 Outcome: Monthly Commits

```{r outcome_comparison, fig.width=8, fig.height=3}
group_labels <- c("0" = "L1 (no AI config)", "1" = "L2+ (AI config)")

df %>%
  mutate(group = factor(treated, labels = group_labels)) %>%
  ggplot(aes(x = log_avg_monthly_commits, fill = group)) +
  geom_density(alpha = 0.5) +
  labs(x = "Log(1 + Avg Monthly Commits)", y = "Density",
       fill = NULL,
       title = "Distribution of Monthly Commits by AI Adoption Status") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(legend.position = "bottom",
        legend.margin = margin(t = -5, r = 0, b = -3, l = 0),
        legend.box.margin = margin(t = -5, r = 0, b = -3, l = 0),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

```{r outcome_summary}
df %>%
  mutate(group = ifelse(treated == 1, "L2+ (treated)", "L1 (control)")) %>%
  group_by(group) %>%
  summarise(
    N = n(),
    `Mean monthly commits` = sprintf("%.1f", mean(avg_monthly_commits)),
    `Median monthly commits` = sprintf("%.1f", median(avg_monthly_commits)),
    `Mean log monthly commits` = sprintf("%.2f", mean(log_avg_monthly_commits)),
    .groups = "drop"
  ) %>%
  kable(caption = "Monthly Commits by Treatment Group") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 1.3 Covariate Balance (Standardised Mean Differences)

```{r covariate_balance, fig.width=8, fig.height=5}
# Compute SMD: (mean_treated - mean_control) / pooled_sd
balance_vars <- c("repo_age_yr", "log_stars", "log_forks", "log_size",
                  "log_total_commits", "log_total_prs", "log_total_issues",
                  "log_releases", "log_open_issues", "is_org", "has_ci",
                  "has_wiki", "has_discuss")

balance_labels <- c(
  "repo_age_yr"       = "Repo age (years)",
  "log_stars"         = "log(Stars)",
  "log_forks"         = "log(Forks)",
  "log_size"          = "log(Size KB)",
  "log_total_commits" = "log(Total commits)",
  "log_total_prs"     = "log(Total PRs)",
  "log_total_issues"  = "log(Total issues)",
  "log_releases"      = "log(Releases)",
  "log_open_issues"   = "log(Open issues)",
  "is_org"            = "Organization owner",
  "has_ci"            = "Has CI/CD",
  "has_wiki"          = "Has wiki",
  "has_discuss"       = "Has discussions"
)

compute_smd <- function(data, var, treat_col = "treated") {
  t_vals <- data[[var]][data[[treat_col]] == 1]
  c_vals <- data[[var]][data[[treat_col]] == 0]
  pooled_sd <- sqrt((var(t_vals, na.rm = TRUE) + var(c_vals, na.rm = TRUE)) / 2)
  if (pooled_sd < 1e-10) return(0)
  (mean(t_vals, na.rm = TRUE) - mean(c_vals, na.rm = TRUE)) / pooled_sd
}

smd_df <- data.frame(
  variable = balance_vars,
  label = balance_labels[balance_vars],
  smd = sapply(balance_vars, function(v) compute_smd(df, v))
) %>%
  mutate(label = reorder(label, abs(smd)))

ggplot(smd_df, aes(x = smd, y = label)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dotted", color = "red", alpha = 0.5) +
  labs(x = "Standardised Mean Difference (L2+ vs L1)",
       y = NULL,
       title = "Covariate Balance Before Adjustment",
       caption = "Dotted red lines mark ±0.1 SMD threshold") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```

```{r smd_table}
smd_df %>%
  arrange(desc(abs(smd))) %>%
  mutate(smd = sprintf("%.3f", smd)) %>%
  select(Covariate = label, SMD = smd) %>%
  kable(caption = "Standardised Mean Differences (L2+ minus L1)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 1.4 Interpretation

The balance diagnostics confirm the selection story.  Every covariate
exceeds the conventional ±0.1 SMD threshold, with total PRs (SMD = 0.91),
repo size (0.75), total commits (0.74), releases (0.61), and CI presence
(0.54) showing the most severe imbalances.

These are not random differences --- they reflect a coherent selection
mechanism: well-resourced, actively maintained, organization-backed
projects are the ones that adopt AI tools.  These same characteristics
independently predict high commit velocity.  Language confounding amplifies
the problem: TypeScript (53% treated) and Go (31%) have strong AI tooling
ecosystems, while C (9%) and Swift (13%) barely participate.

The proposed causal DAG encodes this structure --- though, crucially, the
cross-sectional design introduces ambiguity in classifying several
covariates:

- **Clean confounders** (unambiguously pre-treatment common causes of
  adoption and productivity): repo age, primary language, and organization
  type.  These are structural features determined at repo creation that
  plausibly affect both AI tool adoption and commit velocity.
- **Ambiguous covariates** (confounder proxy, or mediator?): CI/CD
  presence, repo size, and number of releases.  Each has a plausible
  confounder story --- e.g., CI presence proxies for engineering culture,
  which drives both adoption and productivity --- *and* a plausible mediator
  story --- e.g., AI adoption may be part of a modernization wave that
  includes CI; current repo size partly reflects post-treatment code growth;
  total releases includes releases made after adoption.  Since all variables
  are measured at the same snapshot, we cannot separate pre-treatment from
  post-treatment variation.  Conditioning on mediators blocks causal paths
  and biases the estimate *toward* zero, creating a false impression of
  confounding removal.
- **Likely confounder proxies** (misclassifiable as colliders): stars and
  forks are primarily driven by project popularity --- a latent variable
  that also drives both adoption (popular repos attract AI-savvy
  contributors) and productivity (popular repos have larger, more active
  contributor bases).  A collider story exists (AI adoption → hype → stars;
  high activity → visibility → stars) but is weaker than the
  confounder-proxy story.  Moreover, the sample is already conditioned on
  stars (top repos by stars per language), so any collider-induced bias is
  baked in regardless of the regression specification.
- **Unobserved**: developer skill, team culture, organizational investment
  in tooling.  These are the confounders that matter most and that no
  amount of GitHub API data can capture.

This ambiguity --- where we cannot even confidently classify observed
variables as confounders vs. mediators --- is itself evidence of the
cross-sectional design's limitations.  Stage 2 formalises this by
comparing a model with only the three clean confounders against one that
adds the ambiguous covariates: the coefficient drop between them is
uninterpretable.

The 5.8x raw gap in monthly commits (15x at the median) is strikingly
large but causally uninterpretable.  The question is how much survives
after controlling for observable confounders --- and whether the residual
is credible.

# Stage 2: OLS and Omitted Variable Bias Diagnostics

## 2.1 Three OLS Specifications

The DAG review in Section 1.4 motivates three nested specifications that
make the confounder-vs-mediator ambiguity explicit:

- **M_pretreat**: only the three unambiguously pre-treatment confounders
  (repo age, language, org type).
- **M_extended**: adds covariates whose causal status is ambiguous (CI/CD
  presence, repo size, total releases).  These may proxy for unobserved
  engineering culture (useful) or may partly mediate the causal effect
  (harmful).
- **M_all**: further adds stars, forks, PRs, and issues --- variables that
  are plausibly confounder proxies (popularity) but also include
  post-treatment variation.

The coefficient drop from M_pretreat to M_extended is *uninterpretable*:
it could reflect confounding removal (good) or causal-path blocking (bad).
This ambiguity is the heart of the cross-sectional ceiling.

```{r ols_models}
# M_pretreat: only unambiguously pre-treatment confounders
m_pretreat <- lm(log_avg_monthly_commits ~ treated + repo_age_yr +
                   queried_language + is_org,
                 data = df)

# M_extended: adds covariates with ambiguous causal status (proxy or mediator?)
m_ext <- lm(log_avg_monthly_commits ~ treated + repo_age_yr + queried_language +
              is_org + has_ci + log_size + log_releases,
            data = df)

# M_all: adds stars, forks, PRs, issues (likely confounder proxies but also
# include post-treatment variation)
m_all <- lm(log_avg_monthly_commits ~ treated + repo_age_yr + queried_language +
              is_org + has_ci + log_size + log_releases +
              log_stars + log_forks + log_total_prs + log_total_issues,
            data = df)

# Extract treatment coefficients with HC1 robust SEs
extract_treated <- function(model, label) {
  ct <- coeftest(model, vcov = vcovHC(model, type = "HC1"))
  idx <- which(rownames(ct) == "treated")
  data.frame(
    model    = label,
    coef     = ct[idx, "Estimate"],
    se       = ct[idx, "Std. Error"],
    t_val    = ct[idx, "t value"],
    p_val    = ct[idx, "Pr(>|t|)"],
    r2       = summary(model)$r.squared,
    adj_r2   = summary(model)$adj.r.squared,
    n        = nobs(model)
  )
}

ols_results <- bind_rows(
  extract_treated(m_pretreat, "M_pretreat (clean confounders)"),
  extract_treated(m_ext,      "M_extended (+ ambiguous covariates)"),
  extract_treated(m_all,      "M_all (+ stars, forks, ...)")
)

ols_results %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(as.numeric(p_val) < 0.001, "<0.001", sprintf("%.3f", p_val)),
    r2    = sprintf("%.3f", r2)
  ) %>%
  select(Model = model, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, `R²` = r2, N = n) %>%
  kable(caption = "OLS Estimates of AI Adoption Effect on log(Monthly Commits)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The three clean confounders (age, language, org type) already absorb a
substantial share of the raw gap.  Adding the ambiguous covariates (CI,
size, releases) attenuates the coefficient further --- but this additional
drop is uninterpretable: if CI or size partly mediate the causal effect,
the attenuation reflects over-conditioning, not confounding removal.
M_all pushes the coefficient down still further, yet all three controlled
estimates remain highly significant (p < 0.001).

A naive analyst might take this persistence as evidence that the effect is
"robust to controls."  The OVB diagnostics below show why that conclusion
is premature.

## 2.2 Coefficient Stability Across Specifications

```{r coefficient_stability, fig.width=8, fig.height=3}
# Build a sequence of progressively richer models
formulas <- list(
  "Bivariate"                          = log_avg_monthly_commits ~ treated,
  "+ Age + Language"                    = log_avg_monthly_commits ~ treated +
                                           repo_age_yr + queried_language,
  "M_pretreat (+ Org)"                 = log_avg_monthly_commits ~ treated +
                                           repo_age_yr + queried_language + is_org,
  "M_extended (+ CI, Size, Releases)"  = log_avg_monthly_commits ~ treated +
                                           repo_age_yr + queried_language + is_org +
                                           has_ci + log_size + log_releases,
  "M_all (+ Stars, Forks, ...)"        = log_avg_monthly_commits ~ treated +
                                           repo_age_yr + queried_language + is_org +
                                           has_ci + log_size + log_releases +
                                           log_stars + log_forks + log_total_prs +
                                           log_total_issues
)

stability <- lapply(names(formulas), function(label) {
  mod <- lm(formulas[[label]], data = df)
  ct <- coeftest(mod, vcov = vcovHC(mod, type = "HC1"))
  idx <- which(rownames(ct) == "treated")
  data.frame(spec = label, coef = ct[idx, 1], se = ct[idx, 2],
             stringsAsFactors = FALSE)
}) %>%
  bind_rows() %>%
  mutate(spec = factor(spec, levels = names(formulas)))

# Highlight M_pretreat (defensible) in blue, M_extended (ambiguous) in orange
pre_idx <- which(levels(stability$spec) == "M_pretreat (+ Org)")
ext_idx <- which(levels(stability$spec) == "M_extended (+ CI, Size, Releases)")

ggplot(stability, aes(x = spec, y = coef)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  annotate("rect",
           xmin = pre_idx - 0.4, xmax = pre_idx + 0.4,
           ymin = -Inf, ymax = Inf,
           fill = "steelblue", alpha = 0.12) +
  annotate("rect",
           xmin = ext_idx - 0.4, xmax = ext_idx + 0.4,
           ymin = -Inf, ymax = Inf,
           fill = "darkorange", alpha = 0.10) +
  geom_pointrange(aes(ymin = coef - 1.96 * se, ymax = coef + 1.96 * se), size = 0.5) +
  labs(x = NULL, y = "Coefficient on AI Adoption (treated)",
       title = "Coefficient Stability Across Specifications",
       caption = paste("Blue: clean pre-treatment confounders.",
                       "Orange: adds ambiguous covariates (mediator risk).",
                       "95% CI with HC1 SEs.")) +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

The coefficient stability plot shows steady attenuation as controls are
added.  The trajectory is monotonically downward, and even the kitchen-sink
estimate remains far from zero.

The blue-shaded M_pretreat specification contains only unambiguously
pre-treatment confounders; the coefficient drop from the bivariate to
M_pretreat is defensibly attributed to confounding removal.  The
orange-shaded M_extended step adds CI, repo size, and releases ---
covariates whose causal status is ambiguous.  The further attenuation
*could* be additional confounding removal (if these variables proxy for
engineering culture) or over-conditioning on mediators (if AI adoption
partly *causes* CI presence, code growth, or release frequency).  We
cannot tell which from cross-sectional data.

This *looks* robust --- but the attenuation pattern itself is the warning
sign.  Each batch of controls explains away more of the association.  The
question is whether the unobserved confounders (developer skill, team
culture) --- which are plausibly *more* strongly associated with both
treatment and outcome than any individual observed covariate --- would
continue the downward trajectory to zero.

## 2.3 Cinelli & Hazlett Sensitivity Analysis (sensemakr)

```{r sensemakr_pretreat}
# sensemakr on M_pretreat (clean confounders only --- most defensible model)
sense_pre <- sensemakr(m_pretreat, treatment = "treated",
                       benchmark_covariates = c("repo_age_yr", "is_org"),
                       kd = 1:3)

cat("=== M_pretreat: Sensitivity analysis ===\n")
summary(sense_pre)
```

```{r sensemakr_plot, fig.width=8, fig.height=5}
plot(sense_pre, sensitivity.of = "estimate")
```

```{r sensemakr_extended}
# sensemakr on M_extended (adds ambiguous covariates)
sense_ext <- sensemakr(m_ext, treatment = "treated",
                       benchmark_covariates = c("log_size", "repo_age_yr"),
                       kd = 1:3)

cat("=== M_extended: Sensitivity analysis ===\n")
summary(sense_ext)
```

We run the sensitivity analysis on both M_pretreat (the defensible
specification) and M_extended (the specification that includes ambiguous
covariates).

The RV for M_pretreat tells us how strong a single unobserved confounder
would need to be --- in terms of partial R² with both treatment and
outcome --- to nullify the estimate at the 5% level.  M_pretreat has the
higher RV because its coefficient is larger (fewer controls → less
attenuation).  The RV for M_extended is lower because the ambiguous
covariates have already absorbed some of the association --- but whether
they absorbed confounding or the causal effect itself is unclear.

Consider "developer skill" as a confounder.  Skilled developers are more
likely to adopt AI tools (they are early adopters, they configure their
tooling carefully) *and* more likely to commit frequently (they are more
productive).  A confounder this strong is not only plausible --- it is
expected.  Even the higher M_pretreat RV is within the range that such
an unobservable could plausibly achieve.

## 2.4 Oster Proportional Selection Bounds

```{r oster_bounds}
# Oster (2019): δ* = how much selection on unobservables (relative to
# observables) is needed to drive the treatment effect to zero.
# δ* = (β_full * (R_max - R_full)) / ((β_restricted - β_full) * (R_full - R_restricted))
# We use R_max = min(1, 1.3 * R_full) following Oster's recommendation.

oster_delta <- function(model_restricted, model_full, r_max_multiplier = 1.3) {
  b_r <- coef(model_restricted)["treated"]
  b_f <- coef(model_full)["treated"]
  r_r <- summary(model_restricted)$r.squared
  r_f <- summary(model_full)$r.squared
  r_max <- min(1, r_max_multiplier * r_f)

  delta <- (b_f * (r_max - r_f)) / ((b_r - b_f) * (r_f - r_r))
  return(data.frame(
    beta_restricted = b_r,
    beta_full       = b_f,
    R2_restricted   = r_r,
    R2_full         = r_f,
    R2_max          = r_max,
    delta_star      = delta
  ))
}

m_bivariate <- lm(log_avg_monthly_commits ~ treated, data = df)

oster_pre <- oster_delta(m_bivariate, m_pretreat)
oster_ext <- oster_delta(m_bivariate, m_ext)
oster_ambig <- oster_delta(m_pretreat, m_ext)

bind_rows(
  oster_pre   %>% mutate(comparison = "Bivariate → M_pretreat"),
  oster_ext   %>% mutate(comparison = "Bivariate → M_extended"),
  oster_ambig %>% mutate(comparison = "M_pretreat → M_extended")
) %>%
  mutate(across(where(is.numeric), ~sprintf("%.3f", .))) %>%
  select(Comparison = comparison, `β restricted` = beta_restricted,
         `β full` = beta_full, `R² restricted` = R2_restricted,
         `R² full` = R2_full, `R² max` = R2_max, `δ*` = delta_star) %>%
  kable(caption = "Oster Proportional Selection Bounds") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The three comparisons tell different stories:

**Bivariate → M_pretreat** has a large δ\*, meaning that --- relative to
the clean confounders (age, language, org type) --- unobservables would
need to be *many times* stronger to explain the estimate away.  At face
value this looks robust.  But the clean confounders absorb only a modest
share of the raw gap (the R² increase is small), so the "ruler" against
which δ\* is measured is short.  The large δ\* does not mean the estimate
is safe; it means the clean confounders are simply not very informative
about the selection process.

**Bivariate → M_extended** has a much smaller δ\* (around 0.56).  This
comparison is the one a standard analysis would report, but its
interpretation is muddied: some of the R² gain from adding CI, size, and
releases may come from blocking causal paths (mediator contamination)
rather than removing confounding.

**M_pretreat → M_extended** isolates the contribution of the ambiguous
covariates.  Its δ\* tells us how much the unobservables would need to
resemble the ambiguous covariates to drive the coefficient from the
M_pretreat estimate to zero.  But this comparison is doubly hard to
interpret: the "observables" in the denominator are themselves
ambiguous --- they may be proxies (which would make δ\* meaningful) or
mediators (which would make the comparison invalid).

Regardless of which comparison we emphasise, the most important
confounders (developer skill, team culture, organizational resources) are
*precisely* the ones we cannot observe from GitHub API data.  The Oster
framework cannot resolve the confounder-vs-mediator ambiguity for us ---
it can only tell us how far we are from the knife-edge, conditional on
the covariates actually *being* confounders.

# Stage 3: Propensity Score Matching and IPW

Stage 3 asks whether alternative selection-on-observables methods
(matching, IPW) yield different answers from OLS.  If the problem were
model misspecification (nonlinearity, poor overlap), matching or IPW might
help.  If the problem is unobserved confounding, all three estimators will
converge on the same biased answer --- because they share the same
identifying assumption (conditional ignorability given observables).

## 3.1 Propensity Score Estimation

```{r psm_estimation}
# Estimate propensity score using M_extended covariates (includes ambiguous
# covariates --- the convergence argument holds regardless of which set we use)
ps_formula <- treated ~ repo_age_yr + queried_language + is_org + has_ci +
                log_size + log_releases

m_ps <- matchit(ps_formula, data = df, method = "nearest",
                distance = "glm", ratio = 1, replace = FALSE)

summary(m_ps)
```

## 3.2 Balance After Matching

```{r psm_balance, fig.width=8, fig.height=5}
love.plot(m_ps, binary = "std", thresholds = c(m = 0.1),
          var.names = balance_labels,
          colors = c("red", "steelblue"),
          shapes = c(17, 16),
          title = "Covariate Balance: Before vs After Matching") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine")
```

## 3.3 Matched Sample OLS

```{r psm_ols}
matched_data <- match.data(m_ps)

m_psm <- lm(log_avg_monthly_commits ~ treated + repo_age_yr + queried_language +
               is_org + has_ci + log_size + log_releases,
             data = matched_data, weights = weights)

psm_ct <- coeftest(m_psm, vcov = vcovHC(m_psm, type = "HC1"))
idx <- which(rownames(psm_ct) == "treated")

psm_result <- data.frame(
  method = "PSM (1:1 NN)",
  coef   = psm_ct[idx, 1],
  se     = psm_ct[idx, 2],
  p_val  = psm_ct[idx, 4],
  n      = nrow(matched_data)
)
```

## 3.4 Inverse Probability Weighting

```{r ipw}
# Estimate propensity scores
df$ps <- predict(glm(ps_formula, data = df, family = binomial), type = "response")

# IPW weights for ATT: treated get weight 1, controls get PS/(1-PS)
df$ipw_weight <- ifelse(df$treated == 1, 1, df$ps / (1 - df$ps))

# Effective sample size for controls
ess_control <- sum(df$ipw_weight[df$treated == 0])^2 /
               sum(df$ipw_weight[df$treated == 0]^2)
cat(sprintf("IPW effective sample size (controls): %.0f of %d\n",
            ess_control, sum(df$treated == 0)))

m_ipw <- lm(log_avg_monthly_commits ~ treated + repo_age_yr + queried_language +
               is_org + has_ci + log_size + log_releases,
             data = df, weights = ipw_weight)

ipw_ct <- coeftest(m_ipw, vcov = vcovHC(m_ipw, type = "HC1"))
idx <- which(rownames(ipw_ct) == "treated")

ipw_result <- data.frame(
  method = "IPW (ATT)",
  coef   = ipw_ct[idx, 1],
  se     = ipw_ct[idx, 2],
  p_val  = ipw_ct[idx, 4],
  n      = nrow(df)
)
```

## 3.5 Comparison: OLS, PSM, and IPW

```{r three_estimator_comparison}
all_results <- bind_rows(
  ols_results %>%
    filter(model == "M_extended (+ ambiguous covariates)") %>%
    transmute(method = "OLS (M_extended)", coef, se, p_val, n),
  psm_result,
  ipw_result
) %>%
  mutate(method = factor(method, levels = c("OLS (M_extended)", "PSM (1:1 NN)", "IPW (ATT)")))

all_results %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(p_val < 0.001, "<0.001", sprintf("%.3f", p_val))
  ) %>%
  select(Method = method, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, N = n) %>%
  kable(caption = "Three Estimators, Same M_extended Covariates") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r three_estimator_plot, fig.width=8, fig.height=3}
ggplot(all_results, aes(x = method, y = coef)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = coef - 1.96 * se, ymax = coef + 1.96 * se,
                      color = method),
                  size = 0.8) +
  scale_color_manual(values = c("OLS (M_extended)" = "#4E79A7",
                                "PSM (1:1 NN)"     = "#59A14F",
                                "IPW (ATT)"        = "#F28E2B")) +
  labs(x = NULL, y = "Coefficient on AI Adoption",
       title = "Three Estimators Converge Under the Same Covariates",
       caption = "All use M_extended covariates. 95% CI with HC1 robust SEs.") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

# Stage 4: The Cross-Sectional Ceiling

## The diagnostic summary

The four stages build a cumulative case:

1. **The raw gap is enormous** (5.8x in means, 15x at the median), but
   covariate balance shows that L2+ repos differ from L1 on *every*
   observable dimension.  The treated group is a systematically different
   population, not a random subset.

2. **Even clean confounders absorb a large share of the gap**, and the
   coefficient *keeps falling* as more controls are added.  The downward
   trajectory has not flattened.  The additional drop from M_pretreat to
   M_extended is uninterpretable: it could reflect confounding removal
   (if CI, size, and releases proxy for engineering culture) or
   over-conditioning on mediators (if AI adoption partly *causes* those
   variables).

3. **OVB diagnostics say the residual is fragile**: Oster δ* values
   indicate that unobservables need not be dramatically more important
   than observables to explain the entire residual association.

4. **The estimator is not the bottleneck**: OLS, PSM, and IPW converge
   tightly.  Switching from regression to matching to weighting changes
   nothing because all three share the same identifying assumption ---
   and that assumption is violated.

## Why the coefficient survives controls

It is tempting to interpret the persistently significant coefficient as
evidence of a genuine causal effect.  But the persistence is *exactly*
what selection bias predicts:

- **Developer skill** (unobserved) is plausibly the strongest confounder.
  Skilled developers both commit more code *and* adopt AI tools earlier.
  No GitHub API covariate captures this.
- **Organizational investment** (unobserved) drives both tooling adoption
  and development velocity.  Being organization-owned (which we observe)
  is a crude proxy; the actual confounders are team size, budget, and
  engineering culture.
- **Reverse causality** compounds the problem: high-velocity projects
  have more reason to invest in AI tooling, so the causal arrow may
  run from productivity to adoption rather than the reverse.

The observables we control for are *proxies* for the true confounders, not
the confounders themselves --- and some (CI, size, releases) may partly be
mediators rather than confounders, making the coefficient drop from
M_pretreat to M_extended unreliable as evidence of confounding removal.
The proxies absorb some but not all of the selection, which is why
the coefficient attenuates but does not vanish.

## The bridge to longitudinal data

Cross-sectional data cannot separate "AI tools increase productivity"
from "productive teams adopt AI tools."  The identifying variation must
come from *within-repo temporal changes* --- comparing the same project
before and after AI adoption, so that all time-invariant confounders
(developer skill, org culture, project domain) are absorbed by
construction.

This is precisely what Example C (Cursor ITS/DiD) provides.  There, the
naive analysis (without repo fixed effects) produces qualitatively wrong
conclusions about code quality --- the sign *flips* when within-repo
variation is used.  The velocity finding (commits increase) *does* survive
the longitudinal design, but the cross-sectional analysis we performed
here cannot know that in advance.  The lesson is not that the causal effect
is zero, but that the cross-sectional coefficient --- however large and
significant --- is uninterpretable.

# Summary Table

```{r summary_table}
# Pull key numbers from models for the summary table
b_biv  <- sprintf("%.2f", coef(m_bivariate)["treated"])
b_pre  <- sprintf("%.2f", coef(m_pretreat)["treated"])
b_ext  <- sprintf("%.2f", coef(m_ext)["treated"])
drop_pct <- sprintf("%.0f", (1 - coef(m_pretreat)["treated"] /
                                  coef(m_bivariate)["treated"]) * 100)
d_pre  <- sprintf("%.2f", oster_pre$delta_star)

summary_stages <- data.frame(
  Stage = c("Stage 1: Covariate balance",
            "Stage 2: OLS + OVB diagnostics",
            "Stage 3: PSM and IPW",
            "Stage 4: Cross-sectional ceiling"),
  Finding = c(
    paste0("L2+ repos average 5.8x more monthly commits (15x at the median), ",
           "but differ on every observable (total PRs SMD = 0.91, size = 0.75, ",
           "commits = 0.74, CI = 0.54)"),
    paste0("Clean confounders cut the coefficient by ", drop_pct, "% (",
           b_biv, " → ", b_pre, "); adding ambiguous covariates drops it ",
           "further (→ ", b_ext, ") but this additional drop is uninterpretable ",
           "(confounding removal or mediator blocking?). Oster δ* = ", d_pre,
           " (bivariate → M_pretreat)"),
    paste0("OLS, PSM, and IPW converge tightly under the same covariates — ",
           "the estimator is not the bottleneck; the identifying assumption is"),
    paste0("The persistent coefficient is consistent with selection bias (skilled ",
           "teams both adopt AI tools and commit more); the confounder-vs-mediator ",
           "ambiguity compounds the problem; only within-repo longitudinal ",
           "variation can make progress (→ Example C)")
  )
)

summary_stages %>%
  kable(caption = "Diagnostic Arc Summary") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```
