---
title: "Cross-Sectional Analysis: AI Coding Tool Adoption and Development Productivity"
output: html_document
---

This notebook implements the cross-sectional analysis of AI coding tool
adoption for Section 4.1 of the paper.  The analysis follows a three-stage
arc that starts from what a typical SE researcher would do, diagnoses why
it fails, and shows that switching estimators cannot help:

- **Stage 1** starts from the typical approach: a naive comparison and a
  kitchen-sink regression.  The raw gap is enormous and the kitchen-sink
  coefficient is large and highly significant.  A naive analyst might stop
  here.
- **Stage 2** diagnoses why we should distrust the kitchen-sink result.
  The key insight is **temporal collapse**: MSR snapshot data merges
  pre-treatment and post-treatment states of time-varying covariates
  (stars, PRs, issues, CI, size, releases) into a single node, destroying
  the acyclicity that DAGs require and making covariate selection
  incoherent.  Only *structurally pre-treatment* covariates (repo age,
  language, org type) are defensible controls.  The pre-treatment-only
  estimate is large but fragile under sensitivity analysis.
- **Stage 3** shows that OLS, PSM, and IPW converge on the same estimate
  under pre-treatment covariates --- the estimator is not the bottleneck;
  the identifying assumption is.

The takeaway: cross-sectional MSR data cannot credibly answer causal
questions about AI tool adoption.  Only within-repo longitudinal variation
can make progress (Section 4.2).

# Setup

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "knitr", "kableExtra",
                   "lmtest", "sandwich", "sensemakr", "MatchIt", "cobalt",
                   "showtext", "scales")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(lmtest)
library(sandwich)
library(sensemakr)
library(MatchIt)
library(cobalt)
library(showtext)
library(scales)

font_add("Linux Libertine", regular = "LinLibertine_Rah.ttf")
showtext_auto()
```

# Load and Prepare Data

```{r load_data, message=FALSE, warning=FALSE}
repos <- read.csv("../data/ai_adoption_repos.csv", stringsAsFactors = FALSE)
monthly <- read.csv("../data/ai_adoption_monthly_commits.csv",
                     stringsAsFactors = FALSE)

cat(sprintf("Repos: %d\n", nrow(repos)))
cat(sprintf("Monthly commit rows: %d\n", nrow(monthly)))
```

```{r prepare_data}
df <- repos %>%
  mutate(
    # Treatment: L2+ vs L1
    treated     = as.integer(ai_maturity_level >= 2),
    # Derive repo age in years from creation date
    created_at  = as.POSIXct(created_at, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    pushed_at   = as.POSIXct(pushed_at,  format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    repo_age_yr = as.numeric(difftime(Sys.time(), created_at, units = "days")) / 365.25,
    # Log-transform skewed covariates (add 1 to handle zeros)
    log_stars        = log1p(stars),
    log_forks        = log1p(forks),
    log_size         = log1p(size_kb),
    log_total_commits = log1p(total_commits),
    log_total_prs    = log1p(total_prs),
    log_total_issues = log1p(total_issues),
    log_releases     = log1p(total_releases),
    log_open_issues  = log1p(open_issues),
    # Binary covariates
    is_org       = as.integer(owner_type == "Organization"),
    has_ci       = as.integer(has_ci == "True" | has_ci == TRUE),
    has_wiki     = as.integer(has_wiki == "True" | has_wiki == TRUE),
    has_discuss   = as.integer(has_discussions == "True" | has_discussions == TRUE)
  )

# Compute average monthly commits from the 6-month window columns
commit_cols <- grep("^commits_", names(df), value = TRUE)
df$avg_monthly_commits <- rowMeans(df[, commit_cols], na.rm = TRUE)
df$log_avg_monthly_commits <- log1p(df$avg_monthly_commits)

# Total commits in the 6-month window
df$total_window_commits <- rowSums(df[, commit_cols], na.rm = TRUE)
df$log_total_window_commits <- log1p(df$total_window_commits)

cat(sprintf("Treatment (L2+): %d  |  Control (L1): %d\n",
            sum(df$treated == 1), sum(df$treated == 0)))
cat(sprintf("Treatment prevalence: %.1f%%\n", mean(df$treated) * 100))
```

# Stage 1: What Would a Typical SE Researcher Do?

A typical MSR study would start by comparing treated and control groups,
then run a regression with all available covariates.  We follow this
approach first, then diagnose why it fails.

## 1.1 Descriptive Comparison (Table 1)

Table 1 presents the outcome and all covariates side by side for the
control (L1: no AI config) and treated (L2+: AI config present) groups,
with standardised mean differences (SMDs).  Covariates are grouped by
their temporal status --- a distinction we will motivate in Stage 2.

```{r table1_descriptive}
# Helper: compute SMD
compute_smd <- function(data, var, treat_col = "treated") {
  t_vals <- data[[var]][data[[treat_col]] == 1]
  c_vals <- data[[var]][data[[treat_col]] == 0]
  pooled_sd <- sqrt((var(t_vals, na.rm = TRUE) + var(c_vals, na.rm = TRUE)) / 2)
  if (pooled_sd < 1e-10) return(0)
  (mean(t_vals, na.rm = TRUE) - mean(c_vals, na.rm = TRUE)) / pooled_sd
}

# Define variables, labels, and temporal status
table1_vars <- c(
  "avg_monthly_commits",
  "repo_age_yr", "is_org",
  "log_stars", "log_forks", "log_size",
  "log_total_commits", "log_total_prs", "log_total_issues",
  "log_releases", "log_open_issues",
  "has_ci", "has_wiki", "has_discuss"
)

table1_labels <- c(
  "avg_monthly_commits" = "Avg monthly commits",
  "repo_age_yr"         = "Repo age (years)",
  "is_org"              = "Organization owner",
  "log_stars"           = "log(Stars)",
  "log_forks"           = "log(Forks)",
  "log_size"            = "log(Size KB)",
  "log_total_commits"   = "log(Total commits)",
  "log_total_prs"       = "log(Total PRs)",
  "log_total_issues"    = "log(Total issues)",
  "log_releases"        = "log(Releases)",
  "log_open_issues"     = "log(Open issues)",
  "has_ci"              = "Has CI/CD",
  "has_wiki"            = "Has wiki",
  "has_discuss"         = "Has discussions"
)

table1_temporal <- c(
  "avg_monthly_commits" = "Outcome",
  "repo_age_yr"         = "Pre-treatment",
  "is_org"              = "Pre-treatment",
  "log_stars"           = "Time-varying",
  "log_forks"           = "Time-varying",
  "log_size"            = "Time-varying",
  "log_total_commits"   = "Time-varying",
  "log_total_prs"       = "Time-varying",
  "log_total_issues"    = "Time-varying",
  "log_releases"        = "Time-varying",
  "log_open_issues"     = "Time-varying",
  "has_ci"              = "Time-varying",
  "has_wiki"            = "Time-varying",
  "has_discuss"         = "Time-varying"
)

table1_df <- data.frame(
  variable = table1_vars,
  label    = table1_labels[table1_vars],
  temporal = table1_temporal[table1_vars],
  stringsAsFactors = FALSE
) %>%
  rowwise() %>%
  mutate(
    control_mean = mean(df[[variable]][df$treated == 0], na.rm = TRUE),
    treated_mean = mean(df[[variable]][df$treated == 1], na.rm = TRUE),
    smd          = compute_smd(df, variable)
  ) %>%
  ungroup()

table1_df %>%
  mutate(
    control_mean = sprintf("%.2f", control_mean),
    treated_mean = sprintf("%.2f", treated_mean),
    smd          = sprintf("%.3f", smd)
  ) %>%
  select(
    Variable = label,
    `Temporal Status` = temporal,
    `Control (L1)` = control_mean,
    `Treated (L2+)` = treated_mean,
    SMD = smd
  ) %>%
  kable(caption = "Table 1: Descriptive Comparison by AI Adoption Status") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The treated group (L2+) differs from the control group (L1) on *every*
dimension.  Every covariate exceeds the conventional +/-0.1 SMD threshold.
The raw gap in monthly commits is enormous.  These are not random
differences --- they reflect a coherent selection mechanism:
well-resourced, actively maintained, organization-backed projects are the
ones that adopt AI tools.

Note: the primary language is not shown in the table but is another
important pre-treatment covariate.  TypeScript (53% treated) and Go (31%)
have strong AI tooling ecosystems, while C (9%) and Swift (13%) barely
participate.

## 1.2 Kitchen-Sink Regression

A typical SE researcher might now run a regression controlling for
everything available.

```{r kitchen_sink}
# Kitchen-sink: all available covariates
m_kitchen <- lm(log_avg_monthly_commits ~ treated + repo_age_yr +
                  queried_language + is_org + has_ci + log_size +
                  log_releases + log_stars + log_forks + log_total_prs +
                  log_total_issues,
                data = df)

# Extract treatment coefficient with HC1 robust SEs
extract_treated <- function(model, label) {
  ct <- coeftest(model, vcov = vcovHC(model, type = "HC1"))
  idx <- which(rownames(ct) == "treated")
  data.frame(
    model    = label,
    coef     = ct[idx, "Estimate"],
    se       = ct[idx, "Std. Error"],
    t_val    = ct[idx, "t value"],
    p_val    = ct[idx, "Pr(>|t|)"],
    r2       = summary(model)$r.squared,
    adj_r2   = summary(model)$adj.r.squared,
    n        = nobs(model)
  )
}

kitchen_result <- extract_treated(m_kitchen, "Kitchen-sink (all covariates)")

kitchen_result %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(as.numeric(p_val) < 0.001, "<0.001", sprintf("%.3f", p_val)),
    r2    = sprintf("%.3f", r2)
  ) %>%
  select(Model = model, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, `R²` = r2, N = n) %>%
  kable(caption = "Kitchen-Sink OLS: AI Adoption Effect on log(Monthly Commits)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The coefficient is positive and highly significant (p < 0.001), surviving
a battery of controls.  A naive analyst might conclude that AI adoption
*causes* higher productivity and stop here.  But should we trust this?

# Stage 2: Why Should We Distrust This?

## 2.1 The Temporal Collapse Problem

To reason about which covariates to include in a regression, we need a
causal DAG.  But the cross-sectional design introduces a fundamental
problem that is pervasive in MSR studies.

Consider a variable like "stars."  In the true data-generating process,
stars accumulate over time.  The actual causal DAG has time-indexed nodes:

- **stars\_{t-1}** (before AI adoption) is a **confounder**: popular repos
  attract AI-savvy contributors (stars\_{t-1} -> adoption\_t) *and* have
  more active contributor bases (stars\_{t-1} -> commits\_t).
- **stars\_{t+1}** (after AI adoption) is a **collider**: both AI adoption
  and high commit activity increase visibility and attract stars
  (adoption\_t -> stars\_{t+1} <- commits\_t).

In this true temporal DAG, the adjustment set is unambiguous: condition on
stars\_{t-1}, never on stars\_{t+1}.

But a typical MSR study queries the GitHub API once at the time of the
study.  This **collapses** both time-indexed nodes into a single "stars"
variable that contains both pre-treatment and post-treatment variation.
The resulting graph has both incoming and outgoing arrows between "stars"
and both treatment and outcome --- it is **cyclic**, not a DAG.  No
adjustment formula applies.

This temporal collapse is not specific to stars.  Every time-varying
covariate in our dataset (PRs, issues, CI presence, repo size, releases)
suffers from the same problem.  The true temporal DAG is well-defined;
the cross-sectional snapshot destroys it.

(See the paper's Figures X and Y for the true temporal DAG and the
collapsed cross-sectional graph.)

This connects to a recognized problem in the causal inference literature:
Hernan & Robins (2020, Ch. 20) describe this as "treatment-confounder
feedback," and Richardson & Robins have critiqued the use of
cross-sectional DAGs precisely because the nodes must be well-defined
temporal events, not time-collapsed summaries.

## 2.2 Covariate Classification

The temporal collapse argument divides our covariates into two groups:

```{r covariate_classification}
classif <- data.frame(
  Covariate = c("Repo age", "Primary language", "Organization owner",
                "Stars", "Forks", "Repo size (KB)", "Total commits",
                "Total PRs", "Total issues", "Releases", "Open issues",
                "CI/CD presence", "Wiki", "Discussions"),
  `Temporal Status` = c(rep("Pre-treatment", 3), rep("Temporally ambiguous", 11)),
  Justification = c(
    "Determined at creation; cannot be affected by AI adoption",
    "Determined at creation; cannot be affected by AI adoption",
    "Determined at creation; cannot be affected by AI adoption",
    "Accumulates before and after adoption",
    "Accumulates before and after adoption",
    "Includes post-adoption code growth",
    "Includes post-adoption commits",
    "Includes post-adoption PRs",
    "Includes post-adoption issues",
    "Includes post-adoption releases",
    "Snapshot mixes pre- and post-adoption state",
    "May be adopted as part of same modernization wave",
    "Snapshot mixes pre- and post-adoption state",
    "Snapshot mixes pre- and post-adoption state"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

classif %>%
  kable(caption = "Covariate Classification by Temporal Status") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

**Consequence:** Only the three structurally pre-treatment covariates
(repo age, language, organization type) are defensible controls.
Restricting to pre-treatment covariates eliminates mediator bias entirely
(pre-treatment variables cannot be caused by treatment) and reduces
collider risk to the minor M-bias scenario.

## 2.3 Pre-Treatment vs. Kitchen-Sink OLS (Table 2)

We now compare three specifications: the bivariate regression (no
controls), the pre-treatment-only model (the only defensible
specification), and the kitchen-sink model from Stage 1.

```{r ols_comparison}
# Bivariate
m_bivariate <- lm(log_avg_monthly_commits ~ treated, data = df)

# Pre-treatment only: the only defensible specification
m_pretreat <- lm(log_avg_monthly_commits ~ treated + repo_age_yr +
                   queried_language + is_org,
                 data = df)

ols_results <- bind_rows(
  extract_treated(m_bivariate, "Bivariate (no controls)"),
  extract_treated(m_pretreat,  "Pre-treatment covariates only"),
  extract_treated(m_kitchen,   "Kitchen-sink (all covariates)")
)

ols_results %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(as.numeric(p_val) < 0.001, "<0.001", sprintf("%.3f", p_val)),
    r2    = sprintf("%.3f", r2)
  ) %>%
  select(Model = model, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, `R²` = r2, N = n) %>%
  kable(caption = "Table 2: OLS Estimates of AI Adoption Effect on log(Monthly Commits)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The coefficient attenuates monotonically as controls are added.  The
drop from bivariate to pre-treatment is defensibly attributed to
confounding removal (these are unambiguously pre-treatment variables).
The further drop from pre-treatment to kitchen-sink is
**uninterpretable**: the collapsed graph is not a DAG, so no adjustment
formula tells us whether we are removing confounding or blocking causal
paths through temporally ambiguous variables.

## 2.4 Coefficient Stability Across Specifications

```{r coefficient_stability, fig.width=8, fig.height=3}
formulas <- list(
  "Bivariate"                          = log_avg_monthly_commits ~ treated,
  "+ Age + Language"                    = log_avg_monthly_commits ~ treated +
                                           repo_age_yr + queried_language,
  "Pre-treatment (+ Org)"              = log_avg_monthly_commits ~ treated +
                                           repo_age_yr + queried_language + is_org,
  "+ CI, Size, Releases"               = log_avg_monthly_commits ~ treated +
                                           repo_age_yr + queried_language + is_org +
                                           has_ci + log_size + log_releases,
  "Kitchen-sink (+ Stars, ...)"        = log_avg_monthly_commits ~ treated +
                                           repo_age_yr + queried_language + is_org +
                                           has_ci + log_size + log_releases +
                                           log_stars + log_forks + log_total_prs +
                                           log_total_issues
)

stability <- lapply(names(formulas), function(label) {
  mod <- lm(formulas[[label]], data = df)
  ct <- coeftest(mod, vcov = vcovHC(mod, type = "HC1"))
  idx <- which(rownames(ct) == "treated")
  data.frame(spec = label, coef = ct[idx, 1], se = ct[idx, 2],
             stringsAsFactors = FALSE)
}) %>%
  bind_rows() %>%
  mutate(spec = factor(spec, levels = names(formulas)))

pre_idx     <- which(levels(stability$spec) == "Pre-treatment (+ Org)")
kitchen_idx <- which(levels(stability$spec) == "Kitchen-sink (+ Stars, ...)")

ggplot(stability, aes(x = spec, y = coef)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  annotate("rect",
           xmin = pre_idx - 0.4, xmax = pre_idx + 0.4,
           ymin = -Inf, ymax = Inf,
           fill = "steelblue", alpha = 0.12) +
  annotate("rect",
           xmin = kitchen_idx - 0.4, xmax = kitchen_idx + 0.4,
           ymin = -Inf, ymax = Inf,
           fill = "darkorange", alpha = 0.10) +
  geom_pointrange(aes(ymin = coef - 1.96 * se, ymax = coef + 1.96 * se), size = 0.5) +
  labs(x = NULL, y = "Coefficient on AI Adoption (treated)",
       title = "Coefficient Stability Across Specifications",
       caption = paste("Blue: pre-treatment covariates only (defensible).",
                       "Orange: kitchen-sink (temporally ambiguous).",
                       "95% CI with HC1 SEs.")) +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

The trajectory is monotonically downward.  The blue-shaded pre-treatment
specification is the only defensible one; everything to its right adds
temporally ambiguous covariates whose inclusion is unjustified by any
valid DAG.  The coefficient *keeps falling*, and the question is whether
the unobserved confounders (developer skill, team culture) --- which are
plausibly *more* strongly associated with both treatment and outcome than
any observed covariate --- would continue the downward trajectory to zero.

## 2.5 Sensitivity Analysis

### Cinelli & Hazlett (sensemakr) on Pre-Treatment Model

```{r sensemakr_pretreat}
sense_pre <- sensemakr(m_pretreat, treatment = "treated",
                       benchmark_covariates = c("repo_age_yr", "is_org"),
                       kd = 1:3)

cat("=== Pre-treatment model: Sensitivity analysis ===\n")
summary(sense_pre)
```

```{r sensemakr_plot, fig.width=8, fig.height=5}
plot(sense_pre, sensitivity.of = "estimate")
```

The robustness value (RV) tells us how strong a single unobserved
confounder would need to be --- in terms of partial R-squared with both
treatment and outcome --- to nullify the estimate.  Consider "developer
skill" as a confounder: skilled developers are more likely to adopt AI
tools (early adopters, careful tooling configuration) *and* more likely
to commit frequently (more productive).  A confounder this strong is not
only plausible --- it is expected.

### Oster Proportional Selection Bounds

```{r oster_bounds}
oster_delta <- function(model_restricted, model_full, r_max_multiplier = 1.3) {
  b_r <- coef(model_restricted)["treated"]
  b_f <- coef(model_full)["treated"]
  r_r <- summary(model_restricted)$r.squared
  r_f <- summary(model_full)$r.squared
  r_max <- min(1, r_max_multiplier * r_f)

  delta <- (b_f * (r_max - r_f)) / ((b_r - b_f) * (r_f - r_r))
  return(data.frame(
    beta_restricted = b_r,
    beta_full       = b_f,
    R2_restricted   = r_r,
    R2_full         = r_f,
    R2_max          = r_max,
    delta_star      = delta
  ))
}

oster_pre <- oster_delta(m_bivariate, m_pretreat)

oster_pre %>%
  mutate(across(where(is.numeric), ~sprintf("%.3f", .))) %>%
  mutate(comparison = "Bivariate -> Pre-treatment") %>%
  select(Comparison = comparison, `beta restricted` = beta_restricted,
         `beta full` = beta_full, `R2 restricted` = R2_restricted,
         `R2 full` = R2_full, `R2 max` = R2_max, `delta*` = delta_star) %>%
  kable(caption = "Oster Proportional Selection Bounds") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The delta\* for bivariate -> pre-treatment may appear large, suggesting
robustness.  But this is misleading: the pre-treatment confounders (age,
language, org type) absorb only a modest share of the raw gap (the R-squared
increase is small), so the "ruler" against which delta\* is measured is
short.  A large delta\* does not mean the estimate is safe --- it means the
pre-treatment confounders are simply not very informative about the
selection process.  The most important confounders (developer skill, team
culture, organizational resources) are precisely the ones we cannot observe
from GitHub API data.

# Stage 3: Can a Better Estimator Help?

Stage 3 asks whether alternative selection-on-observables methods
(matching, IPW) yield different answers from OLS.  If the problem were
model misspecification (nonlinearity, poor overlap), matching or IPW might
help.  If the problem is unobserved confounding, all three estimators will
converge on the same biased answer --- because they share the same
identifying assumption (conditional ignorability given observables).

We use only the pre-treatment covariates, since they are the only
defensible adjustment set.

## 3.1 Propensity Score Matching

```{r psm_estimation}
ps_formula <- treated ~ repo_age_yr + queried_language + is_org

m_ps <- matchit(ps_formula, data = df, method = "nearest",
                distance = "glm", ratio = 1, replace = FALSE)

summary(m_ps)
```

```{r psm_ols}
matched_data <- match.data(m_ps)

m_psm <- lm(log_avg_monthly_commits ~ treated + repo_age_yr +
               queried_language + is_org,
             data = matched_data, weights = weights)

psm_ct <- coeftest(m_psm, vcov = vcovHC(m_psm, type = "HC1"))
idx <- which(rownames(psm_ct) == "treated")

psm_result <- data.frame(
  method = "PSM (1:1 NN)",
  coef   = psm_ct[idx, 1],
  se     = psm_ct[idx, 2],
  p_val  = psm_ct[idx, 4],
  n      = nrow(matched_data)
)
```

## 3.2 Inverse Probability Weighting

```{r ipw}
df$ps <- predict(glm(ps_formula, data = df, family = binomial), type = "response")

# IPW weights for ATT: treated get weight 1, controls get PS/(1-PS)
df$ipw_weight <- ifelse(df$treated == 1, 1, df$ps / (1 - df$ps))

ess_control <- sum(df$ipw_weight[df$treated == 0])^2 /
               sum(df$ipw_weight[df$treated == 0]^2)
cat(sprintf("IPW effective sample size (controls): %.0f of %d\n",
            ess_control, sum(df$treated == 0)))

m_ipw <- lm(log_avg_monthly_commits ~ treated + repo_age_yr +
               queried_language + is_org,
             data = df, weights = ipw_weight)

ipw_ct <- coeftest(m_ipw, vcov = vcovHC(m_ipw, type = "HC1"))
idx <- which(rownames(ipw_ct) == "treated")

ipw_result <- data.frame(
  method = "IPW (ATT)",
  coef   = ipw_ct[idx, 1],
  se     = ipw_ct[idx, 2],
  p_val  = ipw_ct[idx, 4],
  n      = nrow(df)
)
```

## 3.3 Three-Estimator Comparison (Table 3)

```{r three_estimator_comparison}
all_results <- bind_rows(
  ols_results %>%
    filter(model == "Pre-treatment covariates only") %>%
    transmute(method = "OLS (pre-treatment)", coef, se, p_val, n),
  psm_result,
  ipw_result
) %>%
  mutate(method = factor(method, levels = c("OLS (pre-treatment)",
                                            "PSM (1:1 NN)", "IPW (ATT)")))

all_results %>%
  mutate(
    coef  = sprintf("%.3f", coef),
    se    = sprintf("%.3f", se),
    p_val = ifelse(p_val < 0.001, "<0.001", sprintf("%.3f", p_val))
  ) %>%
  select(Method = method, Coefficient = coef, `Robust SE` = se,
         `p-value` = p_val, N = n) %>%
  kable(caption = "Table 3: Three Estimators, Same Pre-Treatment Covariates") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r three_estimator_plot, fig.width=8, fig.height=3}
ggplot(all_results, aes(x = method, y = coef)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = coef - 1.96 * se, ymax = coef + 1.96 * se,
                      color = method),
                  size = 0.8) +
  scale_color_manual(values = c("OLS (pre-treatment)" = "#4E79A7",
                                "PSM (1:1 NN)"        = "#59A14F",
                                "IPW (ATT)"           = "#F28E2B")) +
  labs(x = NULL, y = "Coefficient on AI Adoption",
       title = "Three Estimators Converge Under Pre-Treatment Covariates",
       caption = "All use pre-treatment covariates only. 95% CI with HC1 robust SEs.") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

The three estimators converge tightly.  Switching from regression to
matching to weighting changes nothing because all three share the same
identifying assumption --- conditional ignorability given pre-treatment
covariates --- and that assumption is violated by the unobserved
confounders that no GitHub API data can capture.  **Using PSM or IPW does
not automatically make a study "causal inference."**

# Synthesis: The Cross-Sectional Ceiling

```{r summary_table}
b_biv  <- sprintf("%.2f", coef(m_bivariate)["treated"])
b_pre  <- sprintf("%.2f", coef(m_pretreat)["treated"])
b_kit  <- sprintf("%.2f", coef(m_kitchen)["treated"])
drop_pct <- sprintf("%.0f", (1 - coef(m_pretreat)["treated"] /
                                  coef(m_bivariate)["treated"]) * 100)
d_pre  <- sprintf("%.2f", oster_pre$delta_star)

summary_stages <- data.frame(
  Stage = c("Stage 1: Naive analysis",
            "Stage 2: Temporal collapse diagnosis",
            "Stage 3: Estimator convergence"),
  Finding = c(
    paste0("L2+ repos have dramatically more monthly commits than L1 repos, ",
           "with massive imbalance on every covariate (all SMDs exceed +/-0.1). ",
           "Kitchen-sink OLS yields a positive, significant coefficient (",
           b_kit, "). A naive analyst might stop here."),
    paste0("Temporal collapse makes DAG-based covariate selection incoherent: ",
           "time-varying covariates (stars, PRs, CI, ...) are simultaneously ",
           "confounders and colliders in the collapsed graph. Only pre-treatment ",
           "covariates (age, language, org type) are defensible. Pre-treatment ",
           "coefficient (", b_pre, ") is ", drop_pct, "% below bivariate (",
           b_biv, ") but fragile: Oster delta* = ", d_pre,
           " (large because the pre-treatment covariates are weak proxies, ",
           "not because the estimate is robust)."),
    paste0("OLS, PSM, and IPW converge tightly under pre-treatment covariates ",
           "--- the estimator is not the bottleneck; the identifying assumption is.")
  )
)

summary_stages %>%
  kable(caption = "Cross-Sectional Diagnostic Summary") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Why the coefficient survives controls

It is tempting to interpret the persistently significant coefficient as
evidence of a genuine causal effect.  But the persistence is *exactly*
what selection bias predicts:

- **Developer skill** (unobserved) is plausibly the strongest confounder.
  Skilled developers both commit more code *and* adopt AI tools earlier.
  No GitHub API covariate captures this.
- **Organizational investment** (unobserved) drives both tooling adoption
  and development velocity.  Being organization-owned (which we observe)
  is a crude proxy; the actual confounders are team size, budget, and
  engineering culture.
- **Reverse causality** compounds the problem: high-velocity projects
  have more reason to invest in AI tooling, so the causal arrow may
  run from productivity to adoption rather than the reverse.

The observables we control for are *proxies* for the true confounders, not
the confounders themselves.  The proxies absorb some but not all of the
selection, which is why the coefficient attenuates but does not vanish.

## The bridge to longitudinal data

Cross-sectional data cannot separate "AI tools increase productivity"
from "productive teams adopt AI tools."  The identifying variation must
come from *within-repo temporal changes* --- comparing the same project
before and after AI adoption, so that all time-invariant confounders
(developer skill, org culture, project domain) are absorbed by
construction.  This is precisely what the longitudinal analysis in
Section 4.2 provides.
