---
title: "Causality in Software Engineering – Literature Review"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
```

```{r load-data}
papers <- read_csv("../data/se_papers_metadata.csv") |>
  mutate(year = as.integer(str_extract(venue, "\\d{4}")))
```

## Methods

We collected metadata for all main-research-track papers published at
three premier SE venues---ICSE, FSE, and ASE---from 2015 to 2025.
Paper titles and DOIs were obtained from DBLP proceedings pages, and
abstracts were retrieved via the Semantic Scholar batch API, with
OpenAlex as a backfill source for missing records. This produced a
corpus of `r nrow(papers)` papers stored in a single CSV.

Each paper was then independently classified by a large language model
(Claude Opus 4.6, `temperature = 0`, with structured JSON output) along three
dimensions: (1) whether the paper is an empirical SE study, (2) whether
it is a mining software repositories (MSR) study, and (3) whether it
employs a causal inference method, makes causal claims, or both. For
papers tagged as using causal methods, a follow-up pass assigned each
paper to one or more categories in the taxonomy defined below. The
deterministic decoding and explicit classification schema were chosen to
maximize reproducibility; all prompts and scripts are available in the
replication package.

## Taxonomy of Causal Methods

The table below defines each causal-method category we assign to
papers tagged `causal_type ∈ {method, both}` (column `causal_method_class`
in the CSV).  A paper's *primary* category is the first match in
taxonomy order; multi-label assignments are stored semicolon-separated.

```{r taxonomy-table}
taxonomy <- tribble(
  ~Category,                            ~Definition,                                                                                                         ~N,
  "Randomized Controlled Experiment",   "Participants or units are randomly assigned to treatment / control groups to isolate the causal effect of an intervention.",   37L,
  "Quasi-/Natural Experiment",          "Exploits naturally occurring variation or non-randomized designs (e.g., case-control, panel data, before-after) to approximate causal inference.",  7L,
  "Causal Graph / DAG",                 "Explicitly constructs a directed acyclic graph encoding assumed causal relationships among variables.",                          7L,
  "Causal Discovery",                   "Algorithmically learns the causal structure (graph) from observational or mixed data.",                                          3L,
  "Counterfactual Analysis",            "Reasons about outcomes under hypothetical alternative conditions (\"what if X had been different?\").",                          6L,
  "Interventional / do-Calculus",       "Estimates the effect of setting a variable to a value using Pearl's do-operator or explicit interventions (e.g., fault injection).", 4L,
  "Granger Causality",                  "Tests whether past values of one time series improve prediction of another, establishing temporal precedence.",                  2L,
  "Structural Equation Modeling",       "Specifies a system of equations with observed/latent variables to estimate direct and indirect causal effects.",                 2L,
  "Causal Impact Analysis",             "Uses Bayesian structural time-series models to estimate the causal effect of an event on a time-series outcome.",                1L,
  "Propensity Score Methods",           "Balances treatment and control groups on observed covariates via matching or weighting to reduce selection bias.",                1L,
  "Survival / Hazard Modeling",         "Applies time-to-event models (e.g., Cox proportional hazards) with a causal interpretation of covariates.",                      2L,
  "Causal Fault Localization",          "Applies causal reasoning (causal testing, causal contribution scores, causal debugging) to localize bugs, root causes, or fairness defects.", 17L,
  "Regression-based Causal Inference",  "Uses regression or observational analysis with explicit causal intent (e.g., controlling for confounders, causal-adjacent language).", 11L,
)

taxonomy <- taxonomy |> arrange(desc(N))

knitr::kable(taxonomy,
             col.names = c("Category", "Definition", "Papers (n)"),
             caption   = "Taxonomy of causal methods applied in SE papers (ASE + FSE + ICSE, 2015–2025). N = 100 papers with causal_type ∈ {method, both}.")
```

```{r taxonomy-bar, fig.width=9, fig.height=5}
taxonomy |>
  mutate(Category = fct_reorder(Category, N)) |>
  ggplot(aes(N, Category)) +
  geom_col(fill = "#0072B2") +
  geom_text(aes(label = N), hjust = -0.3, size = 3.5) +
  labs(title = "Causal Methods Used in SE Papers",
       subtitle = "Primary classification of 100 papers (ASE + FSE + ICSE, 2015–2025)",
       x = "Number of papers", y = NULL) +
  theme_minimal(base_size = 13) +
  xlim(0, 42)
```

```{r causal-trends, fig.width=9, fig.height=5.5}
summary_by_year <- papers |>
  group_by(year) |>
  summarise(
    `All papers`             = n(),
    `Empirical SE`           = sum(is_empirical == TRUE, na.rm = TRUE),
    `Asking causal RQs`      = sum(causal_type %in% c("claim", "both"), na.rm = TRUE),
    `Claiming causal methods` = sum(causal_type %in% c("method", "both"), na.rm = TRUE),
  ) |>
  pivot_longer(-year, names_to = "category", values_to = "count") |>
  mutate(category = factor(category,
    levels = c("All papers", "Empirical SE",
               "Asking causal RQs", "Claiming causal methods")))

ggplot(summary_by_year, aes(year, count, colour = category, shape = category)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = seq(2015, 2025, 1)) +
  scale_colour_manual(values = c("All papers" = "grey40",
                                 "Empirical SE" = "#0072B2",
                                 "Asking causal RQs" = "#E69F00",
                                 "Claiming causal methods" = "#D55E00")) +
  labs(
    title = "Papers Published vs. Papers Making Causal Claims",
    subtitle = "ASE + FSE + ICSE, 2015–2025",
    x = NULL, y = "Number of papers", colour = NULL, shape = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```
