---
title: "Revisiting Bogner & Merkel (MSR 2022): Selection Bias, OVB Analysis, and Design Thinking"
output: html_document
---

This notebook implements a progressive reanalysis of Bogner & Merkel (MSR 2022):
Stage 1 diagnoses selection bias in the original cross-sectional design,
Stage 2 demonstrates cross-sectional OLS and formally assesses omitted variable
bias using three complementary methods (coefficient instability, Cinelli &
Hazlett sensitivity analysis, and Oster proportional selection bounds),
and Stage 3 shows how reframing the treatment as type system adoption intensity
improves the design while remaining fundamentally limited by cross-sectional data.

# Setup

```{r setup, message=FALSE, warning=FALSE}
# Install any missing packages before loading
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "jsonlite", "knitr",
                   "kableExtra", "lmtest", "sandwich", "sensemakr",
                   "showtext", "scales")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(jsonlite)
library(knitr)
library(kableExtra)
library(lmtest)
library(sandwich)
library(sensemakr)
library(showtext)
library(scales)

font_add("Linux Libertine", regular = "LinLibertine_Rah.ttf")

showtext_auto()

if (!dir.exists("../data/js-vs-ts")) {
  unzip("../data/js-vs-ts.zip", exdir = "../data/js-vs-ts")
}

DATA_DIR <- "../data/js-vs-ts/data"
OUTCOMES <- c(
  "code_smells_ncloc"       = "Code smells / nLOC",
  "cog_complexity_ncloc"    = "Cognitive complexity / nLOC",
  "bug_fix_ratio"           = "Bug-fix commit ratio",
  "avg_bug_time_days"       = "Avg. bug resolution time (days)"
)
```

# Load and Merge Data

```{r load_data, message=FALSE, warning=FALSE}
# --- Load metrics CSVs ---
js_metrics <- read.csv(file.path(DATA_DIR, "JavaScriptMetrics.csv"))
ts_metrics <- read.csv(file.path(DATA_DIR, "TypeScriptMetrics.csv"))

# Harmonise column names and add language indicator
na_inf <- function(x) ifelse(is.infinite(x), NA_real_, x)

js_metrics <- js_metrics %>%
  rename(
    code_smells_ncloc    = code.smells_ncloc,
    cog_complexity_ncloc = cognitive.complexity_ncloc,
    bug_fix_ratio        = bug.fix.commits_ratio,
    avg_bug_time_days    = avg_bug.issue_time
  ) %>%
  mutate(
    lang              = "JavaScript",
    any_density       = NA_real_,
    avg_bug_time_days = na_inf(avg_bug_time_days / 86400),
    bug_fix_ratio     = na_inf(bug_fix_ratio),
    code_smells_ncloc = na_inf(code_smells_ncloc),
    cog_complexity_ncloc = na_inf(cog_complexity_ncloc)
  )

ts_metrics <- ts_metrics %>%
  rename(
    code_smells_ncloc    = code.smells_ncloc,
    cog_complexity_ncloc = cognitive.complexity_ncloc,
    bug_fix_ratio        = bug.fix.commits_ratio,
    avg_bug_time_days    = avg_bug.issue_time,
    any_density          = any.type.count_ncloc
  ) %>%
  mutate(
    lang              = "TypeScript",
    avg_bug_time_days = na_inf(avg_bug_time_days / 86400),
    bug_fix_ratio     = na_inf(bug_fix_ratio),
    code_smells_ncloc = na_inf(code_smells_ncloc),
    cog_complexity_ncloc = na_inf(cog_complexity_ncloc)
  )

metrics <- bind_rows(
  js_metrics %>% select(repo_name, lang, code_smells_ncloc, cog_complexity_ncloc,
                         bug_fix_ratio, avg_bug_time_days, any_density),
  ts_metrics %>% select(repo_name, lang, code_smells_ncloc, cog_complexity_ncloc,
                         bug_fix_ratio, avg_bug_time_days, any_density)
)

# --- Load characteristics CSVs ---
js_chars <- read.csv(file.path(DATA_DIR, "JavaScriptReposCharacteristics.csv")) %>%
  rename(
    bug_fix_commits = `bug.fix_commits_count`,
    code_smells     = code_smells
  )
ts_chars <- read.csv(file.path(DATA_DIR, "TypeScriptReposCharacteristics.csv")) %>%
  rename(
    bug_fix_commits = `bug.fix_commits_count`,
    any_type_count  = `any.type_count`
  )

chars <- bind_rows(
  js_chars %>% select(repo_name, ncloc, commits_count, framework),
  ts_chars  %>% select(repo_name, ncloc, commits_count, framework)
)

# --- Extract creation date and stars from JSON ---
read_repo_metadata <- function(path, repo_names) {
  dat   <- fromJSON(path)
  repos <- dat$repositories
  repos %>%
    filter(repoFullName %in% repo_names) %>%
    select(repo_name = repoFullName, creation_date = creationDate, stars)
}

target_repos <- metrics$repo_name
js_meta <- read_repo_metadata(file.path(DATA_DIR, "JavaScriptRepos.txt"), target_repos)
ts_meta <- read_repo_metadata(file.path(DATA_DIR, "TypeScriptRepos.txt"), target_repos)
meta    <- bind_rows(js_meta, ts_meta)

# --- Build unified analysis frame ---
df <- metrics %>%
  left_join(chars, by = "repo_name") %>%
  left_join(meta,  by = "repo_name") %>%
  mutate(
    is_ts         = as.integer(lang == "TypeScript"),
    creation_year = as.integer(substr(creation_date, 1, 4)),
    framework     = case_when(
      framework == "No framework identified" ~ "none",
      TRUE ~ framework
    ),
    framework = factor(framework),
    log_ncloc    = log(ncloc + 1),
    log_commits  = log(commits_count + 1),
    log_stars    = log(stars + 1)
  )

cat(sprintf(
  "Analysis frame: %d repos (%d JS, %d TS) | %d variables\n",
  nrow(df), sum(df$lang == "JavaScript"), sum(df$lang == "TypeScript"), ncol(df)
))
```

# Stage 1: Diagnostic Assessment

## 1.1 Replication of Original Descriptive Results

```{r replicate_descriptives}
desc <- df %>%
  group_by(lang) %>%
  summarise(
    N                       = n(),
    `Code smells / nLOC (median)`   = round(median(code_smells_ncloc, na.rm = TRUE), 4),
    `Cog. complexity / nLOC (median)` = round(median(cog_complexity_ncloc, na.rm = TRUE), 4),
    `Bug-fix ratio (median)`        = round(median(bug_fix_ratio, na.rm = TRUE), 3),
    `Avg. bug time, days (median)`  = round(median(avg_bug_time_days, na.rm = TRUE), 1),
    .groups = "drop"
  )

kable(desc, caption = "Table 1. Descriptive statistics by language (replicating Bogner & Merkel Table 2)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r mwu_tests}
outcomes_vec <- c("code_smells_ncloc", "cog_complexity_ncloc",
                  "bug_fix_ratio", "avg_bug_time_days")

mwu_results <- lapply(outcomes_vec, function(v) {
  x   <- df[[v]][df$lang == "JavaScript"]
  y   <- df[[v]][df$lang == "TypeScript"]
  tst <- wilcox.test(x, y, conf.int = TRUE)
  data.frame(
    Outcome   = OUTCOMES[v],
    W         = tst$statistic,
    p_value   = round(tst$p.value, 4),
    Direction = ifelse(median(x, na.rm = TRUE) > median(y, na.rm = TRUE), "TS better", "JS better")
  )
}) %>% bind_rows()

kable(mwu_results, row.names = FALSE,
      caption = "Table 2. Mann-Whitney U tests (replicating Bogner & Merkel)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 1.2 Covariate Balance Between JS and TS Repositories

The raw comparison in Section 1.1 makes no attempt to control for observable
differences between JS and TS repositories.
We quantify imbalance via standardised mean differences (SMD):
values above 0.1 indicate meaningful imbalance.

```{r covariate_balance}
smd <- function(x, g) {
  xj <- x[g == 0]; xt <- x[g == 1]
  (mean(xt, na.rm = TRUE) - mean(xj, na.rm = TRUE)) /
    sqrt((var(xj, na.rm = TRUE) + var(xt, na.rm = TRUE)) / 2)
}

covariates <- c("ncloc", "commits_count", "stars", "creation_year")
cov_labels <- c("nLOC", "Commit count", "Stars", "Creation year")

bal_tbl <- data.frame(
  Covariate = cov_labels,
  JS_median = sapply(covariates, function(v) median(df[[v]][df$is_ts == 0], na.rm = TRUE)),
  TS_median = sapply(covariates, function(v) median(df[[v]][df$is_ts == 1], na.rm = TRUE)),
  SMD       = sapply(covariates, function(v) round(smd(df[[v]], df$is_ts), 3))
)

# Framework distribution
fw_tbl <- df %>%
  count(lang, framework) %>%
  group_by(lang) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  select(lang, framework, pct) %>%
  pivot_wider(names_from = lang, values_from = pct, values_fill = 0) %>%
  arrange(desc(TypeScript))

kable(bal_tbl,
      caption = "Table 3. Covariate balance: JS vs. TS repositories (SMD > 0.1 indicates imbalance)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

kable(fw_tbl, caption = "Table 4. Framework distribution (%) by language") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r balance_plots, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
plot_vars <- data.frame(
  var   = c("ncloc", "commits_count", "stars", "creation_year"),
  label = c("nLOC", "Commit count", "Stars", "Creation year")
)

balance_long <- df %>%
  select(lang, ncloc, commits_count, stars, creation_year) %>%
  pivot_longer(-lang, names_to = "variable", values_to = "value") %>%
  left_join(plot_vars, by = c("variable" = "var")) %>%
  mutate(label = factor(label, levels = plot_vars$label))

ggplot(balance_long %>% filter(variable != "creation_year"),
       aes(x = lang, y = value, fill = lang)) +
  geom_violin(alpha = 0.6, draw_quantiles = 0.5) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_manual(values = c("JavaScript" = "#4E79A7", "TypeScript" = "#F28E2B")) +
  facet_wrap(~label, scales = "free_y", nrow = 1) +
  labs(x = NULL, y = "Value (log scale)", fill = "Language") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.x   = element_blank(),
    panel.grid.minor.x   = element_blank()
  )
```

```{r creation_year_plot, fig.width=8, fig.height=3}
ggplot(df, aes(x = creation_year, fill = lang)) +
  geom_bar(position = "dodge", width = 0.7, alpha = 0.85) +
  scale_fill_manual(values = c("JavaScript" = "#4E79A7", "TypeScript" = "#F28E2B")) +
  labs(x = "Creation Year", y = "Number of Repos", fill = "Language") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.x   = element_blank(),
    panel.grid.minor.x   = element_blank()
  )
```


# Stage 2: Cross-Sectional OLS and the Omitted Variable Bias Problem

Propensity score matching shares the same conditional ignorability assumption
as OLS: both adjust only for observables.
Since the core concern here is unobserved confounding (team quality,
engineering culture, organizational maturity), teaching both OLS and matching
would be redundant.
We use OLS as the sole cross-sectional estimator and devote our effort to
formally diagnosing the limits of that approach via three complementary OVB
assessment methods.

## 2.1 Unadjusted and Adjusted OLS

We fit two OLS models for each outcome: (M0) language indicator only,
replicating the original unadjusted comparison; (M5) language indicator plus
all observable confounders.
Shrinkage of the language coefficient from M0 to M5 measures how much of the
raw gap is explained by observable differences between JS and TS repositories.

```{r adjusted_regressions, message=FALSE, warning=FALSE}
fit_models <- function(outcome) {
  f0 <- as.formula(paste(outcome, "~ is_ts"))
  f1 <- as.formula(paste(outcome,
    "~ is_ts + log_ncloc + log_commits + log_stars + creation_year + framework"))

  m0 <- lm(f0, data = df)
  m1 <- lm(f1, data = df)

  extract_ts_coef <- function(m) {
    ct <- coeftest(m, vcov = vcovHC(m, type = "HC1"))
    data.frame(
      coef  = round(ct["is_ts", "Estimate"], 4),
      se    = round(ct["is_ts", "Std. Error"], 4),
      pval  = round(ct["is_ts", "Pr(>|t|)"], 4),
      n     = nobs(m),
      r2    = round(summary(m)$r.squared, 3)
    )
  }

  bind_rows(
    extract_ts_coef(m0) %>% mutate(model = "M0: Unadjusted"),
    extract_ts_coef(m1) %>% mutate(model = "M5: Fully adjusted")
  ) %>% mutate(outcome = OUTCOMES[outcome])
}

reg_results <- lapply(outcomes_vec, fit_models) %>% bind_rows()

kable(reg_results %>% select(outcome, model, coef, se, pval, n, r2),
      caption = paste0("Table 5. OLS regression: effect of TypeScript (is_ts) on each outcome.",
                       " Coef. = β for is_ts; HC1 robust SEs.",
                       " *** p<0.001, ** p<0.01, * p<0.05.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r coef_plot, fig.width=8, fig.height=3}
reg_results <- reg_results %>%
  mutate(
    ci_lo = coef - 1.96 * se,
    ci_hi = coef + 1.96 * se,
    model = factor(model, levels = c("M0: Unadjusted", "M5: Fully adjusted"))
  )

ggplot(reg_results, aes(x = coef, xmin = ci_lo, xmax = ci_hi,
                         y = outcome, colour = model, shape = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi),
                 height = 0.2, position = position_dodge(width = 0.5)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c("M0: Unadjusted" = "#4E79A7",
                                  "M5: Fully adjusted" = "#F28E2B")) +
  labs(x = "Coefficient for TypeScript (vs. JavaScript)",
       y = NULL, colour = "Model", shape = "Model") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.y   = element_blank()
  )
```

## 2.2 Coefficient Instability

If the coefficient is still drifting when we add the last observed control,
there is no reason to believe it would stabilise if we could add the
unobserved ones (team quality, engineering culture, management practices).
This is the intuition behind Altonji, Elder & Taber (2005): the trajectory
of the coefficient across control sets signals how sensitive it is to
omitted variables.

```{r coef_instability, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
control_steps <- list(
  "M0: None"             = "is_ts",
  "M1: + log(nLOC)"      = "is_ts + log_ncloc",
  "M2: + log(Commits)"   = "is_ts + log_ncloc + log_commits",
  "M3: + log(Stars)"     = "is_ts + log_ncloc + log_commits + log_stars",
  "M4: + Creation year"  = "is_ts + log_ncloc + log_commits + log_stars + creation_year",
  "M5: + Framework"      = "is_ts + log_ncloc + log_commits + log_stars + creation_year + framework"
)

stability_results <- lapply(outcomes_vec, function(v) {
  lapply(seq_along(control_steps), function(i) {
    f <- as.formula(paste(v, "~", control_steps[[i]]))
    m <- lm(f, data = df)
    ct <- coeftest(m, vcov = vcovHC(m, type = "HC1"))
    data.frame(
      outcome  = OUTCOMES[v],
      step     = names(control_steps)[i],
      step_num = i,
      coef     = ct["is_ts", "Estimate"],
      se       = ct["is_ts", "Std. Error"],
      r2       = summary(m)$r.squared
    )
  }) %>% bind_rows()
}) %>% bind_rows()

stability_results <- stability_results %>%
  mutate(
    ci_lo = coef - 1.96 * se,
    ci_hi = coef + 1.96 * se,
    step  = factor(step, levels = names(control_steps))
  )

ggplot(stability_results, aes(x = step_num, y = coef)) +
  geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), alpha = 0.15, fill = "#4E79A7") +
  geom_line(colour = "#4E79A7") +
  geom_point(size = 2, colour = "#4E79A7") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  facet_wrap(~outcome, scales = "free_y", nrow = 1) +
  scale_x_continuous(breaks = 1:6, labels = paste0("M", 0:5)) +
  labs(x = "Model (controls added sequentially)",
       y = "Coefficient for TypeScript") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    axis.title.y       = element_text(size = rel(0.8)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```

```{r instability_table}
# Show R² progression alongside the coefficient
stab_wide <- stability_results %>%
  mutate(label = sprintf("%.4f (%.4f)", coef, se)) %>%
  select(outcome, step, label, r2) %>%
  mutate(r2 = round(r2, 3))

kable(stability_results %>%
        select(outcome, step, coef, se, r2) %>%
        mutate(coef = round(coef, 4), se = round(se, 4), r2 = round(r2, 3)),
      caption = paste0("Table 6. Coefficient instability: TypeScript coefficient as controls are ",
                       "added sequentially. HC1 robust SEs. R² shows how much outcome variance ",
                       "the observed controls explain — the gap between R² and 1 represents the ",
                       "scope for unobserved confounders.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

**Interpretation**: If the coefficient changes substantially from M0 to M5,
the observed controls are "doing work."
But R² remains far below 1 for all outcomes, meaning most outcome variance is
unexplained — there is plenty of room for unobserved confounders to shift the
coefficient further.
The question is: how strong would those unobserved confounders need to be?

## 2.3 Sensitivity Analysis: Cinelli & Hazlett (2020)

The `sensemakr` framework answers a precise question: what is the minimum
strength of association (in partial R² terms) that an omitted confounder would
need to have with **both** the treatment and the outcome to reduce the
estimated effect to zero?

This quantity is the **robustness value** (RV).
A small RV means a weak confounder suffices to invalidate the result.
We benchmark the RV against the partial R² of observed covariates:
if an unobserved confounder only as strong as, say, project size could explain
away the effect, the finding is fragile.

```{r sensemakr_analysis, message=FALSE, warning=FALSE}
# Run sensemakr for each outcome
sens_list <- lapply(outcomes_vec, function(v) {
  sub <- df %>% filter(is.finite(.data[[v]]))
  f <- as.formula(paste(v,
    "~ is_ts + log_ncloc + log_commits + log_stars + creation_year + framework"))
  m <- lm(f, data = sub)
  sensemakr(model = m, treatment = "is_ts",
            benchmark_covariates = c("log_ncloc", "log_commits",
                                     "log_stars", "creation_year"),
            kd = 1:3, q = 1, alpha = 0.05)
})
names(sens_list) <- outcomes_vec

# Extract robustness values
sens_summary <- lapply(outcomes_vec, function(v) {
  s <- sens_list[[v]]$sensitivity_stats
  data.frame(
    outcome   = OUTCOMES[v],
    estimate  = round(s$estimate, 4),
    se        = round(s$se, 4),
    rv_q      = paste0(round(100 * s$rv_q, 1), "%"),
    rv_qa     = paste0(round(100 * s$rv_qa, 1), "%")
  )
}) %>% bind_rows()

kable(sens_summary,
      col.names = c("Outcome", "Estimate", "SE",
                     "RV (point est. = 0)", "RV (CI includes 0)"),
      caption = paste0("Table 7. Robustness values from Cinelli & Hazlett (2020). ",
                       "RV (point est. = 0): the minimum partial R² an omitted confounder ",
                       "would need with both treatment and outcome to reduce the estimate to zero. ",
                       "RV (CI includes 0): the minimum to make the 95% CI include zero.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r sensemakr_benchmarks}
# Show benchmark bounds for the lead outcome (code_smells_ncloc)
cat("=== Benchmark bounds for Code smells / nLOC ===\n\n")
summary(sens_list[["code_smells_ncloc"]])
```

```{r sensemakr_contour, fig.width=8, fig.height=5}
# Contour plot for the lead outcome
plot(sens_list[["code_smells_ncloc"]],
     sensitivity.of = "estimate")
```

**Interpretation**: The contour plot shows how the estimated TypeScript
coefficient changes for different levels of confounding (x-axis: partial R²
with treatment; y-axis: partial R² with outcome).
The red region is where the estimate crosses zero.
Diamond markers show where observed covariates fall.
If any single benchmark covariate is near the red region, a confounder of
comparable strength would suffice to invalidate the finding.

```{r sensemakr_contour_tstat, fig.width=8, fig.height=5}
plot(sens_list[["code_smells_ncloc"]],
     sensitivity.of = "t-value")
```

## 2.4 Oster (2019) Proportional Selection Bounds

Oster's approach asks a complementary question: if the relationship between
the treatment and unobserved confounders is proportional to the relationship
between the treatment and observed confounders, how large does that
proportionality factor δ need to be to drive the coefficient to zero?

The key formula is:
$$\delta^* = \frac{\hat{\beta}}{\tilde{\beta} - \hat{\beta}} \cdot \frac{\hat{R}^2 - \tilde{R}^2}{R^2_{\max} - \hat{R}^2}$$

where $\tilde{\beta}$ and $\tilde{R}^2$ are from the short (unadjusted)
regression, $\hat{\beta}$ and $\hat{R}^2$ from the long (adjusted) regression,
and $R^2_{\max}$ is the hypothetical R² with all confounders controlled.

- **δ\* > 1**: unobservables would need to be *more* important than
  observables to explain away the result — relatively robust.
- **0 < δ\* < 1**: unobservables need only be a *fraction* as important
  as observables — fragile.
- **δ\* < 0**: the coefficient moved *away* from zero when controls were
  added, meaning the observables work against the selection story.

Following Oster (2019), we set $R^2_{\max} = \min(1,\; 1.3 \times \hat{R}^2)$.

```{r oster_bounds}
oster_delta <- function(outcome, data, r2_max_mult = 1.3) {
  sub <- data %>% filter(is.finite(.data[[outcome]]))
  f_short <- as.formula(paste(outcome, "~ is_ts"))
  f_long  <- as.formula(paste(outcome,
    "~ is_ts + log_ncloc + log_commits + log_stars + creation_year + framework"))

  m_short <- lm(f_short, data = sub)
  m_long  <- lm(f_long, data = sub)

  beta_short <- coef(m_short)["is_ts"]
  beta_long  <- coef(m_long)["is_ts"]
  r2_short   <- summary(m_short)$r.squared
  r2_long    <- summary(m_long)$r.squared
  r2_max     <- min(1, r2_max_mult * r2_long)

  denom_beta <- beta_short - beta_long
  denom_r2   <- r2_max - r2_long

  if (abs(denom_beta) < 1e-10 || abs(denom_r2) < 1e-10) {
    delta <- NA_real_
  } else {
    delta <- beta_long * (r2_long - r2_short) / (denom_beta * denom_r2)
  }

  data.frame(
    outcome    = OUTCOMES[outcome],
    beta_short = round(beta_short, 4),
    beta_long  = round(beta_long, 4),
    r2_short   = round(r2_short, 3),
    r2_long    = round(r2_long, 3),
    r2_max     = round(r2_max, 3),
    delta      = round(delta, 2)
  )
}

oster_results <- lapply(outcomes_vec, oster_delta, data = df) %>% bind_rows()

kable(oster_results,
      col.names = c("Outcome", "β (short)", "β (long)",
                     "R² (short)", "R² (long)", "R²_max", "δ*"),
      caption = paste0("Table 8. Oster (2019) proportional selection bounds. ",
                       "δ* is the ratio of selection on unobservables to selection on ",
                       "observables needed to drive the coefficient to zero. ",
                       "δ* < 1 indicates fragility; δ* > 1 indicates robustness. ",
                       "R²_max = min(1, 1.3 × R²_long).")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r oster_sensitivity}
# Sensitivity to R²_max choice
r2_mults <- c(1.3, 2.0, Inf)
r2_labels <- c("1.3 × R²", "2.0 × R²", "R²_max = 1")

oster_sens <- lapply(seq_along(r2_mults), function(j) {
  lapply(outcomes_vec, function(v) {
    r <- oster_delta(v, df, r2_max_mult = r2_mults[j])
    r$r2_max_rule <- r2_labels[j]
    r
  }) %>% bind_rows()
}) %>% bind_rows()

kable(oster_sens %>% select(outcome, r2_max_rule, r2_max, delta),
      col.names = c("Outcome", "R²_max rule", "R²_max", "δ*"),
      caption = paste0("Table 9. Sensitivity of Oster δ* to the choice of R²_max. ",
                       "The tighter the R²_max bound, the more room there is for ",
                       "unobservables, and the smaller δ* becomes.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 2.5 The OVB Verdict

Three independent diagnostic methods converge:

1. **Coefficient instability** (Section 2.2): The TypeScript coefficient
   changes substantially as controls are added, and R² remains far from 1,
   leaving wide scope for further movement with unobserved confounders.

2. **Cinelli & Hazlett robustness values** (Section 2.3): The robustness
   values can be benchmarked against observed covariates — a confounder of
   comparable strength to project size or commit activity could plausibly
   explain away the remaining effect.

3. **Oster bounds** (Section 2.4): The δ* values quantify the proportional
   selection needed. If δ* is below or near 1, the result is fragile:
   unobserved confounders need not be much stronger than the observed ones
   to explain the entire effect.

**Direction of bias from the DAG**: Team capability and organizational
maturity are positively correlated with both (a) adopting TypeScript and
(b) having lower code smells and better practices.
This means omitted variable bias pushes the TypeScript coefficient in the
direction of "TS looks better" — which is exactly what the original study
reports.
The three OVB diagnostics confirm that this direction of bias is plausible
in magnitude, not just in sign.

**Bottom line**: The cross-sectional OLS estimate — no matter how many
observable controls we add — cannot credibly identify a causal effect of
TypeScript on code quality.
We need either a stronger design or a fundamentally different source of
variation.
Stage 3 explores the former; Stage 4 motivates the latter.


# Stage 3: A Cleverer Design — Type System Adoption Intensity

## 3.1 Reframing the Treatment

Comparing JavaScript to TypeScript compares *different projects in different
ecosystems*.
A sharper design restricts the comparison to within the TypeScript ecosystem,
using `any`-type density as a continuous measure of how much the type system
is actually used.

This improves the design in two ways:

1. **Sharper treatment definition**: The question becomes "how much does
   strict use of the type system associate with quality?" rather than the
   compound "JS vs. TS" comparison.
2. **Narrower comparison group**: All repos are TypeScript, so TS-specific
   tooling, community norms, and IDE integration are held constant.
   The remaining variation is within-ecosystem.

The threat, of course, is that repos choosing strict typing may differ
systematically from lenient-typing repos on unobservables (team discipline,
code review culture).
We assess whether observable balance improves, then estimate the dose-response.

## 3.2 Improved Covariate Balance Within TypeScript

```{r within_ts_balance}
ts_df <- df %>% filter(lang == "TypeScript", !is.na(any_density))

# Split at median any density
ts_df$any_high <- as.integer(ts_df$any_density >= median(ts_df$any_density))

# SMDs within TS (low-any vs. high-any)
within_ts_smds <- sapply(covariates, function(v) {
  round(smd(ts_df[[v]], ts_df$any_high), 3)
})

# Cross-language SMDs (from Stage 1)
cross_lang_smds <- sapply(covariates, function(v) {
  round(smd(df[[v]], df$is_ts), 3)
})

balance_comparison <- data.frame(
  Covariate = cov_labels,
  `JS vs TS (SMD)` = cross_lang_smds,
  `Low-any vs High-any TS (SMD)` = within_ts_smds,
  check.names = FALSE
)

kable(balance_comparison,
      caption = paste0("Table 10. Covariate balance comparison. ",
                       "Within-TS SMDs (low-any vs. high-any, median split) compared with ",
                       "cross-language SMDs (JS vs. TS). Smaller SMDs indicate a more ",
                       "balanced comparison on observables.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 3.3 Dose-Response Within TypeScript Repositories

```{r any_dose_response, message=FALSE, warning=FALSE}
any_reg <- lapply(outcomes_vec, function(v) {
  f  <- as.formula(paste(v,
    "~ any_density + log_ncloc + log_commits + creation_year"))
  m  <- lm(f, data = ts_df)
  ct <- coeftest(m, vcov = vcovHC(m, type = "HC1"))
  data.frame(
    outcome = OUTCOMES[v],
    coef    = round(ct["any_density", "Estimate"], 4),
    se      = round(ct["any_density", "Std. Error"], 4),
    pval    = round(ct["any_density", "Pr(>|t|)"], 4),
    n       = nobs(m),
    r2      = round(summary(m)$r.squared, 3)
  )
}) %>% bind_rows()

kable(any_reg,
      caption = paste0(
        "Table 11. Dose-response: effect of any-type density on quality metrics",
        " within the 305 TypeScript repositories.",
        " Controls: log(nLOC), log(commits), creation year. HC1 robust SEs.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r any_scatter_plots, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
plot_outcomes <- c("code_smells_ncloc", "cog_complexity_ncloc",
                   "bug_fix_ratio", "avg_bug_time_days")

ts_long <- ts_df %>%
  select(any_density, all_of(plot_outcomes)) %>%
  pivot_longer(-any_density, names_to = "variable", values_to = "value") %>%
  mutate(label = factor(OUTCOMES[variable], levels = OUTCOMES))

ggplot(ts_long, aes(x = any_density, y = value)) +
  geom_point(alpha = 0.35, size = 1.2, colour = "#F28E2B") +
  geom_smooth(method = "lm", se = TRUE, colour = "#333333", linewidth = 0.8) +
  facet_wrap(~label, scales = "free_y", nrow = 1) +
  labs(x = "any-Type Density (count / nLOC)", y = "Metric Value") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    axis.title.y       = element_text(size = rel(0.8)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```

## 3.4 Remaining Threats

The within-TS dose-response is a genuine design improvement over the raw
JS-vs-TS comparison: the treatment is sharper, the comparison group is narrower,
and observable balance is better.
But the fundamental problem remains: **this is still a cross-sectional
comparison across different repositories**.

Repos that use `any` heavily may be rushed or legacy projects with weaker
engineering practices; repos that adopt strict typing may be better-staffed
and more disciplined.
The direction of this confounding — disciplined teams → strict typing AND
higher quality — is the same as in the JS-vs-TS comparison, just attenuated.

We cannot resolve this with additional controls or matching, because the key
confounders (team discipline, code review culture, management quality) are
unobserved in these data.
What is needed is a design that exploits **within-project** variation.


# Stage 4: Why Cross-Sectional Data Hits a Ceiling

The three stages above establish a clear progression:

1. The raw JS-vs-TS comparison is severely confounded by selection bias.
2. OLS with observable controls attenuates the gap, but three independent OVB
   diagnostics show the remaining estimate is fragile — a confounder of
   plausible strength could explain it away entirely.
3. Reframing the treatment as `any`-type density within TypeScript improves the
   design (sharper treatment, better balance), but remains fundamentally limited
   by the cross-sectional data structure.

**The core limitation** is that cross-sectional data cannot distinguish
"strict typing improves quality" from "high-quality teams write strict TypeScript."
Both stories are consistent with the observed association.
Separating cause from selection requires **within-project temporal variation**:
observing the *same project* before and after it tightens its type discipline,
or comparing typed vs. untyped files *within the same repository* where
repo-level confounders (team quality, engineering culture, management) are held
constant by construction.

This motivates the next notebook, which collects new longitudinal data —
file-level commit histories, `any` density per file, and bug-fix rates over
time — and applies file-level fixed effects to exploit within-repo variation.
That design absorbs all time-invariant repo-level confounders and brings us
closer to a credible causal estimate.


# Summary

```{r summary_table}
summary_df <- data.frame(
  Stage = c(
    "Stage 1: Raw comparison (M0)",
    "Stage 2: Adjusted OLS (M5)",
    "Stage 2: OVB assessment",
    "Stage 3: any dose-response (within TS)"
  ),
  Finding = c(
    "TS appears better on code smells and complexity; worse on bug-fix ratio",
    "Coefficients attenuate with controls, but some remain significant",
    "All three OVB methods indicate fragility — plausible confounders could explain the remaining effect",
    "Higher any density associated with worse quality within TS, but cross-sectional design cannot rule out selection"
  ),
  Implication = c(
    "Confounded: JS and TS repos differ systematically on observables",
    "Observables explain part of the gap, but R² far from 1 leaves room for unobservables",
    "Cannot credibly claim a causal effect from cross-sectional OLS alone",
    "Design improvement, but within-project longitudinal variation needed for causal identification"
  ),
  check.names = FALSE
)

kable(summary_df,
      caption = "Table 12. Summary of findings across stages.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```
