---
title: "Revisiting Bogner & Merkel (MSR 2022): Selection Bias and Treatment Decomposition"
output: html_document
---

This notebook implements Stages 1–3 of the Example B reanalysis plan:
Stage 1 diagnoses selection bias in the original cross-sectional design,
Stage 2 applies propensity score matching to control for observable confounders,
and Stage 3 decomposes the compound "language" treatment via `any`-type density.

# Setup

```{r setup, message=FALSE, warning=FALSE}
# Install any missing packages before loading
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "jsonlite", "knitr",
                   "kableExtra", "MatchIt", "lmtest", "sandwich", "showtext",
                   "scales")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(jsonlite)
library(knitr)
library(kableExtra)
library(MatchIt)
library(lmtest)
library(sandwich)
library(showtext)
library(scales)

font_add("Linux Libertine", regular = "LinLibertine_R.otf")

showtext_auto()

DATA_DIR <- "../data/js-vs-ts/data"
OUTCOMES <- c(
  "code_smells_ncloc"       = "Code smells / nLOC",
  "cog_complexity_ncloc"    = "Cognitive complexity / nLOC",
  "bug_fix_ratio"           = "Bug-fix commit ratio",
  "avg_bug_time_days"       = "Avg. bug resolution time (days)"
)
```

# Load and Merge Data

```{r load_data, message=FALSE, warning=FALSE}
# --- Load metrics CSVs ---
js_metrics <- read.csv(file.path(DATA_DIR, "JavaScriptMetrics.csv"))
ts_metrics <- read.csv(file.path(DATA_DIR, "TypeScriptMetrics.csv"))

# Harmonise column names and add language indicator
na_inf <- function(x) ifelse(is.infinite(x), NA_real_, x)

js_metrics <- js_metrics %>%
  rename(
    code_smells_ncloc    = code.smells_ncloc,
    cog_complexity_ncloc = cognitive.complexity_ncloc,
    bug_fix_ratio        = bug.fix.commits_ratio,
    avg_bug_time_days    = avg_bug.issue_time
  ) %>%
  mutate(
    lang              = "JavaScript",
    any_density       = NA_real_,
    avg_bug_time_days = na_inf(avg_bug_time_days / 86400),
    bug_fix_ratio     = na_inf(bug_fix_ratio),
    code_smells_ncloc = na_inf(code_smells_ncloc),
    cog_complexity_ncloc = na_inf(cog_complexity_ncloc)
  )

ts_metrics <- ts_metrics %>%
  rename(
    code_smells_ncloc    = code.smells_ncloc,
    cog_complexity_ncloc = cognitive.complexity_ncloc,
    bug_fix_ratio        = bug.fix.commits_ratio,
    avg_bug_time_days    = avg_bug.issue_time,
    any_density          = any.type.count_ncloc
  ) %>%
  mutate(
    lang              = "TypeScript",
    avg_bug_time_days = na_inf(avg_bug_time_days / 86400),
    bug_fix_ratio     = na_inf(bug_fix_ratio),
    code_smells_ncloc = na_inf(code_smells_ncloc),
    cog_complexity_ncloc = na_inf(cog_complexity_ncloc)
  )

metrics <- bind_rows(
  js_metrics %>% select(repo_name, lang, code_smells_ncloc, cog_complexity_ncloc,
                         bug_fix_ratio, avg_bug_time_days, any_density),
  ts_metrics %>% select(repo_name, lang, code_smells_ncloc, cog_complexity_ncloc,
                         bug_fix_ratio, avg_bug_time_days, any_density)
)

# --- Load characteristics CSVs ---
js_chars <- read.csv(file.path(DATA_DIR, "JavaScriptReposCharacteristics.csv")) %>%
  rename(
    bug_fix_commits = `bug.fix_commits_count`,
    code_smells     = code_smells
  )
ts_chars <- read.csv(file.path(DATA_DIR, "TypeScriptReposCharacteristics.csv")) %>%
  rename(
    bug_fix_commits = `bug.fix_commits_count`,
    any_type_count  = `any.type_count`
  )

chars <- bind_rows(
  js_chars %>% select(repo_name, ncloc, commits_count, framework),
  ts_chars  %>% select(repo_name, ncloc, commits_count, framework)
)

# --- Extract creation date and stars from JSON ---
read_repo_metadata <- function(path, repo_names) {
  dat   <- fromJSON(path)
  repos <- dat$repositories
  repos %>%
    filter(repoFullName %in% repo_names) %>%
    select(repo_name = repoFullName, creation_date = creationDate, stars)
}

target_repos <- metrics$repo_name
js_meta <- read_repo_metadata(file.path(DATA_DIR, "JavaScriptRepos.txt"), target_repos)
ts_meta <- read_repo_metadata(file.path(DATA_DIR, "TypeScriptRepos.txt"), target_repos)
meta    <- bind_rows(js_meta, ts_meta)

# --- Build unified analysis frame ---
df <- metrics %>%
  left_join(chars, by = "repo_name") %>%
  left_join(meta,  by = "repo_name") %>%
  mutate(
    is_ts         = as.integer(lang == "TypeScript"),
    creation_year = as.integer(substr(creation_date, 1, 4)),
    framework     = case_when(
      framework == "No framework identified" ~ "none",
      TRUE ~ framework
    ),
    framework = factor(framework),
    log_ncloc    = log(ncloc + 1),
    log_commits  = log(commits_count + 1),
    log_stars    = log(stars + 1)
  )

cat(sprintf(
  "Analysis frame: %d repos (%d JS, %d TS) | %d variables\n",
  nrow(df), sum(df$lang == "JavaScript"), sum(df$lang == "TypeScript"), ncol(df)
))
```

# Stage 1: Diagnostic Assessment

## 1.1 Replication of Original Descriptive Results

```{r replicate_descriptives}
desc <- df %>%
  group_by(lang) %>%
  summarise(
    N                       = n(),
    `Code smells / nLOC (median)`   = round(median(code_smells_ncloc, na.rm = TRUE), 4),
    `Cog. complexity / nLOC (median)` = round(median(cog_complexity_ncloc, na.rm = TRUE), 4),
    `Bug-fix ratio (median)`        = round(median(bug_fix_ratio, na.rm = TRUE), 3),
    `Avg. bug time, days (median)`  = round(median(avg_bug_time_days, na.rm = TRUE), 1),
    .groups = "drop"
  )

kable(desc, caption = "Table 1. Descriptive statistics by language (replicating Bogner & Merkel Table 2)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r mwu_tests}
outcomes_vec <- c("code_smells_ncloc", "cog_complexity_ncloc",
                  "bug_fix_ratio", "avg_bug_time_days")

mwu_results <- lapply(outcomes_vec, function(v) {
  x   <- df[[v]][df$lang == "JavaScript"]
  y   <- df[[v]][df$lang == "TypeScript"]
  tst <- wilcox.test(x, y, conf.int = TRUE)
  data.frame(
    Outcome   = OUTCOMES[v],
    W         = tst$statistic,
    p_value   = round(tst$p.value, 4),
    Direction = ifelse(median(x, na.rm = TRUE) > median(y, na.rm = TRUE), "TS better", "JS better")
  )
}) %>% bind_rows()

kable(mwu_results, row.names = FALSE,
      caption = "Table 2. Mann-Whitney U tests (replicating Bogner & Merkel)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 1.2 Covariate Balance Between JS and TS Repositories

The raw comparison in Section 1.1 makes no attempt to control for observable
differences between JS and TS repositories.
We quantify imbalance via standardised mean differences (SMD):
values above 0.1 indicate meaningful imbalance.

```{r covariate_balance}
smd <- function(x, g) {
  xj <- x[g == 0]; xt <- x[g == 1]
  (mean(xt, na.rm = TRUE) - mean(xj, na.rm = TRUE)) /
    sqrt((var(xj, na.rm = TRUE) + var(xt, na.rm = TRUE)) / 2)
}

covariates <- c("ncloc", "commits_count", "stars", "creation_year")
cov_labels <- c("nLOC", "Commit count", "Stars", "Creation year")

bal_tbl <- data.frame(
  Covariate = cov_labels,
  JS_median = sapply(covariates, function(v) median(df[[v]][df$is_ts == 0], na.rm = TRUE)),
  TS_median = sapply(covariates, function(v) median(df[[v]][df$is_ts == 1], na.rm = TRUE)),
  SMD       = sapply(covariates, function(v) round(smd(df[[v]], df$is_ts), 3))
)

# Framework distribution
fw_tbl <- df %>%
  count(lang, framework) %>%
  group_by(lang) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  select(lang, framework, pct) %>%
  pivot_wider(names_from = lang, values_from = pct, values_fill = 0) %>%
  arrange(desc(TypeScript))

kable(bal_tbl,
      caption = "Table 3. Covariate balance: JS vs. TS repositories (SMD > 0.1 indicates imbalance)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

kable(fw_tbl, caption = "Table 4. Framework distribution (%) by language") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r balance_plots, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
plot_vars <- data.frame(
  var   = c("ncloc", "commits_count", "stars", "creation_year"),
  label = c("nLOC", "Commit count", "Stars", "Creation year")
)

balance_long <- df %>%
  select(lang, ncloc, commits_count, stars, creation_year) %>%
  pivot_longer(-lang, names_to = "variable", values_to = "value") %>%
  left_join(plot_vars, by = c("variable" = "var")) %>%
  mutate(label = factor(label, levels = plot_vars$label))

ggplot(balance_long %>% filter(variable != "creation_year"),
       aes(x = lang, y = value, fill = lang)) +
  geom_violin(alpha = 0.6, draw_quantiles = 0.5) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_manual(values = c("JavaScript" = "#4E79A7", "TypeScript" = "#F28E2B")) +
  facet_wrap(~label, scales = "free_y", nrow = 1) +
  labs(x = NULL, y = "Value (log scale)", fill = "Language") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.x   = element_blank(),
    panel.grid.minor.x   = element_blank()
  )

ggsave("../paper/figures/msr22_covariate_balance.pdf",
       width = 8, height = 3, device = cairo_pdf)
```

```{r creation_year_plot, fig.width=8, fig.height=3}
ggplot(df, aes(x = creation_year, fill = lang)) +
  geom_bar(position = "dodge", width = 0.7, alpha = 0.85) +
  scale_fill_manual(values = c("JavaScript" = "#4E79A7", "TypeScript" = "#F28E2B")) +
  labs(x = "Creation Year", y = "Number of Repos", fill = "Language") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.x   = element_blank(),
    panel.grid.minor.x   = element_blank()
  )

ggsave("../paper/figures/msr22_creation_year.pdf",
       width = 8, height = 3, device = cairo_pdf)
```

## 1.3 Adjusted OLS Regressions

We fit two OLS models for each outcome: (M0) language indicator only,
replicating the original unadjusted comparison; (M1) language indicator plus
observable confounders.
Shrinkage of the language coefficient from M0 to M1 measures how much of the
raw gap is explained by observable differences between JS and TS repositories.

```{r adjusted_regressions, message=FALSE, warning=FALSE}
fit_models <- function(outcome) {
  f0 <- as.formula(paste(outcome, "~ is_ts"))
  f1 <- as.formula(paste(outcome,
    "~ is_ts + log_ncloc + log_commits + log_stars + creation_year + framework"))

  m0 <- lm(f0, data = df)
  m1 <- lm(f1, data = df)

  extract_ts_coef <- function(m) {
    ct <- coeftest(m, vcov = vcovHC(m, type = "HC3"))
    data.frame(
      coef  = round(ct["is_ts", "Estimate"], 4),
      se    = round(ct["is_ts", "Std. Error"], 4),
      pval  = round(ct["is_ts", "Pr(>|t|)"], 4),
      n     = nobs(m),
      r2    = round(summary(m)$r.squared, 3)
    )
  }

  bind_rows(
    extract_ts_coef(m0) %>% mutate(model = "M0: Unadjusted"),
    extract_ts_coef(m1) %>% mutate(model = "M1: Adjusted")
  ) %>% mutate(outcome = OUTCOMES[outcome])
}

reg_results <- lapply(outcomes_vec, fit_models) %>% bind_rows()

reg_wide <- reg_results %>%
  mutate(label = sprintf("%.4f (%.4f)%s",
                         coef, se,
                         ifelse(pval < 0.001, "***",
                           ifelse(pval < 0.01, "**",
                             ifelse(pval < 0.05, "*", ""))))) %>%
  select(outcome, model, label, n, r2) %>%
  pivot_wider(names_from = model, values_from = c(label, n, r2))

kable(reg_results %>% select(outcome, model, coef, se, pval, n, r2),
      caption = paste0("Table 5. OLS regression: effect of TypeScript (is_ts) on each outcome.",
                       " Coef. = β for is_ts; HC3 robust SEs.",
                       " *** p<0.001, ** p<0.01, * p<0.05.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r coef_plot, fig.width=8, fig.height=3}
reg_results <- reg_results %>%
  mutate(
    ci_lo = coef - 1.96 * se,
    ci_hi = coef + 1.96 * se,
    model = factor(model, levels = c("M0: Unadjusted", "M1: Adjusted"))
  )

ggplot(reg_results, aes(x = coef, xmin = ci_lo, xmax = ci_hi,
                         y = outcome, colour = model, shape = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi),
                 height = 0.2, position = position_dodge(width = 0.5)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c("M0: Unadjusted" = "#4E79A7",
                                  "M1: Adjusted"   = "#F28E2B")) +
  labs(x = "Coefficient for TypeScript (vs. JavaScript)",
       y = NULL, colour = "Model", shape = "Model") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.y   = element_blank()
  )

ggsave("../paper/figures/msr22_adjusted_coefs.pdf",
       width = 8, height = 3, device = cairo_pdf)
```

# Stage 2: Propensity Score Matching

## 2.1 Propensity Score Model

```{r pscore_model}
df_match <- df %>%
  filter(!is.na(log_ncloc), !is.na(log_commits), !is.na(log_stars),
         !is.na(creation_year), !is.na(framework))

ps_formula <- is_ts ~ log_ncloc + log_commits + log_stars + creation_year + framework

ps_model <- glm(ps_formula, data = df_match, family = binomial())

df_match$pscore <- predict(ps_model, type = "response")

# C-statistic (AUC) computed via Wilcoxon rank-sum: no extra packages needed
c_stat <- wilcox.test(
  df_match$pscore[df_match$is_ts == 1],
  df_match$pscore[df_match$is_ts == 0]
)$statistic / (sum(df_match$is_ts == 1) * sum(df_match$is_ts == 0))
cat(sprintf("Propensity model C-statistic: %.3f\n", c_stat))

# Distribution of propensity scores
ggplot(df_match, aes(x = pscore, fill = lang)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("JavaScript" = "#4E79A7", "TypeScript" = "#F28E2B")) +
  labs(x = "Propensity Score P(TS | covariates)",
       y = "Number of Repos", fill = "Language") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position    = "bottom",
    legend.margin      = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin  = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y       = element_text(size = rel(0.8)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```

## 2.2 Nearest-Neighbour Matching (Caliper = 0.2 SD)

```{r matching, message=FALSE, warning=FALSE}
m_out <- matchit(ps_formula, data = df_match,
                 method   = "nearest",
                 distance = "logit",
                 caliper  = 0.2,
                 ratio    = 1)

print(summary(m_out)$nn)

# SMDs before and after matching
smd_match <- summary(m_out, standardize = TRUE)$sum.all[, "Std. Mean Diff."]
smd_after <- summary(m_out, standardize = TRUE)$sum.matched[, "Std. Mean Diff."]

smd_comparison <- data.frame(
  Covariate     = names(smd_match),
  SMD_before    = round(smd_match, 3),
  SMD_after     = round(smd_after, 3)
) %>%
  filter(Covariate %in% c("log_ncloc", "log_commits", "log_stars", "creation_year"))

smd_comparison$Covariate <- c("log(nLOC)", "log(Commits)", "log(Stars)", "Creation year")

kable(smd_comparison,
      caption = "Table 6. Standardised mean differences before and after matching") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 2.3 ATT Estimation in Matched Sample

```{r matched_estimates}
df_matched <- match.data(m_out)

att_results <- lapply(outcomes_vec, function(v) {
  sub <- df_matched %>% filter(is.finite(.data[[v]]))
  f  <- as.formula(paste(v, "~ is_ts"))
  m  <- lm(f, data = sub, weights = sub$weights)
  ct <- coeftest(m, vcov = vcovCL(m, cluster = ~subclass))
  data.frame(
    outcome = OUTCOMES[v],
    coef    = round(ct["is_ts", "Estimate"], 4),
    se      = round(ct["is_ts", "Std. Error"], 4),
    pval    = round(ct["is_ts", "Pr(>|t|)"], 4),
    n       = nrow(sub)
  )
}) %>% bind_rows()

kable(att_results,
      caption = paste0(
        "Table 7. ATT estimates in matched sample (1:1 NN, caliper 0.2 SD).",
        " Cluster-robust SEs at matched-pair level.",
        " Compare with M0 in Table 5 for attenuation.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r att_coef_plot, fig.width=8, fig.height=3}
att_plot <- att_results %>%
  mutate(
    ci_lo  = coef - 1.96 * se,
    ci_hi  = coef + 1.96 * se,
    source = "M2: Matched ATT"
  )

coef_compare <- bind_rows(
  reg_results %>%
    filter(model %in% c("M0: Unadjusted", "M1: Adjusted")) %>%
    select(outcome, coef, se, model) %>%
    rename(source = model) %>%
    mutate(ci_lo = coef - 1.96 * se, ci_hi = coef + 1.96 * se),
  att_plot %>% select(outcome, coef, se, source, ci_lo, ci_hi)
) %>%
  mutate(source = factor(source,
    levels = c("M0: Unadjusted", "M1: Adjusted", "M2: Matched ATT")))

ggplot(coef_compare, aes(x = coef, y = outcome, colour = source, shape = source)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi),
                 height = 0.2, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.6)) +
  scale_colour_manual(values = c(
    "M0: Unadjusted" = "#4E79A7",
    "M1: Adjusted"   = "#F28E2B",
    "M2: Matched ATT"= "#59A14F"
  )) +
  labs(x = "Coefficient for TypeScript (vs. JavaScript)",
       y = NULL, colour = "Estimator", shape = "Estimator") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position    = "bottom",
    legend.margin      = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin  = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y       = element_text(size = rel(0.8)),
    panel.grid.major.y = element_blank()
  )

ggsave("../paper/figures/msr22_estimator_comparison.pdf",
       width = 8, height = 3, device = cairo_pdf)
```

## 2.4 Sensitivity: Vary Caliper

```{r caliper_sensitivity, message=FALSE, warning=FALSE}
calipers <- c(0.1, 0.2, 0.3)

cal_results <- lapply(calipers, function(cal) {
  m_cal <- tryCatch(
    matchit(ps_formula, data = df_match,
            method = "nearest", distance = "logit", caliper = cal, ratio = 1),
    error = function(e) NULL
  )
  if (is.null(m_cal)) return(NULL)

  df_cal <- match.data(m_cal)
  lapply(outcomes_vec, function(v) {
    f  <- as.formula(paste(v, "~ is_ts"))
    m  <- lm(f, data = df_cal, weights = weights)
    ct <- coeftest(m, vcov = vcovCL(m, cluster = ~subclass))
    data.frame(
      caliper = cal,
      n_pairs = nrow(df_cal) / 2,
      outcome = OUTCOMES[v],
      coef    = round(ct["is_ts", "Estimate"], 4),
      pval    = round(ct["is_ts", "Pr(>|t|)"], 4)
    )
  }) %>% bind_rows()
}) %>% bind_rows()

kable(cal_results,
      caption = "Table 8. Sensitivity to caliper choice (ATT estimates and matched pairs)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

# Stage 3: Treatment Decomposition via `any`-Type Density

The TypeScript metric `any`-type density (count of `any` annotations per nLOC)
operationalises how much of the type system a repository actually uses.
A high `any` density indicates "TypeScript in name only": the codebase carries
the TypeScript syntax but escapes type checking through liberal use of `any`.

## 3.1 Dose-Response within TypeScript Repositories

```{r any_dose_response, message=FALSE, warning=FALSE}
ts_df <- df %>% filter(lang == "TypeScript", !is.na(any_density))

any_reg <- lapply(outcomes_vec, function(v) {
  f  <- as.formula(paste(v,
    "~ any_density + log_ncloc + log_commits + creation_year"))
  m  <- lm(f, data = ts_df)
  ct <- coeftest(m, vcov = vcovHC(m, type = "HC3"))
  data.frame(
    outcome = OUTCOMES[v],
    coef    = round(ct["any_density", "Estimate"], 4),
    se      = round(ct["any_density", "Std. Error"], 4),
    pval    = round(ct["any_density", "Pr(>|t|)"], 4),
    n       = nobs(m)
  )
}) %>% bind_rows()

kable(any_reg,
      caption = paste0(
        "Table 9. Dose-response: effect of any-type density on quality metrics",
        " within the 305 TypeScript repositories.",
        " Controls: log(nLOC), log(commits), creation year. HC3 robust SEs.",
        " *** p<0.001, ** p<0.01, * p<0.05.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r any_scatter_plots, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
plot_outcomes <- c("code_smells_ncloc", "cog_complexity_ncloc",
                   "bug_fix_ratio", "avg_bug_time_days")

ts_long <- ts_df %>%
  select(any_density, all_of(plot_outcomes)) %>%
  pivot_longer(-any_density, names_to = "variable", values_to = "value") %>%
  mutate(label = factor(OUTCOMES[variable], levels = OUTCOMES))

ggplot(ts_long, aes(x = any_density, y = value)) +
  geom_point(alpha = 0.35, size = 1.2, colour = "#F28E2B") +
  geom_smooth(method = "lm", se = TRUE, colour = "#333333", linewidth = 0.8) +
  facet_wrap(~label, scales = "free_y", nrow = 1) +
  labs(x = "any-type density (count / nLOC)", y = "Metric value") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    axis.title.y       = element_text(size = rel(0.8)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

ggsave("../paper/figures/msr22_any_dose_response.pdf",
       width = 8, height = 3, device = cairo_pdf)
```

## 3.2 Three-Level Treatment: JS, TS-lenient, TS-strict

We split the 305 TypeScript repositories into tertiles by `any`-type density.
The lowest tertile ("TS-strict") uses the type system most fully;
the highest tertile ("TS-lenient") closely approximates an untyped codebase.
Comparing JS, TS-lenient, and TS-strict creates a monotone gradient that
tests whether quality improves as type-system adoption deepens.

```{r three_level_treatment, message=FALSE, warning=FALSE}
ts_df <- ts_df %>%
  mutate(
    any_tertile = ntile(any_density, 3),
    group       = case_when(
      any_tertile == 1 ~ "TS-strict (low any)",
      any_tertile == 2 ~ "TS-mid",
      any_tertile == 3 ~ "TS-lenient (high any)"
    )
  )

js_df <- df %>%
  filter(lang == "JavaScript") %>%
  mutate(group = "JavaScript")

three_level_df <- bind_rows(
  js_df   %>% select(group, all_of(outcomes_vec)),
  ts_df   %>% select(group, all_of(outcomes_vec))
) %>%
  mutate(group = factor(group, levels = c(
    "JavaScript", "TS-lenient (high any)", "TS-mid", "TS-strict (low any)"
  )))

three_level_stats <- three_level_df %>%
  group_by(group) %>%
  summarise(
    N                               = n(),
    `Code smells / nLOC (median)`   = round(median(code_smells_ncloc, na.rm = TRUE), 4),
    `Cog. complexity / nLOC (median)` = round(median(cog_complexity_ncloc, na.rm = TRUE), 4),
    `Bug-fix ratio (median)`        = round(median(bug_fix_ratio, na.rm = TRUE), 3),
    `Avg. bug time, days (median)`  = round(median(avg_bug_time_days, na.rm = TRUE), 1),
    .groups = "drop"
  )

kable(three_level_stats,
      caption = paste0(
        "Table 10. Quality metrics by three-level treatment group.",
        " TS tertiles split by any-type density;",
        " JS repos from the matched sample (Table 6) appended as baseline.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r three_level_plot, fig.width=8, fig.height=3}
three_long <- three_level_df %>%
  pivot_longer(-group, names_to = "variable", values_to = "value") %>%
  mutate(label = factor(OUTCOMES[variable], levels = OUTCOMES))

ggplot(three_long, aes(x = group, y = value, fill = group)) +
  geom_violin(alpha = 0.7, draw_quantiles = 0.5, trim = TRUE) +
  scale_fill_manual(values = c(
    "JavaScript"            = "#4E79A7",
    "TS-lenient (high any)" = "#EDC948",
    "TS-mid"                = "#F28E2B",
    "TS-strict (low any)"   = "#59A14F"
  )) +
  facet_wrap(~label, scales = "free_y", nrow = 1) +
  labs(x = NULL, y = "Metric value", fill = NULL) +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    axis.text.x          = element_blank(),
    panel.grid.major.x   = element_blank(),
    panel.grid.minor.x   = element_blank()
  )

ggsave("../paper/figures/msr22_three_level.pdf",
       width = 8, height = 3, device = cairo_pdf)
```

## 3.3 Pairwise Tests across Three-Level Groups

```{r pairwise_tests, message=FALSE, warning=FALSE}
pairs <- list(
  c("JavaScript",            "TS-lenient (high any)"),
  c("JavaScript",            "TS-strict (low any)"),
  c("TS-lenient (high any)", "TS-strict (low any)")
)

pw_results <- lapply(outcomes_vec, function(v) {
  lapply(pairs, function(p) {
    x <- three_level_df[[v]][three_level_df$group == p[1]]
    y <- three_level_df[[v]][three_level_df$group == p[2]]
    tst <- wilcox.test(x, y)
    data.frame(
      outcome   = OUTCOMES[v],
      group_A   = p[1],
      group_B   = p[2],
      p_raw     = round(tst$p.value, 4),
      direction = ifelse(median(x, na.rm = TRUE) > median(y, na.rm = TRUE),
                         paste(p[2], "better"), paste(p[1], "better"))
    )
  }) %>% bind_rows()
}) %>% bind_rows() %>%
  mutate(p_bonf = round(p.adjust(p_raw, "bonferroni"), 4))

kable(pw_results,
      caption = "Table 11. Pairwise Mann-Whitney tests across treatment groups (Bonferroni corrected)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

# Summary

```{r summary_table}
summary_df <- data.frame(
  Stage = c(
    "Stage 1: Raw comparison (M0)",
    "Stage 1: Adjusted regression (M1)",
    "Stage 2: Matched ATT (M2)",
    "Stage 3: any dose-response (within TS)"
  ),
  `Code smells direction` = c(
    "TS better***", "TS better (attenuated?)", "TS better (attenuated?)", "Higher any → worse"
  ),
  `Cog. complexity direction` = c(
    "TS better***", "TS better (attenuated?)", "TS better (attenuated?)", "Higher any → worse"
  ),
  `Bug-fix ratio direction` = c(
    "TS worse***", "TS worse (attenuated?)", "TS worse (attenuated?)", "Higher any → more bugs"
  ),
  check.names = FALSE
)

kable(summary_df,
      caption = paste0(
        "Table 12. Direction of TypeScript effect across estimation strategies.",
        " Populated after running all chunks above.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```
