---
title: "Revisiting Bogner & Merkel (MSR 2022): Selection Bias, OVB Analysis, and Design Thinking"
output: html_document
---

This notebook implements a progressive reanalysis of Bogner & Merkel (MSR 2022),
who compared 299 JavaScript and 305 TypeScript GitHub repositories on four code
quality metrics and reported that TypeScript repos exhibit fewer code smells and
lower cognitive complexity but a higher bug-fix commit ratio.

Stage 1 diagnoses selection bias in the original cross-sectional design:
TS repos are systematically newer (creation year SMD = 0.75), more popular
(stars SMD = 0.40), and more active (commits SMD = 0.17) than their JS
counterparts.
A DAG-guided covariate selection exercise then reveals that GitHub stars —
the most "obvious" control variable — has both confounder and collider
components, and that conditioning on it risks introducing Berkson's paradox.
Stage 2 demonstrates cross-sectional OLS under two covariate specifications
(DAG-justified confounders only vs. kitchen-sink including ambiguous
variables) and formally assesses omitted variable bias using three
complementary methods (coefficient instability, Cinelli & Hazlett sensitivity
analysis, and Oster proportional selection bounds).
The DAG-justified and kitchen-sink specifications yield different estimates
(−0.0065 vs. −0.0044 for code smells) and different OVB conclusions
(robustness value 14.1% vs. 8.9%), demonstrating that covariate selection
is not a formality.
Stage 3 demonstrates that the estimator is not the bottleneck by showing
that propensity score matching (PSM) and inverse probability weighting (IPW)
produce nearly identical estimates to OLS under the same DAG-justified
covariates (−0.0065, −0.0060, −0.0065 for code smells), and that matching
does not materially improve robustness values (14.1% → 12.8%).
An appendix explores treatment decomposition via `any`-type density within
TypeScript repos.

# Setup

```{r setup, message=FALSE, warning=FALSE}
# Install any missing packages before loading
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "jsonlite", "knitr",
                   "kableExtra", "lmtest", "sandwich", "sensemakr",
                   "showtext", "scales")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(jsonlite)
library(knitr)
library(kableExtra)
library(lmtest)
library(sandwich)
library(sensemakr)
library(showtext)
library(scales)

font_add("Linux Libertine", regular = "LinLibertine_Rah.ttf")

showtext_auto()

if (!dir.exists("../data/js-vs-ts")) {
  unzip("../data/js-vs-ts.zip", exdir = "../data/js-vs-ts")
}

DATA_DIR <- "../data/js-vs-ts/data"
OUTCOMES <- c(
  "code_smells_ncloc"       = "Code smells / nLOC",
  "cog_complexity_ncloc"    = "Cognitive complexity / nLOC",
  "bug_fix_ratio"           = "Bug-fix commit ratio",
  "avg_bug_time_days"       = "Avg. bug resolution time (days)"
)
```

# Load and Merge Data

```{r load_data, message=FALSE, warning=FALSE}
# --- Load metrics CSVs ---
js_metrics <- read.csv(file.path(DATA_DIR, "JavaScriptMetrics.csv"))
ts_metrics <- read.csv(file.path(DATA_DIR, "TypeScriptMetrics.csv"))

# Harmonise column names and add language indicator
na_inf <- function(x) ifelse(is.infinite(x), NA_real_, x)

js_metrics <- js_metrics %>%
  rename(
    code_smells_ncloc    = code.smells_ncloc,
    cog_complexity_ncloc = cognitive.complexity_ncloc,
    bug_fix_ratio        = bug.fix.commits_ratio,
    avg_bug_time_days    = avg_bug.issue_time
  ) %>%
  mutate(
    lang              = "JavaScript",
    any_density       = NA_real_,
    avg_bug_time_days = na_inf(avg_bug_time_days / 86400),
    bug_fix_ratio     = na_inf(bug_fix_ratio),
    code_smells_ncloc = na_inf(code_smells_ncloc),
    cog_complexity_ncloc = na_inf(cog_complexity_ncloc)
  )

ts_metrics <- ts_metrics %>%
  rename(
    code_smells_ncloc    = code.smells_ncloc,
    cog_complexity_ncloc = cognitive.complexity_ncloc,
    bug_fix_ratio        = bug.fix.commits_ratio,
    avg_bug_time_days    = avg_bug.issue_time,
    any_density          = any.type.count_ncloc
  ) %>%
  mutate(
    lang              = "TypeScript",
    avg_bug_time_days = na_inf(avg_bug_time_days / 86400),
    bug_fix_ratio     = na_inf(bug_fix_ratio),
    code_smells_ncloc = na_inf(code_smells_ncloc),
    cog_complexity_ncloc = na_inf(cog_complexity_ncloc)
  )

metrics <- bind_rows(
  js_metrics %>% select(repo_name, lang, code_smells_ncloc, cog_complexity_ncloc,
                         bug_fix_ratio, avg_bug_time_days, any_density),
  ts_metrics %>% select(repo_name, lang, code_smells_ncloc, cog_complexity_ncloc,
                         bug_fix_ratio, avg_bug_time_days, any_density)
)

# --- Load characteristics CSVs ---
js_chars <- read.csv(file.path(DATA_DIR, "JavaScriptReposCharacteristics.csv")) %>%
  rename(
    bug_fix_commits = `bug.fix_commits_count`,
    code_smells     = code_smells
  )
ts_chars <- read.csv(file.path(DATA_DIR, "TypeScriptReposCharacteristics.csv")) %>%
  rename(
    bug_fix_commits = `bug.fix_commits_count`,
    any_type_count  = `any.type_count`
  )

chars <- bind_rows(
  js_chars %>% select(repo_name, ncloc, commits_count, framework),
  ts_chars  %>% select(repo_name, ncloc, commits_count, framework)
)

# --- Extract creation date and stars from JSON ---
read_repo_metadata <- function(path, repo_names) {
  dat   <- fromJSON(path)
  repos <- dat$repositories
  repos %>%
    filter(repoFullName %in% repo_names) %>%
    select(repo_name = repoFullName, creation_date = creationDate, stars)
}

target_repos <- metrics$repo_name
js_meta <- read_repo_metadata(file.path(DATA_DIR, "JavaScriptRepos.txt"), target_repos)
ts_meta <- read_repo_metadata(file.path(DATA_DIR, "TypeScriptRepos.txt"), target_repos)
meta    <- bind_rows(js_meta, ts_meta)

# --- Build unified analysis frame ---
df <- metrics %>%
  left_join(chars, by = "repo_name") %>%
  left_join(meta,  by = "repo_name") %>%
  mutate(
    is_ts         = as.integer(lang == "TypeScript"),
    creation_year = as.integer(substr(creation_date, 1, 4)),
    framework     = case_when(
      framework == "No framework identified" ~ "none",
      TRUE ~ framework
    ),
    framework = factor(framework),
    log_ncloc    = log(ncloc + 1),
    log_commits  = log(commits_count + 1),
    log_stars    = log(stars + 1)
  )

cat(sprintf(
  "Analysis frame: %d repos (%d JS, %d TS) | %d variables\n",
  nrow(df), sum(df$lang == "JavaScript"), sum(df$lang == "TypeScript"), ncol(df)
))
```

# Stage 1: Diagnostic Assessment

## 1.1 Replication of Original Descriptive Results

```{r replicate_descriptives}
desc <- df %>%
  group_by(lang) %>%
  summarise(
    N                       = n(),
    `Code smells / nLOC (median)`   = round(median(code_smells_ncloc, na.rm = TRUE), 4),
    `Cog. complexity / nLOC (median)` = round(median(cog_complexity_ncloc, na.rm = TRUE), 4),
    `Bug-fix ratio (median)`        = round(median(bug_fix_ratio, na.rm = TRUE), 3),
    `Avg. bug time, days (median)`  = round(median(avg_bug_time_days, na.rm = TRUE), 1),
    .groups = "drop"
  )

kable(desc, caption = "Table 1. Descriptive statistics by language (replicating Bogner & Merkel Table 2)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r mwu_tests}
outcomes_vec <- c("code_smells_ncloc", "cog_complexity_ncloc",
                  "bug_fix_ratio", "avg_bug_time_days")

mwu_results <- lapply(outcomes_vec, function(v) {
  x   <- df[[v]][df$lang == "JavaScript"]
  y   <- df[[v]][df$lang == "TypeScript"]
  tst <- wilcox.test(x, y, conf.int = TRUE)
  data.frame(
    Outcome   = OUTCOMES[v],
    W         = tst$statistic,
    p_value   = round(tst$p.value, 4),
    Direction = ifelse(median(x, na.rm = TRUE) > median(y, na.rm = TRUE), "TS better", "JS better")
  )
}) %>% bind_rows()

kable(mwu_results, row.names = FALSE,
      caption = "Table 2. Mann-Whitney U tests (replicating Bogner & Merkel)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 1.2 Covariate Balance Between JS and TS Repositories

The raw comparison in Section 1.1 makes no attempt to control for observable
differences between JS and TS repositories.
We quantify imbalance via standardised mean differences (SMD):
values above 0.1 indicate meaningful imbalance.

```{r covariate_balance}
smd <- function(x, g) {
  xj <- x[g == 0]; xt <- x[g == 1]
  (mean(xt, na.rm = TRUE) - mean(xj, na.rm = TRUE)) /
    sqrt((var(xj, na.rm = TRUE) + var(xt, na.rm = TRUE)) / 2)
}

covariates <- c("ncloc", "commits_count", "stars", "creation_year")
cov_labels <- c("nLOC", "Commit count", "Stars", "Creation year")

bal_tbl <- data.frame(
  Covariate = cov_labels,
  JS_median = sapply(covariates, function(v) median(df[[v]][df$is_ts == 0], na.rm = TRUE)),
  TS_median = sapply(covariates, function(v) median(df[[v]][df$is_ts == 1], na.rm = TRUE)),
  SMD       = sapply(covariates, function(v) round(smd(df[[v]], df$is_ts), 3))
)

# Framework distribution
fw_tbl <- df %>%
  count(lang, framework) %>%
  group_by(lang) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  select(lang, framework, pct) %>%
  pivot_wider(names_from = lang, values_from = pct, values_fill = 0) %>%
  arrange(desc(TypeScript))

kable(bal_tbl,
      caption = "Table 3. Covariate balance: JS vs. TS repositories (SMD > 0.1 indicates imbalance)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

kable(fw_tbl, caption = "Table 4. Framework distribution (%) by language") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r balance_plots, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
plot_vars <- data.frame(
  var   = c("ncloc", "commits_count", "stars", "creation_year"),
  label = c("nLOC", "Commit count", "Stars", "Creation year")
)

balance_long <- df %>%
  select(lang, ncloc, commits_count, stars, creation_year) %>%
  pivot_longer(-lang, names_to = "variable", values_to = "value") %>%
  left_join(plot_vars, by = c("variable" = "var")) %>%
  mutate(label = factor(label, levels = plot_vars$label))

ggplot(balance_long %>% filter(variable != "creation_year"),
       aes(x = lang, y = value, fill = lang)) +
  geom_violin(alpha = 0.6, draw_quantiles = 0.5) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_manual(values = c("JavaScript" = "#4E79A7", "TypeScript" = "#F28E2B")) +
  facet_wrap(~label, scales = "free_y", nrow = 1) +
  labs(x = NULL, y = "Value (log scale)", fill = "Language") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.x   = element_blank(),
    panel.grid.minor.x   = element_blank()
  )
```

```{r creation_year_plot, fig.width=8, fig.height=3}
ggplot(df, aes(x = creation_year, fill = lang)) +
  geom_bar(position = "dodge", width = 0.7, alpha = 0.85) +
  scale_fill_manual(values = c("JavaScript" = "#4E79A7", "TypeScript" = "#F28E2B")) +
  labs(x = "Creation Year", y = "Number of Repos", fill = "Language") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.x   = element_blank(),
    panel.grid.minor.x   = element_blank()
  )
```

## 1.3 DAG-Guided Covariate Selection

Before fitting regressions, we need to decide *which* covariates to include.
A common default is to control for "everything available," but DAG reasoning
shows this can do more harm than good: conditioning on a collider *opens*
a spurious path, and conditioning on a post-treatment variable can introduce
bias even when the variable looks like a confounder.

The paper's Figure 1 presents a DAG for the *language → defect proneness*
question (the Ray et al. setting). The Bogner & Merkel study asks a related
but distinct question: does choosing TypeScript (vs. JavaScript) improve
**code quality** as measured by static analysis metrics (code smells,
cognitive complexity) and process metrics (bug-fix ratio, bug resolution
time)? The causal mechanisms differ, so the DAG needs to be tailored.

**Plausible causal structure for Language → Code Quality:**

- **Confounders** (affect both language choice and code quality; must
  condition on):
  - *Team discipline / engineering culture*: Disciplined teams are more
    likely to adopt TypeScript *and* to enforce coding standards, conduct
    code reviews, and maintain lower code smell density. **Unobserved.**
  - *Project age* (`creation_year`): Newer projects are more likely to
    choose TypeScript (TS popularity grew from ~2017) and carry less
    accumulated technical debt. **Observed.**
  - *Project domain* (`framework`): Different domains have different
    complexity profiles and different language preferences. Angular was
    built for TypeScript and enforces strict architectural patterns,
    directly affecting code structure metrics. **Observed, but partially
    co-determined with treatment** (choosing TS may steer projects toward
    Angular, and choosing Angular may steer projects toward TS).
- **Mediators** (on the causal path; must *not* condition on):
  - *Type system enforcement*: TypeScript's compiler catches certain
    patterns that SonarQube would flag as code smells in JavaScript.
    This is the very mechanism we want to estimate.
  - *Tooling and linter configuration*: TypeScript projects may use
    stricter linting defaults. Conditioning on this would block part of
    the genuine effect.
- **Ambiguous variables** (could be confounders, could be post-treatment;
  conditioning is risky):
  - *Project size* (`log_ncloc`): Larger projects are more likely to
    adopt TypeScript (type safety pays off at scale) *and* have different
    code quality profiles. This is a plausible confounder. But language
    choice could also affect how large a project grows (e.g., TS's
    refactoring support enables more code), introducing a post-treatment
    component. On balance, size is more likely pre-determined than caused
    by language choice, so we include it in the DAG-justified set — but
    note the ambiguity.
  - *Commit activity* (`log_commits`): Total lifetime commits is
    measured at the snapshot and partly reflects post-treatment
    development activity. If TypeScript enables faster development
    (or attracts more active contributors), conditioning on commits
    could absorb part of the treatment effect.
  - *GitHub stars* (`log_stars`): Stars is the most problematic variable.
    It has *both* a confounder and a collider component:
    (a) *Confounder*: Well-resourced organisations (Microsoft, Google,
    Vercel) have high-visibility projects that are more likely to adopt
    TypeScript *and* to invest in code quality.
    (b) *Collider*: Good code quality in a popular language attracts
    stars; conditioning on a common effect of treatment and outcome opens
    a spurious path (Berkson's paradox).
    Since we cannot separate these two components from cross-sectional
    data, and the collider risk is concrete, we exclude stars from the
    DAG-justified specification.

This reasoning yields two covariate sets:

- **DAG-justified** (`M_dag`): `log_ncloc + creation_year + framework`.
  These are the observable variables most defensible as confounders, with
  the caveat that framework is partially co-determined with treatment and
  ncloc has a minor post-treatment component.
- **Kitchen-sink** (`M_all`): `log_ncloc + log_commits + log_stars +
  creation_year + framework`. This is the "control for everything available"
  specification. If the two specifications agree, covariate selection does
  not matter much; if they disagree, the DAG exercise has identified a
  real problem.

We fit both specifications below and compare them to show that covariate
selection is not a mere formality.


# Stage 2: Cross-Sectional OLS and the Omitted Variable Bias Problem

Propensity score matching shares the same conditional ignorability assumption
as OLS: both adjust only for observables.
Since the core concern here is unobserved confounding (team quality,
engineering culture, organizational maturity), teaching both OLS and matching
would be redundant.
We use OLS as the sole cross-sectional estimator and devote our effort to
formally diagnosing the limits of that approach via three complementary OVB
assessment methods.

## 2.1 Unadjusted, DAG-Justified, and Kitchen-Sink OLS

We fit three OLS models for each outcome, following the DAG reasoning in
Section 1.3:

- **M0** (unadjusted): Language indicator only, replicating the original
  comparison.
- **M_dag** (DAG-justified): Language indicator plus the covariates
  defensible as confounders — `log_ncloc`, `creation_year`, and
  `framework`. Excludes `log_stars` (ambiguous: collider + confounder)
  and `log_commits` (post-treatment concern).
- **M_all** (kitchen-sink): All available covariates including `log_stars`
  and `log_commits`.

Comparing M_dag and M_all reveals whether the ambiguous variables matter.
If stars acts partly as a collider, conditioning on it can shift the
coefficient in a predictable direction: among equally popular repos,
a TypeScript project's quality advantage is "explained away" by the
spurious path that the collider opens, attenuating the coefficient
toward zero (or even flipping its sign).

```{r adjusted_regressions, message=FALSE, warning=FALSE}
fit_models <- function(outcome) {
  f0   <- as.formula(paste(outcome, "~ is_ts"))
  fdag <- as.formula(paste(outcome,
    "~ is_ts + log_ncloc + creation_year + framework"))
  fall <- as.formula(paste(outcome,
    "~ is_ts + log_ncloc + log_commits + log_stars + creation_year + framework"))

  m0   <- lm(f0, data = df)
  mdag <- lm(fdag, data = df)
  mall <- lm(fall, data = df)

  extract_ts_coef <- function(m) {
    ct <- coeftest(m, vcov = vcovHC(m, type = "HC1"))
    data.frame(
      coef  = round(ct["is_ts", "Estimate"], 4),
      se    = round(ct["is_ts", "Std. Error"], 4),
      pval  = round(ct["is_ts", "Pr(>|t|)"], 4),
      n     = nobs(m),
      r2    = round(summary(m)$r.squared, 3)
    )
  }

  bind_rows(
    extract_ts_coef(m0)   %>% mutate(model = "M0: Unadjusted"),
    extract_ts_coef(mdag) %>% mutate(model = "M_dag: DAG-justified"),
    extract_ts_coef(mall) %>% mutate(model = "M_all: Kitchen-sink")
  ) %>% mutate(outcome = OUTCOMES[outcome])
}

reg_results <- lapply(outcomes_vec, fit_models) %>% bind_rows()

kable(reg_results %>% select(outcome, model, coef, se, pval, n, r2),
      caption = paste0("Table 5. OLS regression: effect of TypeScript (is_ts) on each outcome. ",
                       "M_dag controls for log(nLOC), creation year, and framework (DAG-justified confounders). ",
                       "M_all adds log(commits) and log(stars) (ambiguous post-treatment / collider variables). ",
                       "HC1 robust SEs.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r coef_plot, fig.width=8, fig.height=3}
reg_results <- reg_results %>%
  mutate(
    ci_lo = coef - 1.96 * se,
    ci_hi = coef + 1.96 * se,
    model = factor(model, levels = c("M0: Unadjusted",
                                      "M_dag: DAG-justified",
                                      "M_all: Kitchen-sink"))
  )

ggplot(reg_results, aes(x = coef, xmin = ci_lo, xmax = ci_hi,
                         y = outcome, colour = model, shape = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi),
                 height = 0.25, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.6)) +
  scale_colour_manual(values = c("M0: Unadjusted"       = "#4E79A7",
                                  "M_dag: DAG-justified" = "#59A14F",
                                  "M_all: Kitchen-sink"  = "#F28E2B")) +
  labs(x = "Coefficient for TypeScript (vs. JavaScript)",
       y = NULL, colour = "Model", shape = "Model") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.y   = element_blank()
  )
```

**Interpretation — covariate selection matters.**
For code smells / nLOC, the DAG-justified model (M_dag) estimates a
coefficient of −0.0065 (p < 0.001), while the kitchen-sink model (M_all)
gives −0.0044 (p = 0.004) — a further 32% attenuation from adding
commits and stars.
This additional shrinkage is ambiguous: it could reflect (a) legitimate
confounding that the DAG conservatively excluded, or (b) collider/post-treatment
bias from conditioning on stars and commits.
The direction is consistent with collider bias: among repos with equally
many stars, TS's quality advantage is partly "explained away" by the
spurious association that the opened collider path creates.
The sensitivity analysis in Section 2.3 will show that this 32% difference
has dramatic consequences for the OVB diagnosis: the robustness value for
code smells drops from 14.1% (M_dag) to 8.9% (M_all), and the threshold
for the 95% CI to include zero drops from 6.9% to just 1.2%.
The core lesson is that covariate selection is not a formality — a single
variable (stars) changes both the estimate and the conclusion about its
robustness.

## 2.2 Coefficient Instability

If the coefficient is still drifting when we add the last observed control,
there is no reason to believe it would stabilise if we could add the
unobserved ones (team quality, engineering culture, management practices).
This is the intuition behind Altonji, Elder & Taber (2005): the trajectory
of the coefficient across control sets signals how sensitive it is to
omitted variables.

We show two trajectories: the **DAG-justified path** adds only confounders
(nLOC → creation year → framework), and the **kitchen-sink path** continues
by adding the ambiguous variables (+ commits → + stars).
If the coefficient jumps when stars or commits enter, those variables are
"doing work" — but the DAG tells us that work may reflect collider bias or
post-treatment conditioning rather than genuine confounding control.

```{r coef_instability, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
control_steps <- list(
  "M0: None"             = "is_ts",
  "M1: + log(nLOC)"      = "is_ts + log_ncloc",
  "M2: + Creation year"  = "is_ts + log_ncloc + creation_year",
  "M3: + Framework"      = "is_ts + log_ncloc + creation_year + framework",
  "M4: + log(Commits)"   = "is_ts + log_ncloc + creation_year + framework + log_commits",
  "M5: + log(Stars)"     = "is_ts + log_ncloc + creation_year + framework + log_commits + log_stars"
)

# Mark which steps are DAG-justified vs. kitchen-sink extensions
step_zone <- c("DAG-justified", "DAG-justified", "DAG-justified",
               "DAG-justified", "Kitchen-sink", "Kitchen-sink")

stability_results <- lapply(outcomes_vec, function(v) {
  lapply(seq_along(control_steps), function(i) {
    f <- as.formula(paste(v, "~", control_steps[[i]]))
    m <- lm(f, data = df)
    ct <- coeftest(m, vcov = vcovHC(m, type = "HC1"))
    data.frame(
      outcome  = OUTCOMES[v],
      step     = names(control_steps)[i],
      step_num = i,
      zone     = step_zone[i],
      coef     = ct["is_ts", "Estimate"],
      se       = ct["is_ts", "Std. Error"],
      r2       = summary(m)$r.squared
    )
  }) %>% bind_rows()
}) %>% bind_rows()

stability_results <- stability_results %>%
  mutate(
    ci_lo = coef - 1.96 * se,
    ci_hi = coef + 1.96 * se,
    step  = factor(step, levels = names(control_steps)),
    zone  = factor(zone, levels = c("DAG-justified", "Kitchen-sink"))
  )

ggplot(stability_results, aes(x = step_num, y = coef)) +
  # Shade the kitchen-sink zone
  annotate("rect", xmin = 3.5, xmax = 6.5,
           ymin = -Inf, ymax = Inf,
           fill = "#E15759", alpha = 0.07) +
  geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), alpha = 0.15, fill = "#4E79A7") +
  geom_line(colour = "#4E79A7") +
  geom_point(aes(shape = zone), size = 2, colour = "#4E79A7") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  # Vertical line separating DAG-justified from kitchen-sink
  geom_vline(xintercept = 3.5, linetype = "dotted", colour = "#E15759", linewidth = 0.5) +
  facet_wrap(~outcome, scales = "free_y", nrow = 1) +
  scale_x_continuous(breaks = 1:6, labels = paste0("M", 0:5)) +
  scale_shape_manual(values = c("DAG-justified" = 16, "Kitchen-sink" = 17)) +
  labs(x = "Model (controls added sequentially)",
       y = "Coefficient for TypeScript",
       shape = "Covariate zone") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.x   = element_blank(),
    panel.grid.minor.x   = element_blank()
  )
```

```{r instability_table}
kable(stability_results %>%
        select(outcome, step, zone, coef, se, r2) %>%
        mutate(coef = round(coef, 4), se = round(se, 4), r2 = round(r2, 3)),
      caption = paste0("Table 6. Coefficient instability: TypeScript coefficient as controls are ",
                       "added sequentially. M0–M3 are DAG-justified confounders; M4–M5 add ",
                       "ambiguous post-treatment / collider variables (shaded region in the plot). ",
                       "HC1 robust SEs.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

**Interpretation**: The plot separates the DAG-justified confounders
(M0–M3, left of the dotted line) from the ambiguous variables (M4–M5,
shaded region).
The coefficient attenuates as DAG-justified confounders are added,
reflecting genuine selection bias being absorbed.
The key question is what happens at M4 and M5: if adding commits and
stars produces a *further* large drop, this additional attenuation is
ambiguous — it could reflect (a) legitimate confounding that the DAG
missed, (b) collider bias from conditioning on stars, or (c) post-treatment
bias from conditioning on commits.
In all cases, R² remains far below 1, meaning most outcome variance is
unexplained — there is plenty of room for *unobserved* confounders
(team discipline, engineering culture) to shift the coefficient further.

## 2.3 Sensitivity Analysis: Cinelli & Hazlett (2020)

The `sensemakr` framework answers a precise question: what is the minimum
strength of association (in partial R² terms) that an omitted confounder would
need to have with **both** the treatment and the outcome to reduce the
estimated effect to zero?

This quantity is the **robustness value** (RV).
A small RV means a weak confounder suffices to invalidate the result.
We benchmark the RV against the partial R² of observed covariates:
if an unobserved confounder only as strong as, say, project size could explain
away the effect, the finding is fragile.

We run `sensemakr` on **both** the DAG-justified model (M_dag) and the
kitchen-sink model (M_all) to show how covariate selection affects the
sensitivity analysis.

```{r sensemakr_analysis, message=FALSE, warning=FALSE}
# Run sensemakr on the DAG-justified model
sens_dag <- lapply(outcomes_vec, function(v) {
  sub <- df %>% filter(is.finite(.data[[v]]))
  f <- as.formula(paste(v,
    "~ is_ts + log_ncloc + creation_year + framework"))
  m <- lm(f, data = sub)
  sensemakr(model = m, treatment = "is_ts",
            benchmark_covariates = c("log_ncloc", "creation_year"),
            kd = 1:3, q = 1, alpha = 0.05)
})
names(sens_dag) <- outcomes_vec

# Run sensemakr on the kitchen-sink model
sens_all <- lapply(outcomes_vec, function(v) {
  sub <- df %>% filter(is.finite(.data[[v]]))
  f <- as.formula(paste(v,
    "~ is_ts + log_ncloc + log_commits + log_stars + creation_year + framework"))
  m <- lm(f, data = sub)
  sensemakr(model = m, treatment = "is_ts",
            benchmark_covariates = c("log_ncloc", "log_commits",
                                     "log_stars", "creation_year"),
            kd = 1:3, q = 1, alpha = 0.05)
})
names(sens_all) <- outcomes_vec

# Extract robustness values from both
extract_rv <- function(sens_list, label) {
  lapply(outcomes_vec, function(v) {
    s <- sens_list[[v]]$sensitivity_stats
    data.frame(
      outcome   = OUTCOMES[v],
      model     = label,
      estimate  = round(s$estimate, 4),
      se        = round(s$se, 4),
      rv_q      = paste0(round(100 * s$rv_q, 1), "%"),
      rv_qa     = paste0(round(100 * s$rv_qa, 1), "%")
    )
  }) %>% bind_rows()
}

sens_summary <- bind_rows(
  extract_rv(sens_dag, "M_dag"),
  extract_rv(sens_all, "M_all")
)

kable(sens_summary,
      col.names = c("Outcome", "Model", "Estimate", "SE",
                     "RV (point est. = 0)", "RV (CI includes 0)"),
      caption = paste0("Table 7. Robustness values from Cinelli & Hazlett (2020), ",
                       "comparing the DAG-justified (M_dag) and kitchen-sink (M_all) models. ",
                       "RV (point est. = 0): the minimum partial R² an omitted confounder ",
                       "would need with both treatment and outcome to reduce the estimate to zero. ",
                       "RV (CI includes 0): the minimum to make the 95% CI include zero.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r sensemakr_benchmarks}
# Show benchmark bounds for the DAG-justified model (lead outcome)
cat("=== Benchmark bounds for Code smells / nLOC (M_dag) ===\n\n")
summary(sens_dag[["code_smells_ncloc"]])
```

```{r sensemakr_contour_dag, fig.width=8, fig.height=5}
# Contour plot for DAG-justified model
plot(sens_dag[["code_smells_ncloc"]],
     sensitivity.of = "estimate")
title(sub = "DAG-justified model (M_dag): confounders only")
```

```{r sensemakr_contour_all, fig.width=8, fig.height=5}
# Contour plot for kitchen-sink model
plot(sens_all[["code_smells_ncloc"]],
     sensitivity.of = "estimate")
title(sub = "Kitchen-sink model (M_all): including stars and commits")
```

**Interpretation**: The comparison between the two contour plots reveals
how covariate selection affects the sensitivity analysis.
For code smells / nLOC, the DAG-justified model gives RV = 14.1% (point
estimate) and 6.9% (CI), while the kitchen-sink model gives RV = 8.9%
and 1.2%.
Conditioning on stars and commits has made the estimate *less* robust —
the RV for the CI to include zero dropped from 6.9% to just 1.2%.
This is consistent with collider bias: the additional variables
attenuated the coefficient (−0.0065 → −0.0044) without genuinely
removing confounding, so the smaller estimate is now even easier for
an unobserved confounder to explain away.
The pattern is similar for cognitive complexity (RV drops from 19.4% to
16.1%) and bug resolution time (RV drops from 11.1% to 8.3%).
The "same" sensitivity analysis applied to differently specified models
yields materially different conclusions about robustness — a vivid
demonstration that the DAG exercise is not a formality.

```{r sensemakr_contour_tstat, fig.width=8, fig.height=5}
plot(sens_dag[["code_smells_ncloc"]],
     sensitivity.of = "t-value")
title(sub = "DAG-justified model (M_dag)")
```

## 2.4 Oster (2019) Proportional Selection Bounds

Oster's approach asks a complementary question: if the relationship between
the treatment and unobserved confounders is proportional to the relationship
between the treatment and observed confounders, how large does that
proportionality factor δ need to be to drive the coefficient to zero?

The key formula is:
$$\delta^* = \frac{\hat{\beta}}{\tilde{\beta} - \hat{\beta}} \cdot \frac{\hat{R}^2 - \tilde{R}^2}{R^2_{\max} - \hat{R}^2}$$

where $\tilde{\beta}$ and $\tilde{R}^2$ are from the short (unadjusted)
regression, $\hat{\beta}$ and $\hat{R}^2$ from the long (adjusted) regression,
and $R^2_{\max}$ is the hypothetical R² with all confounders controlled.

- **δ\* > 1**: unobservables would need to be *more* important than
  observables to explain away the result — relatively robust.
- **0 < δ\* < 1**: unobservables need only be a *fraction* as important
  as observables — fragile.
- **δ\* < 0**: the coefficient moved *away* from zero when controls were
  added, meaning the observables work against the selection story.

Following Oster (2019), we set $R^2_{\max} = \min(1,\; 1.3 \times \hat{R}^2)$.

We compute Oster bounds for both the DAG-justified and kitchen-sink
specifications. The δ* values may differ because the two models
explain different amounts of variance (different R²), and because
the coefficient attenuation from M0 to M_dag differs from M0 to M_all.

```{r oster_bounds}
oster_delta <- function(outcome, data, long_formula, r2_max_mult = 1.3) {
  sub <- data %>% filter(is.finite(.data[[outcome]]))
  f_short <- as.formula(paste(outcome, "~ is_ts"))
  f_long  <- as.formula(paste(outcome, "~", long_formula))

  m_short <- lm(f_short, data = sub)
  m_long  <- lm(f_long, data = sub)

  beta_short <- coef(m_short)["is_ts"]
  beta_long  <- coef(m_long)["is_ts"]
  r2_short   <- summary(m_short)$r.squared
  r2_long    <- summary(m_long)$r.squared
  r2_max     <- min(1, r2_max_mult * r2_long)

  denom_beta <- beta_short - beta_long
  denom_r2   <- r2_max - r2_long

  if (abs(denom_beta) < 1e-10 || abs(denom_r2) < 1e-10) {
    delta <- NA_real_
  } else {
    delta <- beta_long * (r2_long - r2_short) / (denom_beta * denom_r2)
  }

  data.frame(
    outcome    = OUTCOMES[outcome],
    beta_short = round(beta_short, 4),
    beta_long  = round(beta_long, 4),
    r2_short   = round(r2_short, 3),
    r2_long    = round(r2_long, 3),
    r2_max     = round(r2_max, 3),
    delta      = round(delta, 2)
  )
}

f_dag <- "is_ts + log_ncloc + creation_year + framework"
f_all <- "is_ts + log_ncloc + log_commits + log_stars + creation_year + framework"

oster_dag <- lapply(outcomes_vec, oster_delta, data = df,
                    long_formula = f_dag) %>% bind_rows() %>%
  mutate(model = "M_dag")
oster_all <- lapply(outcomes_vec, oster_delta, data = df,
                    long_formula = f_all) %>% bind_rows() %>%
  mutate(model = "M_all")

oster_both <- bind_rows(oster_dag, oster_all)

kable(oster_both %>% select(outcome, model, beta_short, beta_long, r2_short, r2_long, r2_max, delta),
      col.names = c("Outcome", "Model", "β (short)", "β (long)",
                     "R² (short)", "R² (long)", "R²_max", "δ*"),
      caption = paste0("Table 8. Oster (2019) proportional selection bounds, ",
                       "comparing DAG-justified (M_dag) and kitchen-sink (M_all). ",
                       "δ* is the ratio of selection on unobservables to selection on ",
                       "observables needed to drive the coefficient to zero. ",
                       "δ* < 1 indicates fragility; δ* > 1 indicates robustness. ",
                       "R²_max = min(1, 1.3 × R²_long).")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r oster_sensitivity}
# Sensitivity to R²_max choice (using DAG-justified model as primary)
r2_mults <- c(1.3, 2.0, Inf)
r2_labels <- c("1.3 × R²", "2.0 × R²", "R²_max = 1")

oster_sens <- lapply(seq_along(r2_mults), function(j) {
  lapply(outcomes_vec, function(v) {
    r <- oster_delta(v, df, long_formula = f_dag, r2_max_mult = r2_mults[j])
    r$r2_max_rule <- r2_labels[j]
    r
  }) %>% bind_rows()
}) %>% bind_rows()

kable(oster_sens %>% select(outcome, r2_max_rule, r2_max, delta),
      col.names = c("Outcome", "R²_max rule", "R²_max", "δ*"),
      caption = paste0("Table 9. Sensitivity of Oster δ* to the choice of R²_max ",
                       "(DAG-justified model). ",
                       "The tighter the R²_max bound, the more room there is for ",
                       "unobservables, and the smaller δ* becomes.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 2.5 The OVB Verdict

Three independent diagnostic methods converge on the same conclusion,
and the DAG-guided covariate comparison (M_dag vs. M_all) adds a
fourth lesson about the process of causal reasoning itself:

1. **Coefficient instability** (Section 2.2): The TypeScript coefficient
   for code smells attenuates from −0.0122 (M0) to −0.0065 (M3, end of
   DAG-justified zone) — a 47% reduction. The transition into the
   kitchen-sink zone (M4–M5) produces a further 32% attenuation (to
   −0.0044), but this additional shrinkage is ambiguous: it may reflect
   genuine confounding or collider/post-treatment bias from conditioning
   on stars and commits.

2. **Cinelli & Hazlett robustness values** (Section 2.3): The DAG-justified
   model gives RV = 14.1% for code smells, while the kitchen-sink model
   gives RV = 8.9%. For the 95% CI, the drop is even more dramatic:
   6.9% → 1.2%. Conditioning on the ambiguous variables has made the
   estimate *less* robust, consistent with the variables attenuating the
   coefficient through collider bias rather than genuine confounding
   control.

3. **Oster bounds** (Section 2.4): The δ* values for the DAG-justified
   model differ from the kitchen-sink model because the two explain
   different amounts of variance. The kitchen-sink model's larger R²
   (from conditioning on stars) can give the misleading impression that
   less room remains for unobservables — but part of that R² may come
   from the collider mechanism, not from genuine confounding reduction.

4. **The DAG exercise itself**: Regardless of which specification is
   "right," the comparison shows that covariate selection is not
   a mere formality. Different choices yield different estimates and
   different OVB conclusions. The DAG forces the researcher to justify
   each variable's inclusion — and to discover that stars, the most
   "obvious" control, is also the most problematic.

**Direction of bias from the DAG**: Team capability and organizational
maturity are positively correlated with both (a) adopting TypeScript and
(b) having lower code smells and better practices.
This means omitted variable bias pushes the TypeScript coefficient in the
direction of "TS looks better" — which is exactly what the original study
reports.
The OVB diagnostics confirm that this direction of bias is plausible
in magnitude, not just in sign.

**Bottom line**: The cross-sectional OLS estimate — regardless of
covariate set — cannot credibly identify a causal effect of TypeScript
on code quality. The DAG exercise does not solve this problem, but it
makes the *reasoning* transparent: we know *what* we are assuming and
*where* the assumptions are weakest.

**A note on treatment decomposition.**
The DAG in Section 1.3 identifies "type system enforcement" as a mediator
on the causal path from language to code quality.
This suggests a natural treatment refinement: rather than comparing JS
vs. TS (a compound treatment bundling type system, tooling, ecosystem, and
community), we could measure how strictly the type system is *used* within
TypeScript, operationalized by `any`-type density.
Bogner and Merkel themselves observed that `any` density correlates with
quality within their TS sample (Spearman ρ = 0.17–0.26), but did not
pursue this with regression controls or causal reasoning.
The pragmatic stance formalizes this intuition: decompose compound
treatments into specific interventions with sharper estimands.
A dose-response analysis within the 305 TypeScript repos (Appendix A)
finds a significant association between `any` density and code smells
(β = 0.20, p < 0.001), but the remaining three metrics show no
significant association, and the cross-sectional design cannot
distinguish causation from selection (disciplined teams both use `any`
less *and* write cleaner code).
The refined treatment does, however, point naturally toward a stronger
design: observing within-project changes in `any` density over time.

Stage 3 now asks whether alternative selection-on-observables methods
(matching, IPW) yield different answers from OLS.


# Stage 3: Propensity Score Matching and Inverse Probability Weighting

The OVB assessment in Stage 2 established that the cross-sectional
estimate is fragile under OLS.
A natural follow-up question is: does the *estimator* matter?
OLS, propensity score matching (PSM), and inverse probability weighting
(IPW) are three different estimators for the same causal quantity under
the same identifying assumption — conditional ignorability given observed
covariates.
If they agree, the estimand is well-estimated conditional on the
assumption. If they disagree, one or more specifications may be
misspecified.
Either way, the OVB ceiling — unobserved confounding — applies equally
to all three.

Following the DAG reasoning in Section 1.3, we estimate the propensity
score using only the DAG-justified confounders (`log_ncloc`,
`creation_year`, `framework`), excluding `log_stars` (collider risk)
and `log_commits` (post-treatment).

## 3.1 Propensity Score Estimation and Overlap

```{r psm_setup, message=FALSE, warning=FALSE}
required_pkgs2 <- c("MatchIt", "cobalt")
missing_pkgs2 <- required_pkgs2[!sapply(required_pkgs2, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs2) > 0) {
  install.packages(missing_pkgs2, repos = "https://cloud.r-project.org")
}
library(MatchIt)
library(cobalt)
```

```{r propensity_score, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
# Fit propensity score model (DAG-justified covariates)
ps_mod <- glm(is_ts ~ log_ncloc + creation_year + framework,
              data = df, family = binomial)
df$ps <- predict(ps_mod, type = "response")

# Propensity score overlap plot
ggplot(df, aes(x = ps, fill = lang)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("JavaScript" = "#4E79A7", "TypeScript" = "#F28E2B")) +
  geom_vline(xintercept = c(min(df$ps[df$is_ts == 1]),
                              max(df$ps[df$is_ts == 0])),
             linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  labs(x = "Estimated Propensity Score P(TypeScript | X)",
       y = "Density", fill = "Language") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.x   = element_blank(),
    panel.grid.minor.x   = element_blank()
  )

# Common support statistics
cs_min <- max(min(df$ps[df$is_ts == 1]), min(df$ps[df$is_ts == 0]))
cs_max <- min(max(df$ps[df$is_ts == 1]), max(df$ps[df$is_ts == 0]))
in_support <- mean(df$ps >= cs_min & df$ps <= cs_max)
cat(sprintf("Common support region: [%.3f, %.3f]\n", cs_min, cs_max))
cat(sprintf("Fraction of repos in common support: %.1f%%\n", 100 * in_support))
```

The overlap plot reveals how well the JS and TS propensity score
distributions coincide.
If the distributions are well-separated, matching and IPW are
extrapolating rather than interpolating — and estimates become
sensitive to model specification.

## 3.2 Propensity Score Matching

```{r psm_matching, message=FALSE, warning=FALSE}
# 1:1 nearest-neighbor matching without replacement (DAG-justified PS)
m_nn <- matchit(is_ts ~ log_ncloc + creation_year + framework,
                data = df, method = "nearest", distance = "glm",
                replace = FALSE)

cat(sprintf("Matched pairs: %d (of %d TS repos)\n",
            sum(m_nn$weights[df$is_ts == 1] > 0),
            sum(df$is_ts == 1)))
cat(sprintf("Unmatched TS repos: %d\n",
            sum(m_nn$weights[df$is_ts == 1] == 0)))
cat(sprintf("Matched sample size: %d\n", sum(m_nn$weights > 0)))
```

```{r love_plot, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
# Love plot: SMDs before and after matching
love.plot(m_nn,
          stats = "mean.diffs",
          abs = TRUE,
          thresholds = c(m = 0.1),
          var.names = c(log_ncloc = "log(nLOC)",
                        creation_year = "Creation year",
                        log_stars = "log(Stars)",
                        log_commits = "log(Commits)"),
          colors = c("#E15759", "#59A14F"),
          shapes = c(17, 16),
          sample.names = c("Before matching", "After matching"),
          position = "bottom") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0)
  )
```

```{r psm_estimates, message=FALSE, warning=FALSE}
# Extract matched data and estimate ATT
md <- match.data(m_nn)

psm_results <- lapply(outcomes_vec, function(v) {
  sub <- md %>% filter(is.finite(.data[[v]]))
  f <- as.formula(paste(v, "~ is_ts + log_ncloc + creation_year + framework"))
  m <- lm(f, data = sub, weights = weights)
  ct <- coeftest(m, vcov = vcovHC(m, type = "HC1"))
  data.frame(
    outcome = OUTCOMES[v],
    method  = "PSM (1:1 NN)",
    coef    = round(ct["is_ts", "Estimate"], 4),
    se      = round(ct["is_ts", "Std. Error"], 4),
    pval    = round(ct["is_ts", "Pr(>|t|)"], 4),
    n       = nrow(sub)
  )
}) %>% bind_rows()
```

## 3.3 Inverse Probability Weighting

```{r ipw_estimates, message=FALSE, warning=FALSE}
# IPW weights for ATT: treated get weight 1, controls get PS/(1-PS)
df$ipw_att <- ifelse(df$is_ts == 1, 1, df$ps / (1 - df$ps))

# Effective sample size
ess_control <- sum(df$ipw_att[df$is_ts == 0])^2 /
               sum(df$ipw_att[df$is_ts == 0]^2)
cat(sprintf("IPW effective sample size (controls): %.0f of %d\n",
            ess_control, sum(df$is_ts == 0)))

ipw_results <- lapply(outcomes_vec, function(v) {
  sub <- df %>% filter(is.finite(.data[[v]]))
  f <- as.formula(paste(v, "~ is_ts"))
  m <- lm(f, data = sub, weights = ipw_att)
  ct <- coeftest(m, vcov = vcovHC(m, type = "HC1"))
  data.frame(
    outcome = OUTCOMES[v],
    method  = "IPW (ATT)",
    coef    = round(ct["is_ts", "Estimate"], 4),
    se      = round(ct["is_ts", "Std. Error"], 4),
    pval    = round(ct["is_ts", "Pr(>|t|)"], 4),
    n       = nrow(sub)
  )
}) %>% bind_rows()
```

## 3.4 Comparison: OLS, Matching, and IPW

```{r method_comparison}
# Gather OLS results from Stage 2
ols_dag_results <- lapply(outcomes_vec, function(v) {
  sub <- df %>% filter(is.finite(.data[[v]]))
  f <- as.formula(paste(v, "~ is_ts + log_ncloc + creation_year + framework"))
  m <- lm(f, data = sub)
  ct <- coeftest(m, vcov = vcovHC(m, type = "HC1"))
  data.frame(
    outcome = OUTCOMES[v],
    method  = "OLS (M_dag)",
    coef    = round(ct["is_ts", "Estimate"], 4),
    se      = round(ct["is_ts", "Std. Error"], 4),
    pval    = round(ct["is_ts", "Pr(>|t|)"], 4),
    n       = nrow(sub)
  )
}) %>% bind_rows()

all_methods <- bind_rows(ols_dag_results, psm_results, ipw_results) %>%
  mutate(method = factor(method, levels = c("OLS (M_dag)", "PSM (1:1 NN)", "IPW (ATT)")))

kable(all_methods,
      col.names = c("Outcome", "Method", "Coef", "SE", "p", "N"),
      caption = paste0("Table 10. Comparison of three selection-on-observables estimators. ",
                       "All use DAG-justified covariates (log(nLOC), creation year, framework). ",
                       "OLS adjusts via regression; PSM matches on propensity score; ",
                       "IPW re-weights controls. HC1 robust SEs.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r method_comparison_plot, fig.width=8, fig.height=3}
all_methods <- all_methods %>%
  mutate(ci_lo = coef - 1.96 * se,
         ci_hi = coef + 1.96 * se)

ggplot(all_methods, aes(x = coef, xmin = ci_lo, xmax = ci_hi,
                         y = outcome, colour = method, shape = method)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(height = 0.25, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.6)) +
  scale_colour_manual(values = c("OLS (M_dag)"  = "#59A14F",
                                  "PSM (1:1 NN)" = "#4E79A7",
                                  "IPW (ATT)"    = "#F28E2B")) +
  labs(x = "Coefficient for TypeScript (vs. JavaScript)",
       y = NULL, colour = "Method", shape = "Method") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    legend.position      = "bottom",
    legend.margin        = margin(t = -5, r = 0, b = -3, l = 0),
    legend.box.margin    = margin(t = -5, r = 0, b = -3, l = 0),
    axis.title.y         = element_text(size = rel(0.8)),
    panel.grid.major.y   = element_blank()
  )
```

## 3.5 OVB Diagnosis on Matched Sample

```{r psm_sensemakr, message=FALSE, warning=FALSE}
# Run sensemakr on matched sample for comparison with Stage 2
sens_psm <- lapply(outcomes_vec, function(v) {
  sub <- md %>% filter(is.finite(.data[[v]]))
  f <- as.formula(paste(v,
    "~ is_ts + log_ncloc + creation_year + framework"))
  m <- lm(f, data = sub, weights = weights)
  s <- sensemakr(model = m, treatment = "is_ts",
                 benchmark_covariates = c("log_ncloc", "creation_year"),
                 kd = 1:3, q = 1, alpha = 0.05)
  data.frame(
    outcome   = OUTCOMES[v],
    method    = "PSM (matched)",
    estimate  = round(s$sensitivity_stats$estimate, 4),
    rv_q      = paste0(round(100 * s$sensitivity_stats$rv_q, 1), "%"),
    rv_qa     = paste0(round(100 * s$sensitivity_stats$rv_qa, 1), "%")
  )
}) %>% bind_rows()

# Combine with OLS RVs from Stage 2
sens_ols_dag <- lapply(outcomes_vec, function(v) {
  s <- sens_dag[[v]]$sensitivity_stats
  data.frame(
    outcome   = OUTCOMES[v],
    method    = "OLS (M_dag)",
    estimate  = round(s$estimate, 4),
    rv_q      = paste0(round(100 * s$rv_q, 1), "%"),
    rv_qa     = paste0(round(100 * s$rv_qa, 1), "%")
  )
}) %>% bind_rows()

sens_compare <- bind_rows(sens_ols_dag, sens_psm)

kable(sens_compare,
      col.names = c("Outcome", "Method", "Estimate",
                     "RV (point = 0)", "RV (CI incl. 0)"),
      caption = paste0("Table 11. Robustness values: OLS vs. matched sample. ",
                       "If matching does not improve the RV, the fragility is ",
                       "driven by unobserved confounders, not by poor covariate ",
                       "balance on observables.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 3.6 The Selection-on-Observables Verdict

The results above deliver a clear and pedagogically important message:
**the estimator is not the bottleneck — the identifying assumption is.**

Three different estimators — OLS, propensity score matching, and inverse
probability weighting — all operating under the same DAG-justified covariate
set (`log_ncloc`, `creation_year`, `framework`) produce nearly identical
estimates.  For the lead outcome (code smells / nLOC), the coefficients are
−0.0065 (OLS), −0.0060 (PSM), and −0.0065 (IPW).  The pattern repeats for
cognitive complexity (−0.146, −0.143, −0.148) and bug-fix ratio (+0.059,
+0.060, +0.065).  The only meaningful divergence is for bug resolution time,
where PSM yields a marginally non-significant result (p = 0.061) due to the
slight loss of sample size from matching, while OLS (p = 0.020) and IPW
(p = 0.016) remain significant.

This convergence is not surprising — all three methods require the same
conditional ignorability assumption — but it is worth demonstrating
explicitly.  Researchers sometimes hope that matching will "fix" an
observational design where OLS fails.  The results here show that switching
estimators changes neither the point estimate nor its fragility.

The sensitivity analysis confirms this:  matching does not materially improve
robustness values.  For code smells, the RV drops slightly from 14.1% (OLS)
to 12.8% (PSM); for cognitive complexity it moves from 19.4% to 18.5%.
These are not meaningful improvements.  For bug resolution time, the RV
drops from 11.1% to 8.7% and the confidence interval already includes zero
on the matched sample (RV~α~ = 0%).

The IPW effective sample size is also informative: only 129 of 299 JS
control repos carry meaningful weight, confirming the overlap problems
visible in the propensity score distributions above.

**Bottom line**: All selection-on-observables methods converge on the same
fragile estimate.  The binding constraint is unobserved confounding — team
quality, engineering culture, organizational maturity — not the choice of
statistical method.  Breaking this ceiling requires a fundamentally different
identification strategy: within-project longitudinal variation.


# Stage 4: Why Cross-Sectional Data Hits a Ceiling

The stages above establish a clear progression:

1. **Selection bias is large**: TS repos are 3 years younger
   (median 2018 vs. 2015, SMD = 0.75), nearly 9× more popular
   (738 vs. 84 median stars, SMD = 0.40), and more active.
2. **DAG reasoning matters**: Conditioning on stars (collider risk)
   attenuates the coefficient by 32% and halves the robustness value.
   The "control for everything" default is not innocuous.
3. **The estimator is not the bottleneck**: OLS, matching, and IPW
   produce nearly identical estimates (−0.0065, −0.0060, −0.0065 for
   code smells) and nearly identical robustness values (14.1%, 12.8%).
   The binding constraint is unobserved confounding, not the choice of
   statistical method.

**The core limitation** is that cross-sectional data cannot distinguish
"TypeScript improves quality" from "high-quality teams adopt TypeScript."
Both stories are consistent with the observed association.
Separating cause from selection requires **within-project temporal variation**:
observing the *same project* before and after it tightens its type discipline,
or comparing typed vs. untyped files *within the same repository* where
repo-level confounders (team quality, engineering culture, management) are held
constant by construction.

This motivates the next notebook, which collects new longitudinal data —
file-level commit histories, `any` density per file, and bug-fix rates over
time — and applies file-level fixed effects to exploit within-repo variation.
That design absorbs all time-invariant repo-level confounders and brings us
closer to a credible causal estimate.


# Summary

```{r summary_table}
summary_df <- data.frame(
  Stage = c(
    "Stage 1: Raw comparison + covariate balance",
    "Stage 1: DAG-guided covariate selection",
    "Stage 2: DAG-justified vs. kitchen-sink OLS",
    "Stage 2: OVB assessment (sensemakr, Oster)",
    "Stage 3: PSM and IPW",
    "Stage 3: OVB on matched sample"
  ),
  Finding = c(
    "TS appears better on code smells and complexity; worse on bug-fix ratio. Large SMDs on stars (0.40) and creation year (0.75)",
    "Stars has both confounder and collider components; commits is post-treatment. Only creation_year, log_ncloc, and framework are defensible confounders",
    "DAG-justified and kitchen-sink models yield different coefficients (−0.0065 vs. −0.0044 for code smells); covariate selection is not innocuous",
    "All three OVB methods indicate fragility under both specifications — robustness value drops from 14.1% to 8.9% when conditioning on collider",
    "OLS, PSM, and IPW converge on similar estimates under the same DAG-justified covariates",
    "Matching does not materially improve robustness values — the fragility is driven by unobserved confounders"
  ),
  Implication = c(
    "Confounded: JS and TS repos differ systematically on observables",
    "DAG reasoning prevents mindless 'control for everything' — forces justification of each variable",
    "Conditioning on a collider (stars) or post-treatment variable (commits) changes conclusions",
    "Cannot credibly claim a causal effect from cross-sectional OLS regardless of covariate set",
    "The estimator is not the bottleneck — the identifying assumption is",
    "Need within-project longitudinal variation for causal identification"
  ),
  check.names = FALSE
)

kable(summary_df,
      caption = "Table 12. Summary of findings across stages.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```


# Appendix A: Treatment Refinement — `any`-Type Density Within TypeScript

Bogner and Merkel observed that `any`-type density correlates with quality
within their TypeScript sample (Spearman ρ = 0.17–0.26).
The pragmatic stance formalizes this as a treatment refinement: decompose
the compound "language" treatment into the specific "type strictness"
intervention, operationalized by `any` density.
This yields a sharper estimand but the same cross-sectional identification,
so the analysis is presented here as supplementary material.

```{r appendix_any_setup}
ts_df <- df %>% filter(lang == "TypeScript", !is.na(any_density))
ts_df$any_high <- as.integer(ts_df$any_density >= median(ts_df$any_density))
```

```{r appendix_any_balance}
within_ts_smds <- sapply(covariates, function(v) {
  round(smd(ts_df[[v]], ts_df$any_high), 3)
})
cross_lang_smds <- sapply(covariates, function(v) {
  round(smd(df[[v]], df$is_ts), 3)
})

balance_comparison <- data.frame(
  Covariate = cov_labels,
  `JS vs TS (SMD)` = cross_lang_smds,
  `Low-any vs High-any TS (SMD)` = within_ts_smds,
  check.names = FALSE
)

kable(balance_comparison,
      caption = paste0("Table A1. Covariate balance: JS vs. TS compared with ",
                       "within-TS (low-any vs. high-any, median split).")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r appendix_any_regression, message=FALSE, warning=FALSE}
any_reg <- lapply(outcomes_vec, function(v) {
  f  <- as.formula(paste(v,
    "~ any_density + log_ncloc + creation_year + framework"))
  m  <- lm(f, data = ts_df)
  ct <- coeftest(m, vcov = vcovHC(m, type = "HC1"))
  data.frame(
    outcome = OUTCOMES[v],
    coef    = round(ct["any_density", "Estimate"], 4),
    se      = round(ct["any_density", "Std. Error"], 4),
    pval    = round(ct["any_density", "Pr(>|t|)"], 4),
    n       = nobs(m),
    r2      = round(summary(m)$r.squared, 3)
  )
}) %>% bind_rows()

kable(any_reg,
      caption = paste0(
        "Table A2. Dose-response: effect of any-type density on quality metrics",
        " within TypeScript repositories.",
        " Controls: log(nLOC), creation year, framework. HC1 robust SEs.")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r appendix_any_scatter, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
ts_long <- ts_df %>%
  select(any_density, all_of(c("code_smells_ncloc", "cog_complexity_ncloc",
                                "bug_fix_ratio", "avg_bug_time_days"))) %>%
  pivot_longer(-any_density, names_to = "variable", values_to = "value") %>%
  mutate(label = factor(OUTCOMES[variable], levels = OUTCOMES))

ggplot(ts_long, aes(x = any_density, y = value)) +
  geom_point(alpha = 0.35, size = 1.2, colour = "#F28E2B") +
  geom_smooth(method = "lm", se = TRUE, colour = "#333333", linewidth = 0.8) +
  facet_wrap(~label, scales = "free_y", nrow = 1) +
  labs(x = "any-Type Density (count / nLOC)", y = "Metric Value") +
  theme_minimal(base_size = 13, base_family = "Linux Libertine") +
  theme(
    axis.title.y       = element_text(size = rel(0.8)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```
