---
title: "Bootstrap for Outcome Mislabeling (Berger et al. 2019)"
output: html_document
---

This notebook runs the Berger-style bootstrap that propagates outcome
mislabeling noise through the panel models.
It is kept separate from `toplas19_reanalysis.Rmd` because it is slow to run.

# Setup

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(MASS)
library(fixest)
library(parallel)
library(knitr)
library(kableExtra)

WORKING_DIR = "../data"

if (!file.exists(file.path(WORKING_DIR, "commits.csv"))) {
  unzip(file.path(WORKING_DIR, "commits.csv.zip"), exdir = WORKING_DIR)
}
```

# Helper Functions

```{r, helper_functions}
# Extract language coefficients as a named vector (zero-sum contrasts)
extractLangCoefs = function(model, lang_levels) {
    n_langs = length(lang_levels)
    if (inherits(model, "fixest")) {
        ct = as.data.frame(coeftable(model))
    } else {
        ct = as.data.frame(summary(model)$coefficients)
    }
    ct$name = rownames(ct)
    lang_rows = grep("^language", ct$name)
    lang_ct = ct[lang_rows, ]
    modeled = lang_levels[1:(n_langs - 1)]
    ref_coef = -sum(lang_ct[, 1])
    coefs = setNames(c(lang_ct[, 1], ref_coef), c(modeled, lang_levels[n_langs]))
    coefs
}
```

# Load Data

```{r, load_data}
data = read.csv(paste0(WORKING_DIR, "/commits.csv"))
cat(sprintf("Loaded %d commits | %d languages | %d repositories | %d developers\n",
    nrow(data), length(unique(data$language)), length(unique(data$project)), length(unique(data$author))))
```

# Prepare Panel Data

## Panel 1: Repository-Language Level (for M1)

```{r, data_preparation_panel_1}
panel_repo = data %>%
    group_by(project, language) %>%
    summarize(
        commits = n(),
        bcommits = sum(isbug),
        tins = sum(insertion),
        max_commit_age = max(commit_age),
        devs = n_distinct(author),
        .groups = "drop"
    ) %>%
    mutate(
        language = factor(language),
        project = factor(project),
        lcommits = log(commits),
        ltins = log(tins),
        lmax_commit_age = log(max_commit_age),
        ldevs = log(devs)
    )

contrasts(panel_repo$language) = contr.sum(nlevels(panel_repo$language))

cat(sprintf("Repository-Language panel: %d observations | %d repositories | %d languages\n",
    nrow(panel_repo), length(unique(panel_repo$project)), length(unique(panel_repo$language))))
```

## Panel 3: Developer-Repository-Language Level (for M4)

```{r, data_preparation_panel_3}
panel_dev_repo = data %>%
    group_by(author, project, language) %>%
    summarize(
        commits = n(),
        bcommits = sum(isbug),
        tins = sum(insertion),
        max_commit_age = max(commit_age),
        .groups = "drop"
    ) %>%
    mutate(
        language = factor(language),
        author = factor(author),
        project = factor(project),
        lcommits = log(commits),
        ltins = log(tins),
        lmax_commit_age = log(max_commit_age)
    )

contrasts(panel_dev_repo$language) = contr.sum(nlevels(panel_dev_repo$language))

cat(sprintf("Developer-Repository-Language panel: %d observations | %d developers | %d repositories | %d languages\n",
    nrow(panel_dev_repo), length(unique(panel_dev_repo$author)),
    length(unique(panel_dev_repo$project)), length(unique(panel_dev_repo$language))))
```

# Berger-Style Bootstrap for Outcome Mislabeling

Berger et al. (2019, Section 4.2.4) found that the keyword heuristic for
labeling bug-fix commits has a **36% false positive rate** and **11% false
negative rate**. Their bootstrap propagates this outcome measurement error
through the model by (1) resampling projects with replacement and (2)
regenerating bug-fix counts using Binomial draws that reflect the labeling
noise. We apply this procedure to M1 (pooled NBR, for direct comparison with
Berger's Table 6e) and M4 (two-way FE, our strongest specification).

```{r, bootstrap_setup}
FP = 0.36  # false positive rate (Berger Section 4.1.4)
FN = 0.11  # false negative rate
B_m1 = 100000  # M1: replicate Berger et al.'s 100k iterations
B_m4 = 1000    # M4: two-way FE is expensive, use fewer iterations
n_cores = detectCores()

# Set up log file following project convention
LOG_DIR = file.path(dirname(getwd()), "log")
dir.create(LOG_DIR, showWarnings = FALSE)
LOG_FILE = file.path(LOG_DIR, sprintf("bootstrap_%s.log",
    format(Sys.time(), "%Y%m%d_%H%M%S")))

log_msg = function(msg, level = "INFO") {
    line = sprintf("%s [%s] %s\n",
                   format(Sys.time(), "%Y-%m-%d %H:%M:%S"), level, msg)
    cat(line, file = LOG_FILE, append = TRUE)
}

log_msg(sprintf("Logging to %s", LOG_FILE))
log_msg(sprintf("Bootstrap config: M1=%d / M4=%d iterations, %d CPU cores",
    B_m1, B_m4, n_cores))
```

## M1 Bootstrap (Pooled NBR)

```{r, bootstrap_m1, warning=FALSE}
lang_levels_m1 = levels(panel_repo$language)
projects_m1 = unique(panel_repo$project)
n_projects_m1 = length(projects_m1)

# Pre-split data by project for fast resampling
proj_rows_m1 = split(seq_len(nrow(panel_repo)), panel_repo$project)

log_msg(sprintf("Running M1 bootstrap (%d iterations across %d cores)...", B_m1, n_cores))
t0 = Sys.time()

cl = makeCluster(n_cores)
clusterExport(cl, c("panel_repo", "proj_rows_m1", "projects_m1", "n_projects_m1",
                     "lang_levels_m1", "FP", "FN", "extractLangCoefs",
                     "B_m1", "LOG_FILE"))
clusterEvalQ(cl, { library(MASS) })
clusterSetRNGStream(cl, iseed = 42)

boot_results_m1 = parLapply(cl, 1:B_m1, function(b) {
    if (b %% 1000 == 0) {
        cat(sprintf("%s [INFO] [M1] %d / %d (%.0f%%)\n",
                    format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
                    b, B_m1, b / B_m1 * 100),
            file = LOG_FILE, append = TRUE)
    }
    sampled = sample(projects_m1, n_projects_m1, replace = TRUE)
    idx = unlist(proj_rows_m1[as.character(sampled)])
    boot_data = panel_repo[idx, ]

    boot_data$bcommits = rbinom(nrow(boot_data), boot_data$bcommits, 1 - FP) +
                         rbinom(nrow(boot_data), boot_data$commits - boot_data$bcommits, FN)

    fit = tryCatch(
        glm.nb(bcommits ~ lmax_commit_age + ltins + ldevs + lcommits + language,
               contrasts = list(language = contr.sum),
               data = boot_data),
        error = function(e) NULL, warning = function(w) suppressWarnings(
            glm.nb(bcommits ~ lmax_commit_age + ltins + ldevs + lcommits + language,
                   contrasts = list(language = contr.sum),
                   data = boot_data)
        )
    )
    if (!is.null(fit)) {
        tryCatch(extractLangCoefs(fit, lang_levels_m1),
                 error = function(e) rep(NA, length(lang_levels_m1)))
    } else {
        rep(NA, length(lang_levels_m1))
    }
})

stopCluster(cl)

boot_coefs_m1 = do.call(rbind, boot_results_m1)
colnames(boot_coefs_m1) = lang_levels_m1

elapsed_m1 = difftime(Sys.time(), t0, units = "mins")
n_success_m1 = sum(complete.cases(boot_coefs_m1))
log_msg(sprintf("M1 done in %.1f minutes. %d/%d iterations converged.",
    elapsed_m1, n_success_m1, B_m1))
```

## M4 Bootstrap (Two-Way FE Poisson)

```{r, bootstrap_m4, warning=FALSE}
lang_levels_m4 = levels(panel_dev_repo$language)
projects_m4 = unique(panel_dev_repo$project)
n_projects_m4 = length(projects_m4)

# Pre-split data by project for fast resampling
proj_rows_m4 = split(seq_len(nrow(panel_dev_repo)), panel_dev_repo$project)

log_msg(sprintf("Running M4 bootstrap (%d iterations across %d cores)...", B_m4, n_cores))
t0 = Sys.time()

cl = makeCluster(n_cores)
clusterExport(cl, c("panel_dev_repo", "proj_rows_m4", "projects_m4", "n_projects_m4",
                     "lang_levels_m4", "FP", "FN", "extractLangCoefs",
                     "B_m4", "LOG_FILE"))
clusterEvalQ(cl, { library(fixest) })
clusterSetRNGStream(cl, iseed = 123)

boot_results_m4 = parLapply(cl, 1:B_m4, function(b) {
    if (b %% 100 == 0) {
        cat(sprintf("%s [INFO] [M4] %d / %d (%.0f%%)\n",
                    format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
                    b, B_m4, b / B_m4 * 100),
            file = LOG_FILE, append = TRUE)
    }
    sampled = sample(projects_m4, n_projects_m4, replace = TRUE)
    idx = unlist(proj_rows_m4[as.character(sampled)])
    boot_data = panel_dev_repo[idx, ]

    boot_data$bcommits = rbinom(nrow(boot_data), boot_data$bcommits, 1 - FP) +
                         rbinom(nrow(boot_data), boot_data$commits - boot_data$bcommits, FN)

    boot_data$author = factor(boot_data$author)
    boot_data$project = factor(boot_data$project)
    boot_data$language = factor(boot_data$language)
    contrasts(boot_data$language) = contr.sum(nlevels(boot_data$language))

    fit = tryCatch(
        fepois(bcommits ~ lmax_commit_age + ltins + lcommits + language | author + project,
               data = boot_data, vcov = ~author + project),
        error = function(e) NULL, warning = function(w) tryCatch(
            suppressWarnings(
                fepois(bcommits ~ lmax_commit_age + ltins + lcommits + language | author + project,
                       data = boot_data, vcov = ~author + project)
            ), error = function(e) NULL
        )
    )
    if (!is.null(fit)) {
        tryCatch(extractLangCoefs(fit, lang_levels_m4),
                 error = function(e) rep(NA, length(lang_levels_m4)))
    } else {
        rep(NA, length(lang_levels_m4))
    }
})

stopCluster(cl)

boot_coefs_m4 = do.call(rbind, boot_results_m4)
colnames(boot_coefs_m4) = lang_levels_m4

elapsed_m4 = difftime(Sys.time(), t0, units = "mins")
n_success_m4 = sum(complete.cases(boot_coefs_m4))
log_msg(sprintf("M4 done in %.1f minutes. %d/%d iterations converged.",
    elapsed_m4, n_success_m4, B_m4))
```

## Bootstrap Results

```{r, bootstrap_results}
# Summarize bootstrap: mean coef, SE (sd of bootstrap dist), Bonferroni significance
summarizeBootstrap = function(boot_matrix, alpha = 0.01) {
    n_langs = ncol(boot_matrix)
    lang_names = colnames(boot_matrix)

    mean_coef = colMeans(boot_matrix, na.rm = TRUE)
    boot_se = apply(boot_matrix, 2, sd, na.rm = TRUE)

    # Bonferroni-corrected quantile test (following Berger)
    q_lo = apply(boot_matrix, 2, quantile, probs = alpha / (2 * n_langs), na.rm = TRUE)
    q_hi = apply(boot_matrix, 2, quantile, probs = 1 - alpha / (2 * n_langs), na.rm = TRUE)
    significant = !(q_lo <= 0 & q_hi >= 0)

    data.frame(
        Variable = lang_names,
        coef = round(mean_coef, 3),
        se = round(boot_se, 3),
        q_lo = round(q_lo, 3),
        q_hi = round(q_hi, 3),
        sig_bonf = significant,
        stringsAsFactors = FALSE,
        row.names = NULL
    )
}

boot_summary_m1 = summarizeBootstrap(boot_coefs_m1)
boot_summary_m4 = summarizeBootstrap(boot_coefs_m4)

cat("=== M1 Bootstrap (Pooled NBR + Labeling Noise) ===\n")
cat(sprintf("Significant languages (Bonferroni): %s\n",
    paste(boot_summary_m1$Variable[boot_summary_m1$sig_bonf], collapse = ", ")))
cat(sprintf("Count: %d / %d\n\n", sum(boot_summary_m1$sig_bonf), nrow(boot_summary_m1)))

cat("=== M4 Bootstrap (Two-Way FE Poisson + Labeling Noise) ===\n")
cat(sprintf("Significant languages (Bonferroni): %s\n",
    paste(boot_summary_m4$Variable[boot_summary_m4$sig_bonf], collapse = ", ")))
cat(sprintf("Count: %d / %d\n\n", sum(boot_summary_m4$sig_bonf), nrow(boot_summary_m4)))

# Display both tables
kable(boot_summary_m1, format = "html", caption = "M1 Bootstrap: Mean Coefficients and Bonferroni CIs") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

kable(boot_summary_m4, format = "html", caption = "M4 Bootstrap: Mean Coefficients and Bonferroni CIs") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```
