---
title: "Sensitivity Analysis for Outcome Mislabeling"
output: html_document
---

This notebook assesses the sensitivity of the panel models to outcome
mislabeling noise (false positive and false negative bug-fix labels).
It applies two approaches:

1. **Berger-style bootstrap** for M1 (pooled NBR), replicating their procedure.
2. **Monte Carlo sensitivity analysis** for M4 (two-way FE), which is more
   appropriate for models that already use cluster-robust standard errors.

# Why MC Sensitivity Rather Than Bootstrap for Panel FE Models

Berger et al. (2019, Section 4.2.4) use a bootstrap that simultaneously
(1) resamples projects with replacement and (2) perturbs bug-fix counts via
binomial draws reflecting labeling noise (FP = 36%, FN = 11%).
This conflates two distinct sources of uncertainty:

1. **Sampling/estimation uncertainty** — how much the coefficients would vary
   across hypothetical repeated samples from the same population.
2. **Measurement error uncertainty** — how much the coefficients shift when the
   mislabeled outcome is corrected.

For their pooled NBR (M1), which uses default MLE standard errors, the
project-level resampling is defensible: it serves as a nonparametric substitute
for cluster-robust inference. But for our panel FE models (M2–M4), which
already use cluster-robust standard errors (clustered at the repository and/or
developer level), the resampling component of the bootstrap re-estimates
sampling variability that the sandwich estimator already accounts for. This
double-counts estimation uncertainty and inflates the intervals, making
significance harder to achieve for reasons unrelated to measurement error.

Layering Bonferroni correction (α = 0.01 / 16 tests) on top of these
already-widened intervals compounds the conservatism further. The effective
per-language confidence level becomes 99.9375%, using the 0.03th and 99.97th
percentiles of a distribution that is itself wider than it should be.

For M4, we therefore use a **Monte Carlo sensitivity analysis** that isolates
measurement error. For each of *B* iterations, we perturb the bug-fix labels
on the *original* data (no resampling) and refit the model. Each refit
produces coefficients and cluster-robust standard errors. We then report:

- What **fraction of MC draws** produce a significant result for each language
  (at the α = 0.05 level, using the model's own cluster-robust *z*-test).
  A coefficient that is significant in ≥95% of draws is robust to label noise.
- The **mean coefficient** and its **MC standard deviation** across draws,
  which shows how much the point estimate moves under mislabeling.

# Setup

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(MASS)
library(fixest)
library(parallel)
library(knitr)
library(kableExtra)

WORKING_DIR = "../data"

if (!file.exists(file.path(WORKING_DIR, "commits.csv"))) {
  unzip(file.path(WORKING_DIR, "commits.csv.zip"), exdir = WORKING_DIR)
}
```

# Helper Functions

```{r, helper_functions}
# Extract language coefficients, SEs, and p-values (zero-sum contrasts)
extractLangResults = function(model, lang_levels) {
    n_langs = length(lang_levels)
    if (inherits(model, "fixest")) {
        ct = as.data.frame(coeftable(model))
    } else {
        ct = as.data.frame(summary(model)$coefficients)
    }
    ct$name = rownames(ct)
    lang_rows = grep("^language", ct$name)
    lang_ct = ct[lang_rows, ]

    modeled = lang_levels[1:(n_langs - 1)]
    ref_coef = -sum(lang_ct[, 1])
    ref_se = tryCatch(sqrt(sum(vcov(model)[lang_rows, lang_rows])),
                      error = function(e) NA)
    ref_pval = if (!is.na(ref_se)) 2 * pnorm(-abs(ref_coef / ref_se)) else NA

    coefs = setNames(c(lang_ct[, 1], ref_coef), c(modeled, lang_levels[n_langs]))
    ses   = setNames(c(lang_ct[, 2], ref_se),   c(modeled, lang_levels[n_langs]))
    pvals = setNames(c(lang_ct[, 4], ref_pval),  c(modeled, lang_levels[n_langs]))

    list(coef = coefs, se = ses, pval = pvals)
}
```

# Load Data

```{r, load_data}
data = read.csv(paste0(WORKING_DIR, "/commits.csv"))
cat(sprintf("Loaded %d commits | %d languages | %d repositories | %d developers\n",
    nrow(data), length(unique(data$language)), length(unique(data$project)),
    length(unique(data$author))))
```

# Prepare Panel Data

## Panel 1: Repository-Language Level (for M1)

```{r, data_preparation_panel_1}
panel_repo = data %>%
    group_by(project, language) %>%
    summarize(
        commits = n(),
        bcommits = sum(isbug),
        tins = sum(insertion),
        max_commit_age = max(commit_age),
        devs = n_distinct(author),
        .groups = "drop"
    ) %>%
    mutate(
        language = factor(language),
        project = factor(project),
        lcommits = log(commits),
        ltins = log(tins),
        lmax_commit_age = log(max_commit_age),
        ldevs = log(devs)
    )

contrasts(panel_repo$language) = contr.sum(nlevels(panel_repo$language))

cat(sprintf("Repository-Language panel: %d observations | %d repositories | %d languages\n",
    nrow(panel_repo), length(unique(panel_repo$project)),
    length(unique(panel_repo$language))))
```

## Panel 3: Developer-Repository-Language Level (for M4)

```{r, data_preparation_panel_3}
panel_dev_repo = data %>%
    group_by(author, project, language) %>%
    summarize(
        commits = n(),
        bcommits = sum(isbug),
        tins = sum(insertion),
        max_commit_age = max(commit_age),
        .groups = "drop"
    ) %>%
    mutate(
        language = factor(language),
        author = factor(author),
        project = factor(project),
        lcommits = log(commits),
        ltins = log(tins),
        lmax_commit_age = log(max_commit_age)
    )

contrasts(panel_dev_repo$language) = contr.sum(nlevels(panel_dev_repo$language))

cat(sprintf("Developer-Repository-Language panel: %d observations | %d developers | %d repositories | %d languages\n",
    nrow(panel_dev_repo), length(unique(panel_dev_repo$author)),
    length(unique(panel_dev_repo$project)), length(unique(panel_dev_repo$language))))
```

# Mislabeling Rates

Berger et al. (2019, Section 4.2.4) found that the keyword heuristic for
labeling bug-fix commits has a **36% false positive rate** and **11% false
negative rate**.

```{r, mc_setup}
FP = 0.36  # false positive rate (Berger Section 4.1.4)
FN = 0.11  # false negative rate
B_m1 = 100000  # M1: Berger-style bootstrap (replicate their 100k)
B_m4 = 1000    # M4: MC sensitivity (two-way FE is expensive)
n_cores = detectCores()

LOG_DIR = file.path(dirname(getwd()), "log")
dir.create(LOG_DIR, showWarnings = FALSE)
LOG_FILE = file.path(LOG_DIR, sprintf("mc_sensitivity_%s.log",
    format(Sys.time(), "%Y%m%d_%H%M%S")))

log_msg = function(msg, level = "INFO") {
    line = sprintf("%s [%s] %s\n",
                   format(Sys.time(), "%Y-%m-%d %H:%M:%S"), level, msg)
    cat(line, file = LOG_FILE, append = TRUE)
}

log_msg(sprintf("Logging to %s", LOG_FILE))
log_msg(sprintf("Config: M1 bootstrap=%d / M4 MC=%d iterations, %d CPU cores",
    B_m1, B_m4, n_cores))
```

## M1: Berger-Style Bootstrap (Pooled NBR)

We replicate the Berger et al. bootstrap for M1: resample projects with
replacement **and** perturb bug-fix labels. This is appropriate here because M1
uses default MLE standard errors (no cluster-robust SEs), so the project-level
resampling provides the nonparametric cluster-level inference that the model
itself does not.

```{r, mc_m1, warning=FALSE}
lang_levels_m1 = levels(panel_repo$language)
n_langs_m1 = length(lang_levels_m1)
projects_m1 = unique(panel_repo$project)
n_projects_m1 = length(projects_m1)
proj_rows_m1 = split(seq_len(nrow(panel_repo)), panel_repo$project)

log_msg(sprintf("Running M1 bootstrap (%d iterations across %d cores)...",
    B_m1, n_cores))
t0 = Sys.time()

cl = makeCluster(n_cores)
clusterExport(cl, c("panel_repo", "proj_rows_m1", "projects_m1",
                     "n_projects_m1", "lang_levels_m1", "n_langs_m1",
                     "FP", "FN", "extractLangResults", "B_m1", "LOG_FILE"))
clusterEvalQ(cl, { library(MASS) })
clusterSetRNGStream(cl, iseed = 42)

mc_results_m1 = parLapply(cl, 1:B_m1, function(b) {
    if (b %% 1000 == 0) {
        cat(sprintf("%s [INFO] [M1] %d / %d (%.0f%%)\n",
                    format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
                    b, B_m1, b / B_m1 * 100),
            file = LOG_FILE, append = TRUE)
    }

    # Resample projects (needed for M1 which lacks cluster-robust SEs)
    sampled = sample(projects_m1, n_projects_m1, replace = TRUE)
    idx = unlist(proj_rows_m1[as.character(sampled)])
    mc_data = panel_repo[idx, ]

    # Perturb labels
    mc_data$bcommits = rbinom(nrow(mc_data), mc_data$bcommits, 1 - FP) +
                       rbinom(nrow(mc_data), mc_data$commits - mc_data$bcommits, FN)

    fit = tryCatch(
        glm.nb(bcommits ~ lmax_commit_age + ltins + ldevs + lcommits + language,
               contrasts = list(language = contr.sum),
               data = mc_data),
        error = function(e) NULL,
        warning = function(w) suppressWarnings(
            glm.nb(bcommits ~ lmax_commit_age + ltins + ldevs + lcommits + language,
                   contrasts = list(language = contr.sum),
                   data = mc_data)
        )
    )

    if (!is.null(fit)) {
        res = tryCatch(extractLangResults(fit, lang_levels_m1),
                       error = function(e) NULL)
        if (!is.null(res)) {
            return(list(coef = res$coef, se = res$se, pval = res$pval))
        }
    }
    list(coef = rep(NA, n_langs_m1), se = rep(NA, n_langs_m1),
         pval = rep(NA, n_langs_m1))
})

stopCluster(cl)

mc_coefs_m1 = do.call(rbind, lapply(mc_results_m1, `[[`, "coef"))
mc_ses_m1   = do.call(rbind, lapply(mc_results_m1, `[[`, "se"))
mc_pvals_m1 = do.call(rbind, lapply(mc_results_m1, `[[`, "pval"))
colnames(mc_coefs_m1) = colnames(mc_ses_m1) = colnames(mc_pvals_m1) = lang_levels_m1

elapsed_m1 = difftime(Sys.time(), t0, units = "mins")
n_success_m1 = sum(complete.cases(mc_coefs_m1))
log_msg(sprintf("M1 done in %.1f minutes. %d/%d iterations converged.",
    elapsed_m1, n_success_m1, B_m1))
```

## M4: Two-Way FE Poisson (MC Sensitivity Only — No Resampling)

For M4, cluster-robust SEs (two-way: author + project) already account for
sampling uncertainty. The MC only perturbs labels on the original data.

```{r, mc_m4, warning=FALSE}
lang_levels_m4 = levels(panel_dev_repo$language)
n_langs_m4 = length(lang_levels_m4)

log_msg(sprintf("Running M4 MC sensitivity (%d iterations across %d cores)...",
    B_m4, n_cores))
t0 = Sys.time()

cl = makeCluster(n_cores)
clusterExport(cl, c("panel_dev_repo", "lang_levels_m4", "n_langs_m4",
                     "FP", "FN", "extractLangResults", "B_m4", "LOG_FILE"))
clusterEvalQ(cl, { library(fixest) })
clusterSetRNGStream(cl, iseed = 123)

mc_results_m4 = parLapply(cl, 1:B_m4, function(b) {
    if (b %% 10 == 0) {
        cat(sprintf("%s [INFO] [M4] %d / %d (%.0f%%)\n",
                    format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
                    b, B_m4, b / B_m4 * 100),
            file = LOG_FILE, append = TRUE)
    }

    # No resampling — perturb labels on original data
    mc_data = panel_dev_repo
    mc_data$bcommits = rbinom(nrow(mc_data), mc_data$bcommits, 1 - FP) +
                       rbinom(nrow(mc_data), mc_data$commits - mc_data$bcommits, FN)

    mc_data$author = factor(mc_data$author)
    mc_data$project = factor(mc_data$project)
    mc_data$language = factor(mc_data$language)
    contrasts(mc_data$language) = contr.sum(nlevels(mc_data$language))

    fit = tryCatch(
        fepois(bcommits ~ lmax_commit_age + ltins + lcommits + language | author + project,
               data = mc_data, vcov = ~author + project),
        error = function(e) NULL,
        warning = function(w) tryCatch(
            suppressWarnings(
                fepois(bcommits ~ lmax_commit_age + ltins + lcommits + language | author + project,
                       data = mc_data, vcov = ~author + project)
            ), error = function(e) NULL
        )
    )

    if (!is.null(fit)) {
        res = tryCatch(extractLangResults(fit, lang_levels_m4),
                       error = function(e) NULL)
        if (!is.null(res)) {
            return(list(coef = res$coef, se = res$se, pval = res$pval))
        }
    }
    list(coef = rep(NA, n_langs_m4), se = rep(NA, n_langs_m4),
         pval = rep(NA, n_langs_m4))
})

stopCluster(cl)

mc_coefs_m4 = do.call(rbind, lapply(mc_results_m4, `[[`, "coef"))
mc_ses_m4   = do.call(rbind, lapply(mc_results_m4, `[[`, "se"))
mc_pvals_m4 = do.call(rbind, lapply(mc_results_m4, `[[`, "pval"))
colnames(mc_coefs_m4) = colnames(mc_ses_m4) = colnames(mc_pvals_m4) = lang_levels_m4

elapsed_m4 = difftime(Sys.time(), t0, units = "mins")
n_success_m4 = sum(complete.cases(mc_coefs_m4))
log_msg(sprintf("M4 done in %.1f minutes. %d/%d iterations converged.",
    elapsed_m4, n_success_m4, B_m4))
```

# Results

```{r, mc_results}
summarizeMC = function(mc_coefs, mc_pvals, alpha = 0.05) {
    lang_names = colnames(mc_coefs)
    n_valid = colSums(!is.na(mc_coefs))

    mean_coef = colMeans(mc_coefs, na.rm = TRUE)
    mc_sd = apply(mc_coefs, 2, sd, na.rm = TRUE)

    # Fraction of MC draws where the coefficient is significant at alpha
    frac_sig = colSums(mc_pvals < alpha, na.rm = TRUE) / n_valid

    # Unadjusted percentile CI from MC distribution
    ci_lo = apply(mc_coefs, 2, quantile, probs = alpha / 2, na.rm = TRUE)
    ci_hi = apply(mc_coefs, 2, quantile, probs = 1 - alpha / 2, na.rm = TRUE)

    # Fraction of draws where sign matches the mean (sign stability)
    sign_stable = ifelse(
        mean_coef > 0,
        colSums(mc_coefs > 0, na.rm = TRUE) / n_valid,
        colSums(mc_coefs < 0, na.rm = TRUE) / n_valid
    )

    data.frame(
        Language = lang_names,
        coef = round(mean_coef, 3),
        mc_sd = round(mc_sd, 3),
        ci_lo = round(ci_lo, 3),
        ci_hi = round(ci_hi, 3),
        frac_sig = round(frac_sig, 3),
        sign_stable = round(sign_stable, 3),
        stringsAsFactors = FALSE,
        row.names = NULL
    )
}

mc_summary_m1 = summarizeMC(mc_coefs_m1, mc_pvals_m1)
mc_summary_m4 = summarizeMC(mc_coefs_m4, mc_pvals_m4)

printMCSummary = function(summary_df, model_label) {
    cat(sprintf("=== %s ===\n", model_label))
    robust = summary_df$Language[summary_df$frac_sig >= 0.95]
    fragile = summary_df$Language[summary_df$frac_sig > 0 & summary_df$frac_sig < 0.95]
    never = summary_df$Language[summary_df$frac_sig == 0]
    cat(sprintf("Robust (sig in ≥95%% of draws):  %s\n",
        if (length(robust) > 0) paste(robust, collapse = ", ") else "none"))
    cat(sprintf("Fragile (sig in <95%% of draws): %s\n",
        if (length(fragile) > 0) paste(fragile, collapse = ", ") else "none"))
    cat(sprintf("Never significant:               %s\n\n",
        if (length(never) > 0) paste(never, collapse = ", ") else "none"))
}

printMCSummary(mc_summary_m1, "M1: Pooled NBR (Berger-style bootstrap)")
printMCSummary(mc_summary_m4, "M4: Two-Way FE Poisson (MC sensitivity, no resampling)")
```

```{r, mc_tables}
formatMCTable = function(summary_df, caption) {
    display_df = summary_df
    display_df$frac_sig = sprintf("%.1f%%", display_df$frac_sig * 100)
    display_df$sign_stable = sprintf("%.1f%%", display_df$sign_stable * 100)
    colnames(display_df) = c("Language", "Mean Coef", "MC SD", "CI lo", "CI hi",
                              "% Sig", "% Sign Stable")
    kable(display_df, format = "html", caption = caption, align = c("l", rep("r", 6))) %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width = FALSE) %>%
        footnote(general = paste(
            "Mean Coef: average across MC draws.",
            "MC SD: standard deviation of coefficient across draws.",
            "CI: 95% percentile interval of MC distribution.",
            "% Sig: fraction of draws where the coefficient's cluster-robust p < 0.05.",
            "% Sign Stable: fraction of draws where the coefficient has the same sign as the mean."
        ))
}

formatMCTable(mc_summary_m1,
    "M1 (Pooled NBR): Berger-Style Bootstrap for Outcome Mislabeling")
formatMCTable(mc_summary_m4,
    "M4 (Two-Way FE Poisson): MC Sensitivity to Outcome Mislabeling")
```
